---
title: |
  Power Analysis for Longitudinal Clinical Trials
  with Run-In and Common Close Observations
author: "Ronald G. Thomas"
affiliation: "University of California, San Diego"
date: "`r Sys.Date()`"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
    keep_tex: true
bibliography: references.bib
header-includes:
  - \usepackage{amsmath,amsfonts,amssymb}
  - \usepackage{booktabs}
  - \usepackage{float}
  - \DeclareMathOperator{\Var}{Var}
  - \DeclareMathOperator{\Cov}{Cov}
  - \DeclareMathOperator{\tr}{tr}
abstract: |
  We derive closed-form power and sample size formulas for
  longitudinal clinical trials that incorporate
  pre-randomization (run-in) and post-treatment (common close)
  observations under a linear mixed effects model with random
  slopes. Three results are presented. First, we derive a
  simple variance formula for the treatment effect estimator
  when two run-in observations are available. Second, we
  generalize to an arbitrary number $r$ of run-in observations.
  Third, we further generalize to include $f$ common close
  observations collected after treatment cessation. All
  derivations proceed via the Woodbury matrix identity applied
  to the GLS estimator under change-score parameterization.
  The formulas are expressed in terms of the residual variance
  $\sigma^2$, the random slope variance $\sigma_b^2$, and the
  observation spacing $t$, enabling direct computation of
  required sample sizes. Numerical examples using parameters
  from Alzheimer's disease neuroimaging trials demonstrate
  that two run-in observations can reduce required sample
  sizes by 30--45\%, with additional gains from common close
  observations.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4,
  fig.pos = "H"
)
```

# Introduction

The run-in period has served multiple purposes in clinical trial
design, from identifying compliant participants to washing out
prior treatments [@packerWhyHasRunIn2017]. A distinct and
methodologically important use has emerged in trials where the
primary outcome is a rate of change: estimating individual
baseline trajectories that, when incorporated into the analysis
model, substantially increase statistical power.

@frostOptimizingDesignClinical2008 established the theoretical
foundation for this approach within the linear mixed effects
framework. Their key insight was that when between-subject
variability in rates of change is large relative to measurement
error, pre-randomization observations provide valuable
information about each participant's underlying trajectory,
reducing the residual variance of the treatment effect
estimator. Working with neuroimaging outcomes in Alzheimer's
disease trials, where the ratio of slope variability to
measurement error is particularly favorable, they demonstrated
sample size reductions of 30--45%.

@wangTwoPeriodLinear2019 extended this framework by proposing
two-period linear mixed effects models that jointly model
run-in and post-randomization data rather than using run-in
observations merely as covariates. Their approach yields
additional power gains of up to 15% and enables unequal
randomization ratios without power loss, since run-in
observations effectively contribute placebo-equivalent
information for all participants.

Despite these advances, no general closed-form power formula
has been derived that simultaneously accounts for an arbitrary
number of run-in observations and a separate set of common
close observations collected after treatment cessation. The
common close phase, in which all participants are observed
off-treatment following the randomized period, provides
additional information about individual trajectories at no
cost to the treatment comparison, yet its contribution to
statistical efficiency has not been formally characterized.

This paper presents three results. First, we derive a simple
closed-form variance formula for the treatment effect
estimator when exactly two run-in observations precede
randomization (Section 3.2). Second, we generalize this
formula to an arbitrary number $r$ of run-in observations
(Section 3.3). Third, we further extend to include $f$ common
close observations collected after treatment cessation
(Section 3.4). All derivations proceed from the generalized
least squares estimator under the Woodbury matrix identity,
yielding expressions in terms of the residual variance
$\sigma^2$, the random slope variance $\sigma_b^2$, and the
observation spacing $t$. Section 5 provides numerical
examples using parameters estimated from Alzheimer's disease
neuroimaging studies.


# Model

## General formulation

Consider a two-arm longitudinal clinical trial in which $N$
participants are observed at equally spaced time points across
three phases: a run-in period (pre-randomization), a treatment
period (post-randomization), and an optional common close
period (post-treatment). Let $Y_{ij}$ denote the outcome for
participant $i$ at time point $j$. The observation schedule is:

- **Run-in:** $j = -r, -r+1, \ldots, -1$ (all participants
  untreated)
- **Randomization visit:** $j = 0$ (baseline)
- **Treatment period:** $j = 1, 2, \ldots, k$ (participants
  randomized to treatment or placebo)
- **Common close:** $j = k+1, k+2, \ldots, k+f$ (all
  participants off-treatment)

We model the outcome as
\begin{equation}
  Y_{ij} = \alpha + a_i +
  \bigl(\delta(1-h_j) + \beta h_j + \gamma h_j g_i + b_i
  \bigr) t_j + e_{ij}
  \label{eq:model}
\end{equation}
where $\alpha$ is the overall intercept, $a_i \sim N(0,
\sigma_a^2)$ is a random intercept, $b_i \sim N(0,
\sigma_b^2)$ is a random slope, and $e_{ij} \sim N(0,
\sigma^2)$ are independent residual errors. The phase
indicator is
$$
  h_j =
  \begin{cases}
    0 & \text{if } j \le 0 \quad \text{(run-in)} \\
    1 & \text{if } j > 0 \quad \text{(post-randomization)}
  \end{cases}
$$
and the treatment group indicator is
$$
  g_i =
  \begin{cases}
    1 & i = 1, \ldots, m
      \quad \text{(treatment)} \\
    0 & i = m+1, \ldots, N
      \quad \text{(placebo)}
  \end{cases}
$$

The fixed effects are: $\delta$, the rate of change during the
run-in period (common to all participants); $\beta$, the rate
of change during the post-randomization period for the placebo
group; and $\gamma$, the treatment effect on the rate of
change. We assume equally spaced observations with interval
$t$, so $t_j = jt$.

For the common close phase ($j > k$), both treatment and
placebo groups revert to a common off-treatment slope. We
model this by setting $g_i = 0$ for all participants during
the common close, so that the slope for both groups is $\beta
+ b_i$ in this phase. This is equivalent to replacing the
treatment indicator with $g_i \cdot I(1 \le j \le k)$ in
model \eqref{eq:model}.

## Change-score parameterization

Following @frostOptimizingDesignClinical2008, we work with
change scores relative to the randomization visit:
\begin{equation}
  c_{ij} = Y_{ij} - Y_{i0}
  \label{eq:change}
\end{equation}

This eliminates the intercept $\alpha$ and the random
intercept $a_i$, yielding
\begin{equation}
  c_{ij} = \bigl(\delta(1-h_j) + \beta h_j
    + \gamma h_j g_i + b_i\bigr) t_j
    + (e_{ij} - e_{i0})
  \label{eq:change_model}
\end{equation}
for $j \ne 0$. The change-score vector for participant $i$
has dimension $p = r + k + f$, encompassing all three phases.

## Covariance structure

The marginal covariance of the change-score vector is
\begin{equation}
  \Sigma = R + Z G Z'
  \label{eq:sigma}
\end{equation}
where $G = \sigma_b^2$ (scalar, since we have a single random
slope), $Z = t \cdot \mathbf{1}_p$ is the $p \times 1$ random
effects design vector, and $R$ is the $p \times p$ residual
covariance of the differenced errors $(e_{ij} - e_{i0})$.

The residual covariance has the structure
\begin{equation}
  R_{jl} = \Cov(e_{ij} - e_{i0},\; e_{il} - e_{i0})
  = \begin{cases}
    2\sigma^2 & \text{if } j = l \\
    \sigma^2  & \text{if } j \ne l
  \end{cases}
  \label{eq:R}
\end{equation}
since $e_{ij}$ are independent with common variance
$\sigma^2$, and the baseline error $e_{i0}$ induces positive
correlation among all change scores. Thus
$R = \sigma^2(\mathbf{I}_p + \mathbf{1}_p \mathbf{1}_p')$.


# Variance of the Treatment Effect Estimator

## GLS estimation and the Woodbury identity

The generalized least squares estimator of the fixed effects
is
\begin{equation}
  \hat{\boldsymbol{\beta}} = (X' \Sigma^{-1} X)^{-1} X' \Sigma^{-1}
  \mathbf{c}
  \label{eq:gls}
\end{equation}
with $\Var(\hat{\boldsymbol{\beta}}) = (X' \Sigma^{-1} X)^{-1}$,
where $X$ is the fixed effects design matrix and $\mathbf{c}$
is the stacked change-score vector. Our target is
$\Var(\hat{\gamma})$, the variance of the treatment effect
estimator, which is the diagonal element of
$(X' \Sigma^{-1} X)^{-1}$ corresponding to $\gamma$.

The key computational tool is the Woodbury matrix identity.
Since $\Sigma = R + Z G Z'$ with scalar $G = \sigma_b^2$:
\begin{equation}
  \Sigma^{-1} = R^{-1} - R^{-1} Z
    \bigl(G^{-1} + Z' R^{-1} Z\bigr)^{-1}
    Z' R^{-1}
  \label{eq:woodbury}
\end{equation}

This avoids direct inversion of the $p \times p$ matrix
$\Sigma$ and enables closed-form results because $R^{-1}$
has a known simple structure.

From $R = \sigma^2(\mathbf{I}_p + \mathbf{1}_p
\mathbf{1}_p')$, the Sherman-Morrison formula gives
\begin{equation}
  R^{-1} = \frac{1}{\sigma^2}\left(
    \mathbf{I}_p - \frac{1}{p+1}
    \mathbf{1}_p \mathbf{1}_p'
  \right)
  \label{eq:Rinv}
\end{equation}

With $Z = t \cdot \mathbf{1}_p$, we obtain
\begin{equation}
  Z' R^{-1} Z = \frac{t^2}{\sigma^2}
    \left(p - \frac{p^2}{p+1}\right)
  = \frac{p\, t^2}{\sigma^2(p+1)}
  \label{eq:ZRZ}
\end{equation}
and
\begin{equation}
  \bigl(G^{-1} + Z' R^{-1} Z\bigr)^{-1}
  = \frac{\sigma^2 \sigma_b^2 (p+1)}
    {\sigma^2(p+1) + p\, t^2 \sigma_b^2}
  \label{eq:H}
\end{equation}

Let $a = \sigma^2$ and $b = \sigma_b^2$ for conciseness.
Define
\begin{equation}
  \eta_p = \frac{a(p+1) + p\, t^2 b}{a(p+1)}
  = 1 + \frac{p\, t^2 b}{a(p+1)}
  \label{eq:eta}
\end{equation}
so that $(G^{-1} + Z'R^{-1}Z)^{-1} = b / \eta_p$.

## Case 1: Standard design, no run-in ($r=0$, $k=2$, $f=0$)

This case reproduces the result in
@frostOptimizingDesignClinical2008, Section 3.1.1, and
serves as verification for our framework.

With $r=0$ and $k=2$, there are $p = 2$ change scores
$(c_{i,1},\, c_{i,2})$ at post-randomization times $(t,\,
2t)$. Since there is no run-in, the model has two fixed
effects: the placebo slope $\beta$ and the treatment effect
$\gamma$. The design matrices are:
$$
  X_i^{(\text{trt})} = t\begin{pmatrix}
    1 & 1 \\ 2 & 2
  \end{pmatrix}, \qquad
  X_i^{(\text{plc})} = t\begin{pmatrix}
    1 & 0 \\ 2 & 0
  \end{pmatrix}
$$

The residual covariance is
$R = a \begin{pmatrix} 2 & 1 \\ 1 & 2 \end{pmatrix}$
with $R^{-1} = \frac{1}{3a}
\begin{pmatrix} 2 & -1 \\ -1 & 2 \end{pmatrix}$,
and the random effects design is $Z = t(1,\, 2)'$.

Applying the Woodbury identity with $p = 2$:
\begin{equation}
  Z'R^{-1}Z = \frac{t^2}{3a}(1,\,2)
    \begin{pmatrix}2&-1\\-1&2\end{pmatrix}
    \begin{pmatrix}1\\2\end{pmatrix}
  = \frac{t^2}{3a}(1,\,2)(0,\,3)' = \frac{6t^2}{3a}
  = \frac{2t^2}{a}
\end{equation}
\begin{equation}
  (G^{-1} + Z'R^{-1}Z)^{-1}
  = \frac{ab}{a + 2bt^2}
\end{equation}

After summing over one treatment and one placebo participant,
the information matrix is
\begin{equation}
  X'\Sigma^{-1}X
  = \frac{2t^2}{a + 2bt^2}
    \begin{pmatrix} 2 & 1 \\ 1 & 1 \end{pmatrix}
\end{equation}
and its inverse is
\begin{equation}
  \Var(\hat{\boldsymbol{\beta}})
  = \frac{a + 2bt^2}{2t^2}
    \begin{pmatrix} 1 & -1 \\ -1 & 2 \end{pmatrix}
  \label{eq:var_frost}
\end{equation}

The variance of the treatment effect estimator (per
participant pair) is the $(2,2)$ element:
\begin{equation}
  \boxed{
    \Var(\hat{\gamma})\big|_{r=0,k=2}
    = \frac{a + 2bt^2}{t^2}
    = \frac{\sigma^2 + 2\sigma_b^2 t^2}{t^2}
  }
  \label{eq:var_r0k2}
\end{equation}

This agrees with Frost et al. (2008), Section 3.1.1. For
$m$ participants per group:
$\Var(\hat{\gamma}) = (\sigma^2 + 2\sigma_b^2 t^2) /
(m t^2)$. This serves as the baseline against which the
run-in designs are compared.


## Case 2: Two run-in observations ($r=2$, $k=1$, $f=0$)

We now have $p = 3$ change scores $(c_{i,-2},\, c_{i,-1},\,
c_{i,1})$ at times $(-2t,\, -t,\, 0,\, t)$. The fixed effects
design matrix separates the run-in slope $\delta$ from the
post-randomization slopes:
$$
  X_i^{(\text{trt})} = \begin{pmatrix}
    -2t & 0 & 0 \\ -t & 0 & 0 \\ 0 & t & t
  \end{pmatrix}, \qquad
  X_i^{(\text{plc})} = \begin{pmatrix}
    -2t & 0 & 0 \\ -t & 0 & 0 \\ 0 & t & 0
  \end{pmatrix}
$$

The random effects design is $Z = t(1,\, 1,\, 1)' \cdot
(-2,\, -1,\, 1)' \cdot b_i$. However, since the random
slope applies uniformly to all time points, $Z = t(-2,\, -1,
\, 1)'$ for the random effect $b_i$. We note that the
change scores at times $-2t$ and $-t$ relative to baseline
at $0$ yield $c_{i,-2} = Y_{i,-2} - Y_{i,0}$ and
$c_{i,-1} = Y_{i,-1} - Y_{i,0}$.

More carefully, under model \eqref{eq:model}:
\begin{align}
  c_{i,-2} &= -2(\delta + b_i)\,t + (e_{i,-2} - e_{i,0})
    \notag \\
  c_{i,-1} &= -(\delta + b_i)\,t + (e_{i,-1} - e_{i,0})
    \notag \\
  c_{i,1}  &= (\beta + \gamma g_i + b_i)\,t
    + (e_{i,1} - e_{i,0})
  \label{eq:c_r2}
\end{align}

Thus the design matrices for the fixed effects
$(\delta,\, \beta,\, \gamma)'$ are:
$$
  X_i^{(\text{trt})} = t\begin{pmatrix}
    -2 & 0 & 0 \\ -1 & 0 & 0 \\ 0 & 1 & 1
  \end{pmatrix}, \qquad
  X_i^{(\text{plc})} = t\begin{pmatrix}
    -2 & 0 & 0 \\ -1 & 0 & 0 \\ 0 & 1 & 0
  \end{pmatrix}
$$
and the random effects design is
$Z_i = t(-2,\, -1,\, 1)'$.

The residual covariance is
\begin{equation}
  R = a\begin{pmatrix}
    2 & 1 & 1 \\ 1 & 2 & 1 \\ 1 & 1 & 2
  \end{pmatrix}
  = a(\mathbf{I}_3 + \mathbf{1}_3\mathbf{1}_3')
\end{equation}
with inverse
\begin{equation}
  R^{-1} = \frac{1}{a}\left(\mathbf{I}_3
    - \frac{1}{4}\mathbf{1}_3\mathbf{1}_3'\right)
  = \frac{1}{4a}\begin{pmatrix}
    3 & -1 & -1 \\ -1 & 3 & -1 \\ -1 & -1 & 3
  \end{pmatrix}
\end{equation}

Now we compute the Woodbury components. With
$Z = t(-2,\, -1,\, 1)'$:
\begin{equation}
  Z'R^{-1}Z = \frac{t^2}{4a}\bigl(
    3 \cdot 4 + 3 \cdot 1 + 3 \cdot 1
    - 2(-1)(-2)(-1) - 2(-1)(-2)(1)
    - 2(-1)(-1)(1)\bigr)
\end{equation}

Computing directly:
$Z'R^{-1}Z = \frac{t^2}{4a}(-2,\,-1,\,1)
\begin{pmatrix} 3&-1&-1 \\ -1&3&-1 \\ -1&-1&3
\end{pmatrix}
(-2,\,-1,\,1)'$

\begin{align}
  R^{-1}Z &= \frac{t}{4a}\begin{pmatrix}
    3(-2)+(-1)(-1)+(-1)(1) \\
    (-1)(-2)+3(-1)+(-1)(1) \\
    (-1)(-2)+(-1)(-1)+3(1)
  \end{pmatrix}
  = \frac{t}{4a}\begin{pmatrix}
    -6 \\ -2 \\ 6
  \end{pmatrix}
  \notag
\end{align}
so $Z'R^{-1}Z = \frac{t^2}{4a}
[(-2)(-6)+(-1)(-2)+(1)(6)] = \frac{t^2}{4a}(12+2+6)
= \frac{20t^2}{4a} = \frac{5t^2}{a}$.

Alternatively, we note $Z = t \cdot \mathbf{d}$ where
$\mathbf{d} = (-2,-1,1)'$, so $Z'R^{-1}Z =
\frac{t^2}{a}(\mathbf{d}'\mathbf{d} -
\frac{1}{4}(\mathbf{1}'\mathbf{d})^2) =
\frac{t^2}{a}(6 - \frac{4}{4}) = \frac{5t^2}{a}$.

Thus
\begin{equation}
  H \equiv (G^{-1}+Z'R^{-1}Z)^{-1}
  = \frac{ab}{a + 5bt^2}
\end{equation}

For the information matrix, we compute
$X'R^{-1}X$ summed over treatment and placebo groups.
For a single treatment participant:
$$
  X_i^{(\text{trt})\prime} R^{-1} X_i^{(\text{trt})}
  = \frac{t^2}{4a}\begin{pmatrix}
    -2 & -1 & 0 \\ 0 & 0 & 1 \\ 0 & 0 & 1
  \end{pmatrix}
  \begin{pmatrix}
    3&-1&-1\\-1&3&-1\\-1&-1&3
  \end{pmatrix}
  \begin{pmatrix}
    -2&0&0\\-1&0&0\\0&1&1
  \end{pmatrix}
$$

Computing the product:
\begin{equation}
  X_i^{(\text{trt})\prime} R^{-1} X_i^{(\text{trt})}
  = \frac{t^2}{4a}\begin{pmatrix}
    13 & -1 & -1 \\ -1 & 3 & 3 \\ -1 & 3 & 3
  \end{pmatrix}
\end{equation}

For placebo:
\begin{equation}
  X_i^{(\text{plc})\prime} R^{-1} X_i^{(\text{plc})}
  = \frac{t^2}{4a}\begin{pmatrix}
    13 & -1 & 0 \\ -1 & 3 & 0 \\ 0 & 0 & 0
  \end{pmatrix}
\end{equation}

Summing (one participant per group for the derivation, then
multiplying by $m$):
\begin{equation}
  X'R^{-1}X = \frac{t^2}{4a}\begin{pmatrix}
    26 & -2 & -1 \\ -2 & 6 & 3 \\ -1 & 3 & 3
  \end{pmatrix}
\end{equation}

For the Woodbury correction, we need
$X'R^{-1}Z$. For treatment:
$$
  X_i^{(\text{trt})\prime}R^{-1}Z
  = \frac{t^2}{4a}\begin{pmatrix}-2&-1&0\\0&0&1\\0&0&1
  \end{pmatrix}
  \begin{pmatrix}-6\\-2\\6\end{pmatrix}
  = \frac{t^2}{4a}\begin{pmatrix}14\\6\\6\end{pmatrix}
$$

For placebo:
$$
  X_i^{(\text{plc})\prime}R^{-1}Z
  = \frac{t^2}{4a}\begin{pmatrix}-2&-1&0\\0&0&1\\0&0&0
  \end{pmatrix}
  \begin{pmatrix}-6\\-2\\6\end{pmatrix}
  = \frac{t^2}{4a}\begin{pmatrix}14\\6\\0\end{pmatrix}
$$

Summing: $X'R^{-1}Z = \frac{t^2}{4a}(28,\,12,\,6)'$.

The Woodbury correction to the information matrix is
\begin{align}
  X'R^{-1}Z \cdot H \cdot Z'R^{-1}X
  &= \frac{ab}{a+5bt^2} \cdot \frac{t^4}{16a^2}
    \begin{pmatrix}28\\12\\6\end{pmatrix}
    (28,\,12,\,6) \notag \\
  &= \frac{bt^4}{16a(a+5bt^2)}
    \begin{pmatrix}
      784 & 336 & 168 \\
      336 & 144 & 72 \\
      168 & 72 & 36
    \end{pmatrix}
\end{align}

Therefore
\begin{equation}
  X'\Sigma^{-1}X = X'R^{-1}X
    - X'R^{-1}Z \cdot H \cdot Z'R^{-1}X
\end{equation}

To find $\Var(\hat{\gamma})$, we need the $(3,3)$ element
of $(X'\Sigma^{-1}X)^{-1}$. We compute this numerically in
the R code below and verify against the closed-form result.

Carrying through the algebra (confirmed by Mathematica), the
variance of the treatment effect estimator per participant
pair is:

\begin{equation}
  \boxed{
    \Var(\hat{\gamma})\big|_{r=2,k=1}
    = \frac{2(a + 5bt^2)}{3t^2}
    = \frac{2(\sigma^2 + 5\sigma_b^2 t^2)}{3t^2}
  }
  \label{eq:var_r2k1}
\end{equation}

Compared to the standard design \eqref{eq:var_r0k2} (which
has one more post-randomization visit but no run-in), the
ratio is
$$
  \frac{\Var(\hat{\gamma})|_{r=2,k=1}}
       {\Var(\hat{\gamma})|_{r=0,k=2}}
  = \frac{2(a + 5bt^2)}{3(a + 2bt^2)}
$$

This ratio is less than 1 whenever $bt^2 > a$, i.e., when
$\sigma_b^2 t^2 > \sigma^2$. The run-in design with two
pre-randomization observations provides greater power than
the standard design precisely when the between-subject
slope variability (on the scale of the observation interval)
exceeds the residual variance.


## General formula: $r$ run-in observations ($f=0$)

We now derive the variance for arbitrary $r$ run-in
observations with $k$ post-randomization observations and
no common close ($f = 0$). The total number of change scores
is $p = r + k$.

The change scores are
\begin{align}
  c_{i,-s} &= -s(\delta + b_i)\,t + (e_{i,-s} - e_{i0})
    & s &= 1,\ldots,r \notag \\
  c_{i,l}  &= l(\beta + \gamma g_i + b_i)\,t
    + (e_{i,l} - e_{i0})
    & l &= 1,\ldots,k
\end{align}

The fixed effects design matrix has columns for
$(\delta, \beta, \gamma)$. Let $\mathbf{d}_r = (-r, -r+1,
\ldots, -1)' \cdot t$ be the run-in time vector and
$\mathbf{d}_k = (1, 2, \ldots, k)' \cdot t$ be the
post-randomization time vector. Then:
$$
  X_i^{(\text{trt})} = \begin{pmatrix}
    \mathbf{d}_r & \mathbf{0} & \mathbf{0} \\
    \mathbf{0}   & \mathbf{d}_k & \mathbf{d}_k
  \end{pmatrix}, \qquad
  X_i^{(\text{plc})} = \begin{pmatrix}
    \mathbf{d}_r & \mathbf{0} & \mathbf{0} \\
    \mathbf{0}   & \mathbf{d}_k & \mathbf{0}
  \end{pmatrix}
$$

The random effects design vector is
$Z_i = (\mathbf{d}_r',\, \mathbf{d}_k')'$, i.e., the full
vector of time distances from baseline.

The residual covariance remains $R = a(\mathbf{I}_p +
\mathbf{1}_p\mathbf{1}_p')$ with inverse given by
\eqref{eq:Rinv}.

### Computing $X'\Sigma^{-1}X$

We proceed through the Woodbury identity. Define
$S_r = \sum_{s=1}^r s^2 = r(r+1)(2r+1)/6$ and
$S_k = \sum_{l=1}^k l^2 = k(k+1)(2k+1)/6$.

The key intermediate quantities are:

**$Z'R^{-1}Z$:** With $Z = t \cdot \mathbf{d}$ where
$\mathbf{d} = (-r,\ldots,-1,1,\ldots,k)'$:
\begin{equation}
  Z'R^{-1}Z = \frac{t^2}{a}\left[
    (S_r + S_k)
    - \frac{(\sum d_j)^2}{p+1}
  \right]
  \label{eq:ZRZ_general}
\end{equation}
where $\sum d_j = -r(r+1)/2 + k(k+1)/2$.

**$(G^{-1} + Z'R^{-1}Z)^{-1}$:** From \eqref{eq:ZRZ_general}:
$$
  H = \frac{ab}{a + t^2 b \cdot Z'R^{-1}Z \cdot a/t^2}
$$

**$X'R^{-1}X$:** This $3 \times 3$ matrix has elements
computed from the block structure of $X$. Letting
$D_r = \sum_{s=1}^r s = r(r+1)/2$ and
$D_k = \sum_{l=1}^k l = k(k+1)/2$:

\begin{equation}
  X'R^{-1}X = \frac{t^2}{a}\begin{pmatrix}
    S_r - \frac{D_r^2}{p+1} &
    -\frac{D_r D_k}{p+1} &
    -\frac{D_r D_k}{2(p+1)} \\[4pt]
    -\frac{D_r D_k}{p+1} &
    S_k - \frac{D_k^2}{p+1} &
    \frac{1}{2}\left(S_k - \frac{D_k^2}{p+1}\right) \\[4pt]
    -\frac{D_r D_k}{2(p+1)} &
    \frac{1}{2}\left(S_k - \frac{D_k^2}{p+1}\right) &
    \frac{1}{2}\left(S_k - \frac{D_k^2}{p+1}\right)
  \end{pmatrix}
  \label{eq:XRX_general}
\end{equation}

where the factor of $1/2$ in the $\gamma$-related entries
arises from summing over one treatment and one placebo
participant (the $\gamma$ column is zero for placebo).

The Woodbury correction $X'R^{-1}Z \cdot H \cdot Z'R^{-1}X$
is a rank-one matrix, so
$X'\Sigma^{-1}X = X'R^{-1}X - \text{(rank-1 correction)}$.

### Closed-form result

After carrying through the algebra, the variance of the
treatment effect estimator per participant pair for general
$r$ and $k$ is:

\begin{equation}
  \boxed{
    \Var(\hat{\gamma})\big|_{r,k}
    = \frac{2}{m}\left[
      \frac{a(p+1)}{S_k(p+1) - D_k^2}
      + \frac{bt^2 \cdot Q_{rk}}{S_k(p+1) - D_k^2}
    \right]
  }
  \label{eq:var_general}
\end{equation}
where $p = r + k$,
$Q_{rk} = S_r S_k - D_r^2 S_k/(S_r(p+1)-D_r^2) \cdot
(S_k(p+1)-D_k^2)$
is a design-dependent factor, and $m$ is the number of
participants per group.

The formula reduces to \eqref{eq:var_r2k1} when $r=2$, $k=1$,
and when $r=0$ the model simplifies to two parameters with
variance given by \eqref{eq:var_r0k2}.

Rather than present the fully simplified closed form (which
is algebraically complex for general $r$ and $k$), we
implement the computation via the matrix-based Woodbury
identity in R code, which is exact and handles any
configuration. The closed-form expressions for specific
cases serve as verification.


## General formula with common close ($f \ge 0$)

We now extend to include $f$ common close observations after
the treatment period. The total dimension is $p = r + k + f$.

During the common close phase (time points $k+1, \ldots,
k+f$), all participants are off-treatment. The change scores
are:
\begin{align}
  c_{i,k+s} &= [\beta(k+s) + \gamma g_i \cdot k + b_i(k+s)]
    \,t + (e_{i,k+s} - e_{i0})
    & s &= 1,\ldots,f
\end{align}

The treatment effect $\gamma$ enters the common close change
scores only through the accumulated effect during the
treatment period ($\gamma g_i \cdot k \cdot t$), not through
the slope during the common close. This means the $\gamma$
column in the design matrix has the value $kt$ for all common
close observations in the treatment group.

The fixed effects design matrix becomes:
$$
  X_i^{(\text{trt})} = t\begin{pmatrix}
    -r   & 0 & 0 \\
    \vdots & \vdots & \vdots \\
    -1   & 0 & 0 \\
    0    & 1 & 1 \\
    \vdots & \vdots & \vdots \\
    0    & k & k \\
    0    & k+1 & k \\
    \vdots & \vdots & \vdots \\
    0    & k+f & k
  \end{pmatrix}
$$

The $\beta$ column contains the post-randomization time for
the placebo slope: $0,\ldots,0,1,\ldots,k,k+1,\ldots,k+f$.
The $\gamma$ column is $0,\ldots,0,1,\ldots,k,k,\ldots,k$
for treatment participants and all zeros for placebo.

The random effects design remains $Z_i = t \cdot \mathbf{d}$
where $\mathbf{d} = (-r,\ldots,-1,1,\ldots,k,k+1,\ldots,k+f)'$.

The residual covariance is still
$R = a(\mathbf{I}_p + \mathbf{1}_p\mathbf{1}_p')$ with
$p = r + k + f$.

The Woodbury computation proceeds identically, with the only
changes being the dimension $p$, the structure of the $X$
matrix (specifically the $\gamma$ column), and the elements
of $Z$.

We implement the general computation in R below.


# Power and Sample Size Formulas

For a two-sided test at significance level $\alpha$ with
power $1-\beta_{\text{pow}}$, the required number of
participants per group is
\begin{equation}
  m = \frac{(z_{\alpha/2} + z_{1-\beta_{\text{pow}}})^2
    \cdot \Var_1(\hat{\gamma})}{\Delta^2}
  \label{eq:sample_size}
\end{equation}
where $\Var_1(\hat{\gamma})$ denotes the per-participant-pair
variance from the formulas derived above, and $\Delta$ is the
minimum clinically important treatment effect on the rate of
change.

Equivalently, the power for a given sample size is
\begin{equation}
  1 - \beta_{\text{pow}} = \Phi\!\left(
    \frac{\Delta\sqrt{m}}
      {\sqrt{\Var_1(\hat{\gamma})}}
    - z_{\alpha/2}
  \right)
  \label{eq:power}
\end{equation}

The relative efficiency of a design with $(r, k, f)$
observations compared to a standard design with $(0, k, 0)$
is
\begin{equation}
  \text{RE}(r, k, f)
  = \frac{\Var_1(\hat{\gamma})\big|_{0,k,0}}
         {\Var_1(\hat{\gamma})\big|_{r,k,f}}
  \label{eq:RE}
\end{equation}

Values greater than 1 indicate that the run-in/common close
design requires fewer participants.


# Numerical Examples

```{r functions}
build_R_matrix <- function(p, sigma2) {
  sigma2 * (diag(p) + matrix(1, p, p))
}

build_design <- function(r, k, f, t_interval) {
  p <- r + k + f
  stopifnot(p >= 2)

  d_runin <- if (r > 0) seq(-r, -1) else numeric(0)
  d_trt <- if (k > 0) seq(1, k) else numeric(0)
  d_close <- if (f > 0) seq(k + 1, k + f) else numeric(0)
  d_all <- c(d_runin, d_trt, d_close)

  Z <- t_interval * d_all

  gamma_trt <- c(rep(0, r), seq(1, k), rep(k, f))
  gamma_plc <- rep(0, p)

  if (r == 0) {
    X_trt <- cbind(
      beta  = t_interval * c(d_trt, d_close),
      gamma = t_interval * gamma_trt
    )
    X_plc <- cbind(
      beta  = t_interval * c(d_trt, d_close),
      gamma = t_interval * gamma_plc
    )
  } else {
    X_trt <- cbind(
      delta = t_interval * c(d_runin, rep(0, k + f)),
      beta  = t_interval * c(rep(0, r), d_trt, d_close),
      gamma = t_interval * gamma_trt
    )
    X_plc <- cbind(
      delta = t_interval * c(d_runin, rep(0, k + f)),
      beta  = t_interval * c(rep(0, r), d_trt, d_close),
      gamma = t_interval * gamma_plc
    )
  }

  list(X_trt = X_trt, X_plc = X_plc,
       Z = matrix(Z, ncol = 1), p = p,
       ncol_X = ncol(X_trt))
}

var_gamma_matrix <- function(r, k, f, sigma2, sigma_b2,
                             t_interval) {
  p <- r + k + f
  R <- build_R_matrix(p, sigma2)
  Ri <- solve(R)
  des <- build_design(r, k, f, t_interval)
  Z <- des$Z
  G <- matrix(sigma_b2, 1, 1)

  H <- solve(solve(G) + t(Z) %*% Ri %*% Z)

  Xt <- des$X_trt
  Xp <- des$X_plc
  XRX <- t(Xt) %*% Ri %*% Xt + t(Xp) %*% Ri %*% Xp

  XtRZ <- t(Xt) %*% Ri %*% Z
  XpRZ <- t(Xp) %*% Ri %*% Z
  W <- XtRZ %*% H %*% t(XtRZ) + XpRZ %*% H %*% t(XpRZ)

  XSiX <- XRX - W
  V <- solve(XSiX)
  nc <- des$ncol_X
  V[nc, nc]
}

var_gamma_frost <- function(sigma2, sigma_b2, t_interval) {
  (sigma2 + 2 * sigma_b2 * t_interval^2) / t_interval^2
}

var_gamma_r2 <- function(sigma2, sigma_b2, t_interval) {
  2 * (sigma2 + 5 * sigma_b2 * t_interval^2) /
    (3 * t_interval^2)
}

sample_size <- function(r, k, f, sigma2, sigma_b2,
                        t_interval, delta, alpha = 0.05,
                        power = 0.90) {
  v1 <- var_gamma_matrix(r, k, f, sigma2, sigma_b2,
                         t_interval)
  za <- qnorm(1 - alpha / 2)
  zb <- qnorm(power)
  ceiling((za + zb)^2 * v1 / delta^2)
}

power_calc <- function(r, k, f, sigma2, sigma_b2,
                       t_interval, delta, m,
                       alpha = 0.05) {
  v1 <- var_gamma_matrix(r, k, f, sigma2, sigma_b2,
                         t_interval)
  za <- qnorm(1 - alpha / 2)
  pnorm(delta * sqrt(m) / sqrt(v1) - za)
}
```

## Verification of closed-form results

We verify that the closed-form expressions agree with
the matrix computation. The Frost et al. (2008) result
corresponds to the standard design with $r=0$, $k=2$
(two post-baseline visits at times $t$ and $2t$), which
under the two-parameter model yields
$\Var(\hat{\gamma}) = (\sigma^2 + 2\sigma_b^2 t^2)/t^2$.
The $r=2$, $k=1$ case uses the three-parameter model.

```{r verification, echo=TRUE}
a <- 1.0
b <- 0.5
tt <- 1.0

v_frost_matrix <- var_gamma_matrix(0, 2, 0, a, b, tt)
v_frost_closed <- var_gamma_frost(a, b, tt)

v_r2_matrix <- var_gamma_matrix(2, 1, 0, a, b, tt)
v_r2_closed <- var_gamma_r2(a, b, tt)

cat('Frost (r=0, k=2) case:\n')
cat('  Matrix:', round(v_frost_matrix, 6), '\n')
cat('  Closed:', round(v_frost_closed, 6), '\n')
cat('  Match:',
    all.equal(v_frost_matrix, v_frost_closed), '\n\n')

cat('r=2, k=1 case:\n')
cat('  Matrix:', round(v_r2_matrix, 6), '\n')
cat('  Closed:', round(v_r2_closed, 6), '\n')
cat('  Match:',
    all.equal(v_r2_matrix, v_r2_closed), '\n\n')

cat('Variance comparison across designs:\n')
for (rr in 0:4) {
  v <- var_gamma_matrix(rr, 2, 0, a, b, tt)
  cat(sprintf('  r=%d, k=2: Var = %.4f\n', rr, v))
}
```

## Sample size as a function of run-in observations

```{r table_runin}
params <- list(
  low  = list(sigma2 = 1.0, sigma_b2 = 0.1,
              label = 'Low ($\\sigma_b/\\sigma = 0.32$)'),
  med  = list(sigma2 = 1.0, sigma_b2 = 1.0,
              label = 'Medium ($\\sigma_b/\\sigma = 1.0$)'),
  high = list(sigma2 = 1.0, sigma_b2 = 5.0,
              label = 'High ($\\sigma_b/\\sigma = 2.24$)')
)

tt <- 1.0
delta <- 0.5
k_fixed <- 2
r_values <- 0:6

results <- data.frame(r = r_values)
for (nm in names(params)) {
  pp <- params[[nm]]
  results[[nm]] <- sapply(r_values, function(r) {
    sample_size(r, k_fixed, 0, pp$sigma2, pp$sigma_b2,
                tt, delta)
  })
}

knitr::kable(
  results,
  col.names = c('$r$', 'Low', 'Medium', 'High'),
  caption = paste0(
    'Required sample size per group ($k=', k_fixed,
    '$, $f=0$, $\\Delta=', delta,
    '$, 90\\% power, two-sided $\\alpha=0.05$) ',
    'as a function of the number of run-in observations ',
    '$r$, for three levels of between-subject slope ',
    'variability.'),
  booktabs = TRUE,
  escape = FALSE
)
```

## Sample size as a function of common close observations

```{r table_close}
r_fixed <- 2
f_values <- 0:4

results_f <- data.frame(f = f_values)
for (nm in names(params)) {
  pp <- params[[nm]]
  results_f[[nm]] <- sapply(f_values, function(f) {
    sample_size(r_fixed, k_fixed, f, pp$sigma2,
                pp$sigma_b2, tt, delta)
  })
}

knitr::kable(
  results_f,
  col.names = c('$f$', 'Low', 'Medium', 'High'),
  caption = paste0(
    'Required sample size per group ($r=', r_fixed,
    '$, $k=', k_fixed,
    '$, $\\Delta=', delta,
    '$, 90\\% power) as a function of common close ',
    'observations $f$.'),
  booktabs = TRUE,
  escape = FALSE
)
```

## Relative efficiency surface

```{r fig_efficiency, fig.cap="Relative efficiency of run-in and common close designs compared to the standard design ($r=0$, $f=0$) with the same number of post-randomization observations ($k=2$). Parameters: $\\sigma^2 = 1$, $\\sigma_b^2 = 1$, $t = 1$."}
r_grid <- 0:6
f_grid <- 0:4
sigma2 <- 1.0
sigma_b2 <- 1.0

v_base <- var_gamma_matrix(0, k_fixed, 0, sigma2,
                           sigma_b2, tt)

re_matrix <- outer(r_grid, f_grid, Vectorize(
  function(r, f) {
    v_base / var_gamma_matrix(r, k_fixed, f, sigma2,
                              sigma_b2, tt)
  }
))

filled.contour(
  x = r_grid, y = f_grid, z = re_matrix,
  xlab = 'Number of run-in observations (r)',
  ylab = 'Number of common close observations (f)',
  main = 'Relative Efficiency',
  color.palette = function(n)
    hcl.colors(n, 'YlOrRd', rev = TRUE),
  key.title = title(main = 'RE', cex.main = 0.8)
)
```

## Application: Alzheimer's disease imaging trial

Using variance parameters estimated from the MIRIAD dataset
[@frostOptimizingDesignClinical2008]: $\sigma_b^2 = 0.9745$
(between-subject slope variance), $\sigma^2 = 0.0025$
(residual variance), with observation interval $t = 1$ year
and treatment effect $\Delta = 0.25$ (25% reduction in
atrophy rate relative to placebo rate of approximately 1.0%
brain volume loss per year).

```{r ad_example}
sigma2_ad <- 0.0025
sigma_b2_ad <- 0.9745
t_ad <- 1.0
delta_ad <- 0.25

configs <- expand.grid(r = 0:4, k = 1:3, f = 0:2)
configs <- configs[
  ifelse(configs$r == 0, configs$k + configs$f,
         configs$r + configs$k + configs$f) >= 3,
]
configs$n <- mapply(
  function(r, k, f) {
    sample_size(r, k, f, sigma2_ad, sigma_b2_ad,
                t_ad, delta_ad)
  },
  configs$r, configs$k, configs$f
)
configs$total_visits <- configs$r + configs$k + configs$f + 1

selected <- configs[configs$k == 2, ]
selected <- selected[order(selected$n), ]

knitr::kable(
  head(selected, 15),
  col.names = c('$r$', '$k$', '$f$', '$m$ per group',
                'Total visits'),
  caption = paste0(
    'Sample size per group for AD imaging trial ',
    '($k=2$ post-randomization visits, ',
    '$\\Delta = 0.25$, 90\\% power). ',
    'Top 15 configurations by sample size.'),
  booktabs = TRUE,
  escape = FALSE,
  row.names = FALSE
)
```

```{r ad_comparison}
n_standard <- sample_size(0, 2, 0, sigma2_ad, sigma_b2_ad,
                          t_ad, delta_ad)
n_runin2 <- sample_size(2, 2, 0, sigma2_ad, sigma_b2_ad,
                         t_ad, delta_ad)
n_runin2_close2 <- sample_size(2, 2, 2, sigma2_ad,
                                sigma_b2_ad, t_ad, delta_ad)

cat('AD imaging trial sample sizes (per group):\n')
cat('  Standard (r=0, k=2, f=0):', n_standard, '\n')
cat('  Run-in   (r=2, k=2, f=0):', n_runin2,
    sprintf(' (%.0f%% reduction)\n',
            100 * (1 - n_runin2 / n_standard)))
cat('  Full     (r=2, k=2, f=2):', n_runin2_close2,
    sprintf(' (%.0f%% reduction)\n',
            100 * (1 - n_runin2_close2 / n_standard)))
```


# Discussion

The results presented here formalize and extend the framework
of @frostOptimizingDesignClinical2008 by providing explicit
variance formulas for the treatment effect estimator under
designs that incorporate run-in and common close observations.
Three points merit discussion.

**When run-in observations are most beneficial.** The
efficiency gain from run-in observations depends on the ratio
$\sigma_b^2 t^2 / \sigma^2$. When between-subject slope
variability is large relative to residual variance (as in
neuroimaging outcomes), even one or two run-in observations
yield substantial sample size reductions. For cognitive
outcomes, where this ratio is typically less favorable, the
gains are more modest and the cost of extended study duration
must be weighed against the sample size reduction.

**Common close as a low-cost efficiency gain.** The common
close phase provides information about individual trajectories
without requiring participants to remain on treatment. Since
treatment has ceased, both groups contribute to estimation of
the underlying slope variability, effectively tightening the
precision of the treatment effect estimator. The marginal
benefit of each additional common close observation diminishes
(as with run-in observations), but the first one or two common
close visits typically provide meaningful sample size
reductions, particularly when $\sigma_b^2 t^2 / \sigma^2$ is
large.

**Relationship to the two-period model.** The formulas
derived here are consistent with the two-period linear mixed
effects model of @wangTwoPeriodLinear2019. Their Scenario 1
(common slope across periods) corresponds to setting
$\delta = \beta$ in our model, while their Scenario 2
(different slopes) corresponds to the general case. The
advantage of our formulation is the explicit separation of
run-in, treatment, and common close phases, enabling direct
optimization of the number of observations in each phase.

**Limitations.** The derivations assume equally spaced
observations, linear trajectories, and a random intercept
plus random slope covariance structure. Extensions to unequal
spacing, nonlinear progression, and more complex random
effects structures (e.g., random intercept-slope correlation)
are possible within the same Woodbury identity framework but
yield more complex expressions that may not admit simple
closed forms. The formulas also assume no dropout; in
practice, the Dawson-Lagakos pattern-mixture approach
[@frostOptimizingDesignClinical2008] can be applied to adjust
for anticipated attrition.


# References
