---
title: "Clinical Trial Power Analysis With One Or More Run-In Observations, X04"
author: "R.G. Thomas"
date: "`r Sys.Date()`"
fontsize: 11pt
geometry: "left=3cm,right=3in,top=2cm,bottom=2cm"
output:
  pdf_document:
    toc: true
    number_sections: true
    highlight: tango
    keep_tex: true 
    includes:
      in_header: "~/shr/preamble.tex"
header-includes:
   - \rhead{Lab report}
   - \fancyfoot[L]{\currfilename} %put current file in footer
   - \setlength{\headheight}{13.59999pt}
latex_engine: xelatex
---

```{r  include=FALSE, echo=F}
# clear env: objects and packages

rm(list = ls())

options(
  knitr.kable.NA = "", dplyr.summarise.inform = FALSE,
  dplyr.print_max = 1e9, knitr.table.format = "latex"
)
library(pacman)
p_load(skimr, janitor, kableExtra, tidyverse, knitr, lubridate, readxl)
opts_chunk$set(
  warning = F, message = F, echo = F, fig.width = 3.2,
  fig.height = 3, results = "asis", dev = "pdf",
  opts_chunk$set(cache.path = "cache/")
)
source("~/shr/zz.tools.R")
options(scipen = 1, digits = 2)
a <- 1
```
# Notes

The first part of this project is to read the Frost paper "cover-to-cover".
Notes of import will be reproduced here...
@frostOptimizingDesignClinical2008

Second part is to record all useful calculations in the black "Longitudinal"
notebook. 

Third is to use mathematica to compute the variance of the fixed parameters. 

The paper by Wang et al targets the same problem we are interested in. 
@wangTwoPeriodLinear2019. The key question is how best to model the run-in
information to get maximum advantage. What if the time course profile is not
linear? should we model the slope or the variance? 

2022-10-20
made sure the key pubs from the folder are contained in Zotero container


2022-10-20
setting up mathematica. 
start by working on frost3.1.1
Math file names frost311.m
> wolframscript -f frost311.m

test = \cite{frostAnalysisRepeatedDirect2004}
\cite{frisonLinearlyDivergentTreatment1997}

# Introduction

(Start with an Extended intro regarding run-in information in clinical trials.
Pros and Cons.  Not restricted to AD trials. )

# Methods

# Results
# Code 

<!-- your comment 
```{r, attr.source='.numberLines', ref.label=knitr::all_labels(),echo=TRUE,eval=FALSE}
```
-->

\bibliography{x04.bib}
\bibliographystyle{plain}
\nocite{*}
