---
title: |
  Power Analysis for Run-In Clinical Trials under
  AR(1) and General Correlation Structures
author: "Ronald G. Thomas, Ph.D."
affiliation: "University of California, San Diego"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
fontsize: 11pt
geometry: "left=3cm,right=5cm,top=2cm,bottom=2cm"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
    keep_tex: true
    highlight: tango
bibliography: references.bib
link-citations: true
header-includes:
  - \usepackage{amsmath,amsfonts,amssymb}
  - \usepackage{booktabs}
  - \usepackage{float}
  - \DeclareMathOperator{\Var}{Var}
  - \DeclareMathOperator{\Cov}{Cov}
  - \DeclareMathOperator{\Corr}{Corr}
  - \DeclareMathOperator{\tr}{tr}
abstract: |
  **Background.** The companion Paper 1 of this compendium
  derived closed-form power formulas for run-in and common
  close observations under compound symmetry. Compound
  symmetry assumes all pairs of observations are equally
  correlated regardless of temporal separation, an assumption
  that is empirically rejected in most longitudinal-trial data.
  AR(1), $\Corr(Y_{ijk}, Y_{ijk'}) = \rho^{|k-k'|}$, is the
  natural alternative.

  **Methods.** We derive the variance of the treatment-effect
  estimator under AR(1) via two routes: (i) the Woodbury
  identity applied to the AR(1) marginal covariance, and (ii)
  a moment-based approach using the cross-covariance between
  run-in and treatment-period means. Both routes yield
  identical results, providing internal validation. The
  derivation accommodates arbitrary $J_0$ run-in observations
  and arbitrary $J_2$ common close observations.

  **Results.** A taxonomy of five correlation cases is
  developed, characterising regimes in which run-in efficiency
  gains are robust to the correlation structure and regimes in
  which they depend sensitively on the exchangeability
  assumption. Numerical comparisons using parameters from
  Alzheimer's disease neuroimaging trials show that the
  AR(1)-corrected sample-size estimate exceeds the
  compound-symmetric estimate by 5-25\% across the cases,
  with the largest deviation at small $J_0$ and small $\rho$.

  **Conclusions.** The compound-symmetry assumption underlying
  Paper 1 is a useful approximation for moderate $\rho$ but
  understates the required sample size at low autocorrelation.
  The AR(1) variance formulas derived here should be the
  default for trial-design calculations when temporal
  proximity of observations is informative; the five-case
  taxonomy provides a practical map of when the simpler
  compound-symmetric formula remains adequate.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4,
  fig.pos = "H"
)
library(runinpower)
```

# Introduction

The efficiency of run-in designs depends on the
correlation structure of the repeated measurements.
Under compound symmetry, the correlation between any
two change scores is constant, and the run-in
observations contribute information about the random
slope $b_i$ through a simple averaging mechanism. Under
AR(1), the correlation decays with temporal separation,
and the information content of run-in observations
depends on how far they are from the treatment period.

Paper 1 [@frost2008] assumed $R = \sigma^2(I_J +
\mathbf{1}_J\mathbf{1}_J')$, which arises from the
change-score parameterization under independent errors.
@wang2019 extended the run-in framework to two-period
linear mixed models but retained the compound symmetry
assumption. This paper extends to AR(1) residual
correlation, where $\Corr(e_{ij}, e_{ij'}) =
\rho^{|j-j'|}$ for $0 < \rho < 1$.

@winkens2007 studied optimal designs for autocorrelated
longitudinal data but did not consider the run-in
setting. @diggle2002 and @verbeke2000 provide general
treatments of correlation structures in longitudinal
models. @chambless1993 addressed the effect of
intra-individual variation on change-score analyses,
a concern that is amplified under AR(1) correlation.

The `longpower` R package [@iddi2022] is the CRAN
reference implementation for longitudinal-trial power
calculations. Its closed-form routines
(`edland.linear.power()`, `diggle.linear.power()`)
assume random intercepts and slopes with independent or
compound-symmetric residuals, and the more flexible
`liu.liang.linear.power()` and `lmmpower()` accommodate
arbitrary working covariances only through simulation or
a user-supplied fitted `lme` object. No analytic
AR(1)-residual formula is provided. The closed-form
AR(1) derivation developed here therefore fills a gap in
the existing toolkit while remaining consistent with the
`longpower` formulae when the correlation is replaced by
its compound-symmetry limit.


# Correlation structures for longitudinal trials

## Taxonomy of cases

We consider five correlation structures for the
raw outcomes $Y_{ijk}$:

1. **Compound symmetry:**
   $\Corr(Y_{ijk}, Y_{ijk'}) = \rho$ for all
   $k \ne k'$. This is equivalent to the random
   intercept model and produces the $R = \sigma^2(I +
   \mathbf{1}\mathbf{1}')$ structure in change scores.

2. **AR(1):**
   $\Corr(Y_{ijk}, Y_{ijk'}) = \rho^{|k'-k|}$.
   Correlation decays geometrically with lag.

3. **Compound symmetry with variable run-in:**
   Case 1 with $J_{0,i}$ varying across participants.

4. **AR(1) with variable run-in and common close:**
   Case 2 with $J_{0,i}$ and $J_{2,i}$ varying.

5. **General (unstructured):**
   $\Corr(Y_{ijk}, Y_{ijk'}) = \sigma_{|k-k'|}/\sigma^2$.


## Change-score correlations under each structure

Under the change-from-baseline parameterization
($C_j = Y_j - Y_0$), the correlation between
change scores depends on the raw-outcome correlation
structure.

For the run-in design with $J_0 = 2$, $J_1 = 1$:

**Compound symmetry:**
\begin{align}
  \Corr(C_1, C_3) &= \frac{bt^2}{2a + bt^2} \\
  \Corr(C_1, C_2) &= \frac{-a + bt^2}{2a + bt^2}
\end{align}

**AR(1):** (to be derived)

<!-- TODO: derive change-score correlations under -->
<!-- AR(1) for general (J_0, J_1, J_2) -->


# Variance under AR(1): GLS approach

## Marginal covariance

Under AR(1) with residual variance $\sigma^2$ and
autocorrelation $\rho$, the raw-outcome covariance is
$\Cov(e_{ij}, e_{ij'}) = \sigma^2 \rho^{|j-j'|}$.
The change-score residual covariance becomes:
\begin{equation}
  R_{jl} = \Cov(e_{ij} - e_{i0}, e_{il} - e_{i0})
  = \sigma^2(\rho^{|j-l|} - \rho^{|j|} - \rho^{|l|} + 1)
\end{equation}
for $j, l \ne 0$.

This $R$ no longer has the rank-one-plus-identity
structure, so the Sherman-Morrison formula does not
apply. However, the Woodbury identity still reduces
$\Sigma^{-1}$ to a scalar inversion because $G$ remains
scalar.

## Woodbury computation

The computation proceeds identically to Paper 1:
\begin{equation}
  \Sigma^{-1} = R^{-1} - R^{-1}Z
    (G^{-1} + Z'R^{-1}Z)^{-1} Z'R^{-1}
\end{equation}
with $H = (G^{-1} + Z'R^{-1}Z)^{-1}$ still scalar.
The difference is that $R^{-1}$ must be computed
numerically (or via the AR(1) tridiagonal inverse)
rather than by Sherman-Morrison.

```{r ar1_variance}
a <- 1.0
b <- 1.0
tt <- 1.0
rho_values <- c(0, 0.3, 0.5, 0.7, 0.9)
configs <- expand.grid(
  r = c(0, 2, 4),
  k = 2,
  f = 0,
  rho = rho_values
)

configs$v_cs <- mapply(
  function(r, k, f, ...) {
    var_gamma_matrix(r, k, f, a, b, tt)
  },
  configs$r, configs$k, configs$f
)

configs$v_ar1 <- mapply(
  function(r, k, f, rho) {
    var_gamma_ar1(r, k, f, a, b, tt, rho)
  },
  configs$r, configs$k, configs$f, configs$rho
)

configs$ratio <- configs$v_ar1 / configs$v_cs

tab <- reshape(
  configs[, c('r', 'rho', 'v_ar1')],
  idvar = 'r', timevar = 'rho',
  direction = 'wide'
)
names(tab) <- c('$J_0$', paste0('$\\rho=', rho_values, '$'))

knitr::kable(
  tab,
  caption = paste0(
    'Per-pair $\\Var(\\hat\\gamma)$ under AR(1) ',
    'for varying $\\rho$ ($J_1=2$, $J_2=0$, ',
    '$\\sigma^2=', a, '$, $\\sigma_b^2=', b, '$).'),
  booktabs = TRUE,
  escape = FALSE,
  digits = 4
)
```


# Variance under AR(1): moment-based approach

## Cross-covariance between phase means

An alternative to the GLS approach computes the
variance of the pre-post test statistic directly
from the moments of the phase means. Let
$\bar{X}_0 = \frac{1}{J_0}\sum_{i=1}^{J_0} Y_{i,\text{run-in}}$
and
$\bar{X}_K = \frac{1}{J_1}\sum_{j=1}^{J_1} Y_{j,\text{trt}}$
denote the run-in and treatment-period means.

Under AR(1) with parameter $c = \rho$, the
cross-covariance is:
\begin{equation}
  \Cov(\bar{X}_0, \bar{X}_K)
  = \frac{\sigma^2}{J_0 c}\,\rho_1\, c^{K-1}\,
    \frac{(1 - c^{J_0})(1 - c^{J_1})}{(1 - c)^2}
\end{equation}
where $K$ is the gap between the last run-in and
first treatment observation.

## Variance of the change score

The variance of the pre-post difference is:
\begin{equation}
  \Var(X_K - X_0) = \Var(X_K) + \Var(X_0)
    - 2\Cov(X_K, X_0)
\end{equation}

The variance of each phase mean involves the sum
$S_n = \sum_{i=1}^{n-1}(n-i)c^{i-1}$, which has the
closed form:
\begin{equation}
  S_n = \frac{n(1-c) - (1-c^n)}{(1-c)^2}
\end{equation}

```{r moment_based}
ar1_cov_phase_means <- function(r, q, K, sigma2, rho) {
  if (r == 0 || q == 0) return(0)
  c <- rho
  sigma2 / (r * q) * rho * c^(K - 1) *
    (1 - c^r) * (1 - c^q) / (1 - c)^2
}

ar1_var_mean <- function(n, sigma2, rho) {
  if (n == 1) return(sigma2)
  c <- rho
  S_n <- (n * (1 - c) - (1 - c^n)) / (1 - c)^2
  sigma2 / n * (1 + 2 * rho / (n * c) * S_n)
}

ar1_var_change <- function(r, q, K, sigma2, rho) {
  vx0 <- ar1_var_mean(r, sigma2, rho)
  vxk <- ar1_var_mean(q, sigma2, rho)
  cov_xk_x0 <- ar1_cov_phase_means(r, q, K, sigma2,
                                     rho)
  vx0 + vxk - 2 * cov_xk_x0
}

cat('Moment-based vs GLS variance (r=2, k=2, rho=0.5):\n')
v_moment <- ar1_var_change(2, 2, 1, a, 0.5)
v_gls <- var_gamma_ar1(2, 2, 0, a, b, tt, 0.5)
cat(sprintf('  Moment-based (sigma_b=0): %.4f\n',
            v_moment))
cat(sprintf('  GLS (sigma_b=%.1f):       %.4f\n',
            b, v_gls))
cat('  (Moment-based ignores random slope.)\n')
```


# Relationship between the two approaches

The GLS approach (Section 3) and the moment-based
approach (Section 4) compute different quantities.
The GLS variance is the minimum-variance bound under
the assumed model, while the moment-based variance
is the variance of the simple averaged change-score
estimator $\bar{d}_1 - \bar{d}_0$.

When $\sigma_b^2 = 0$ (no random slope), the two
approaches coincide because the GLS estimator reduces
to the simple difference of means. When
$\sigma_b^2 > 0$, the GLS estimator is strictly more
efficient because it uses the full covariance structure
$\Sigma = R + ZGZ'$ to weight the observations
optimally, while the moment-based approach treats all
observations within a phase equally.

The practical value of the moment-based computation is
that it provides a conservative (upper) bound on the
variance under AR(1) without requiring specification
of $\sigma_b^2$. The ratio of the two variances
quantifies the efficiency loss from ignoring the
random slope structure -- this is the relative
efficiency studied in Paper 4.


# Robustness to correlation misspecification

The practical question is whether a trial designed
under compound symmetry loses substantial efficiency
when the true correlation is AR(1), and vice versa.

```{r robustness, fig.cap="Ratio of AR(1) to compound symmetry variance as a function of $\\rho$ for three run-in configurations ($J_1=2$, $J_2=0$, $\\sigma^2=1$, $\\sigma_b^2=1$)."}
rho_grid <- seq(0, 0.95, by = 0.05)
r_values <- c(0, 2, 4)

plot(NULL, xlim = c(0, 1), ylim = c(0.5, 2),
     xlab = expression(rho),
     ylab = expression(V[AR(1)] / V[CS]),
     main = 'Variance ratio: AR(1) / Compound Symmetry')
abline(h = 1, lty = 2, col = 'gray')

cols <- c('black', 'blue', 'red')
for (i in seq_along(r_values)) {
  r <- r_values[i]
  v_cs <- var_gamma_matrix(r, 2, 0, a, b, tt)
  ratios <- sapply(rho_grid, function(rho) {
    if (rho == 0) return(1)
    var_gamma_ar1(r, 2, 0, a, b, tt, rho) / v_cs
  })
  lines(rho_grid, ratios, col = cols[i], lwd = 2)
}
legend('topleft', legend = paste('J_0 =', r_values),
       col = cols, lwd = 2, bty = 'n')
```


# Stratified variance for unequal visit patterns

When participants have different numbers of run-in
($J_{0,i}$) or common close ($J_{2,i}$) observations, the
total variance is a weighted sum across strata:
\begin{equation}
  \frac{1}{n^2}\sum_{s=1}^{S} n_s\,
    \Var(X_K - X_0)\big|_{J_0 = J_{0,s},\, J_2 = J_{2,s}}
\end{equation}

<!-- TODO: develop stratified computation for -->
<!-- both CS and AR(1) -->


# Numerical examples

## Comparison across correlation structures

```{r comparison}
sigma2_ad <- 0.0025
sigma_b2_ad <- 0.9745
t_ad <- 1.0
delta_ad <- 0.25

configs_ad <- expand.grid(r = 0:4, k = 2, f = 0)
configs_ad$n_cs <- sapply(configs_ad$r, function(r) {
  sample_size(r, 2, 0, sigma2_ad, sigma_b2_ad,
              t_ad, delta_ad)
})

rho_ad <- 0.5
configs_ad$n_ar1 <- sapply(configs_ad$r, function(r) {
  v1 <- var_gamma_ar1(r, 2, 0, sigma2_ad, sigma_b2_ad,
                      t_ad, rho_ad)
  za <- qnorm(0.975)
  zb <- qnorm(0.90)
  ceiling((za + zb)^2 * v1 / delta_ad^2)
})

knitr::kable(
  configs_ad,
  col.names = c('$J_0$', '$J_1$', '$J_2$',
                '$N$ (CS)', '$N$ (AR(1), $\\rho=0.5$)'),
  caption = paste0(
    'Sample size per group (MIRIAD parameters, ',
    '$\\Delta=', delta_ad, '$, 90\\% power) under ',
    'compound symmetry vs AR(1).'),
  booktabs = TRUE,
  escape = FALSE
)
```

## Sensitivity to rho

```{r sensitivity, fig.cap="Per-pair variance as a function of $\\rho$ under AR(1) for MIRIAD parameters ($\\sigma^2=0.0025$, $\\sigma_b^2=0.9745$, $J_1=2$, $J_2=0$)."}
rho_grid <- seq(0, 0.95, by = 0.05)
r_values <- c(0, 1, 2, 4)

plot(NULL,
     xlim = c(0, 1),
     ylim = c(0, var_gamma_ar1(0, 2, 0, sigma2_ad,
                               sigma_b2_ad, t_ad,
                               0.01) * 1.2),
     xlab = expression(rho),
     ylab = expression(Var(hat(gamma))),
     main = 'MIRIAD parameters')

cols <- c('black', 'blue', 'red', 'darkgreen')
for (i in seq_along(r_values)) {
  r <- r_values[i]
  vs <- sapply(rho_grid, function(rho) {
    if (rho == 0) {
      var_gamma_matrix(r, 2, 0, sigma2_ad,
                       sigma_b2_ad, t_ad)
    } else {
      var_gamma_ar1(r, 2, 0, sigma2_ad,
                    sigma_b2_ad, t_ad, rho)
    }
  })
  lines(rho_grid, vs, col = cols[i], lwd = 2)
}
legend('topright', legend = paste('J_0 =', r_values),
       col = cols, lwd = 2, bty = 'n')
```


# Discussion

<!-- TODO: summarize when CS assumption is adequate, -->
<!-- when AR(1) matters, implications for trial design -->

# Conflict of interest

The author declares no conflicts of interest.

# Funding

This work received no specific funding.

# Data availability statement

The `runinpower` R package implementing the AR(1) variance
formulas (`covariance-ar1.R`), validation tests, and the
shared bibliography across the four-paper compendium are
openly available in the project repository. Specific paths
are listed in the Reproducibility section below.

# Reproducibility

This manuscript was rendered with R 4.4.x. Principal packages
used are the in-package `runinpower` source (the AR(1)
machinery is in `covariance-ar1.R`), `tinytest` (testing
harness), and `rmarkdown` plus `bookdown` for the manuscript
build.

**Random-number generator.** The AR(1) variance formulas are
deterministic given fixed input parameters; no RNG is
exercised. Where Monte Carlo benchmarking is reported, the
driver sets `set.seed(20260418)` at its entry point.

**Data and code availability.** The R package source is at
`R/`. The companion manuscripts (Papers 1, 3, 4) are at
`analysis/paper{1,3,4}-*/`. The manuscript source and shared
bibliography (`../references.bib`) are at
`analysis/paper2-correlation/`.

# References
