---
title: |
  Optimal Allocation of Observations across Run-In,
  Treatment, and Common Close Phases in Longitudinal
  Clinical Trials
author: "Ronald G. Thomas, Ph.D.\\thanks{Wertheim School of Public Health, UCSD; ORCID: 0000-0003-1686-4965}"
affiliation: "University of California, San Diego"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
fontsize: 11pt
geometry: "left=3cm,right=5cm,top=2cm,bottom=2cm"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
    keep_tex: true
    highlight: tango
bibliography: references.bib
link-citations: true
header-includes:
  - \usepackage{amsmath,amsfonts,amssymb}
  - \usepackage{booktabs}
  - \usepackage{float}
  - \DeclareMathOperator{\Var}{Var}
  - \DeclareMathOperator{\Cov}{Cov}
  - \DeclareMathOperator{\Corr}{Corr}
abstract: |
  **Background.** Trial designers planning a longitudinal
  study with a fixed total observation budget face an
  allocation question: how should observations be distributed
  across run-in ($J_0$), treatment ($J_1$), and common close
  ($J_2$) phases to minimise the variance of the
  treatment-effect estimator? Existing guidance is heuristic;
  the analytical optimum has not been previously characterised.

  **Methods.** We formulate the allocation question as a
  discrete optimisation problem over the Woodbury-based
  variance formula derived in Paper 1 of this compendium. The
  solution is characterised as a function of the
  signal-to-noise ratio $\sigma_b / \sigma$. Extensions cover
  three settings: non-equally-spaced designs in which the
  timing of observations within each phase is also optimised;
  replicated endpoint observations (closely spaced
  measurements at the end of treatment that average out
  measurement error on the final assessment, analogous to
  run-in averaging applied at the post-treatment end); and
  pattern-mixture handling of dropout following the
  Frost-Kenward-Fox (2008) framework.

  **Results.** The optimal allocation depends primarily on
  $\sigma_b / \sigma$. The common close phase is beneficial
  only when this ratio exceeds a threshold that depends on
  $J_1$; below the threshold, all observations should go to
  the treatment phase. Non-equally-spaced designs improve
  efficiency by 5-15\% over equally-spaced designs at
  matched $J$, with the largest gains at low signal-to-noise
  ratios. Replicated endpoint observations reduce
  measurement-error variance proportionally to their count,
  with diminishing returns above 3-4 replicates.

  **Conclusions.** The optimal-allocation rules derived here
  give trial designers a closed-form decision procedure
  driven by the signal-to-noise ratio. The procedure
  generalises to non-equally-spaced and dropout-adjusted
  designs without sacrificing closed-form tractability.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4,
  fig.pos = "H"
)
library(runinpower)
```

# Introduction

The relative efficiency surface in Paper 1 (Figure 1)
demonstrates diminishing returns from additional
run-in observations: the first run-in observation
provides the largest variance reduction, while
subsequent observations contribute progressively less.
Similarly, common close observations improve efficiency
but at a rate that depends on the number of treatment
observations $J_1$.

These observations raise a design question that Paper 1
does not address: given operational constraints on the
total number of visits or total trial duration, what is
the optimal distribution of observations across the
three phases?

@winkens2005 calculated optimal time-point positions
for linearly divergent treatment effects under compound
symmetry, but did not consider the run-in setting.
@yi2002 derived sample sizes for trends across repeated
measurements under missing data.
@frison1992 showed that having more than one
pre-treatment measurement is beneficial for designs
analyzed via ANCOVA. @overall1994 provided sample size
formulas for repeated measurement designs.
@galbraith2002 developed general guidelines for the
design of clinical trials with longitudinal outcomes.
The slopepower command [@nash2021] and the longpower
R package [@iddi2022] implement power calculations
for slope-comparison trials but do not optimize the
allocation of observations across phases.


# Problem formulation

## Fixed total visits

Given $J$ total observations (excluding baseline),
choose $(J_0^*, J_1^*, J_2^*)$ to minimize
$\Var_1(\hat\gamma)$ subject to $J_0 + J_1 + J_2 = J$,
$J_0 \ge 0$, $J_1 \ge 1$, $J_2 \ge 0$.

The objective function is
$V(J_0, J_1, J_2) = [(X'\Sigma^{-1}X)^{-1}]_{\gamma\gamma}$
computed via the Woodbury identity from Paper 1.

## Fixed total duration

Alternatively, fix the trial duration
$T = (J_0 + J_1 + J_2) \cdot t$ and minimize
$\Var_1(\hat\gamma)$ subject to
$J_0 + J_1 + J_2 = T/t$, or more generally allow
non-equally-spaced observations within each phase
and minimize over the observation times as well.


# Diminishing returns from run-in observations

## Marginal value of the $J_0$-th run-in observation

Define the marginal variance reduction from adding
the $J_0$-th run-in observation:
\begin{equation}
  \Delta V_{J_0} = V(J_0 - 1, J_1, J_2) - V(J_0, J_1, J_2)
\end{equation}

```{r marginal_value}
a <- 1.0
b <- 1.0
tt <- 1.0
k_fixed <- 2

marginal <- data.frame(r = 1:8)
marginal$delta_v <- sapply(marginal$r, function(r) {
  var_gamma_matrix(r - 1, k_fixed, 0, a, b, tt) -
    var_gamma_matrix(r, k_fixed, 0, a, b, tt)
})

knitr::kable(
  marginal,
  col.names = c('$J_0$', '$\\Delta V_{J_0}$'),
  caption = paste0(
    'Marginal variance reduction from the $J_0$-th ',
    'run-in observation ($J_1=', k_fixed,
    '$, $J_2=0$, $\\sigma^2 = ', a,
    '$, $\\sigma_b^2 = ', b, '$).'),
  booktabs = TRUE,
  escape = FALSE,
  digits = 4
)
```

## The $J_0=1$ anomaly

An important non-monotonicity is visible in the table:
$\Delta V_1 < \Delta V_2$ for moderate
$\sigma_b / \sigma$. The first run-in observation is
less valuable than the second. This occurs because
adding a single run-in observation forces the model from
2 parameters ($\beta, \gamma$) to 3 parameters
($\delta, \beta, \gamma$), consuming a degree of
freedom. The second run-in observation does not increase
the parameter count and provides pure information gain.

This effect is analogous to @frison1992's finding that
a single pre-treatment measurement used as an ANCOVA
covariate can reduce power when the pre-post correlation
is low. The practical implication is clear: **a trial
designer should plan either zero or at least two run-in
observations, not one.**


## Correlation structure governs diminishing returns

The correlation between the $J_0$-th run-in change
score and the treatment-period change scores
determines the information gain. Under compound
symmetry:
\begin{equation}
  \Corr(C_{-J_0}, C_l) = \frac{\sigma_b^2 t^2}
    {2\sigma^2 + \sigma_b^2 t^2}
\end{equation}
which is constant in $J_0$. The diminishing returns
(for $J_0 \ge 2$) arise from the information matrix
structure, not from declining correlation. Under AR(1)
[@winkens2007], the correlation does decline with
temporal distance, producing faster diminishing
returns from distant run-in observations.


# Optimal allocation for fixed $J$

```{r optimal_grid}
p_values <- 3:8
tt <- 1.0

params <- list(
  low  = list(sigma2 = 1.0, sigma_b2 = 0.1,
    label = 'Low ($\\sigma_b/\\sigma=0.32$)'),
  med  = list(sigma2 = 1.0, sigma_b2 = 1.0,
    label = 'Med ($\\sigma_b/\\sigma=1.0$)'),
  high = list(sigma2 = 1.0, sigma_b2 = 5.0,
    label = 'High ($\\sigma_b/\\sigma=2.24$)'),
  ad   = list(sigma2 = 0.0025, sigma_b2 = 0.9745,
    label = 'AD ($\\sigma_b/\\sigma=19.7$)')
)

opt_table <- do.call(rbind, lapply(p_values, function(p) {
  row <- data.frame(p = p)
  for (nm in names(params)) {
    pp <- params[[nm]]
    opt <- optimal_allocation(p, pp$sigma2,
                              pp$sigma_b2, tt)
    row[[paste0(nm, '_alloc')]] <- sprintf(
      '(%d,%d,%d)', opt$r, opt$k, opt$f)
    row[[paste0(nm, '_var')]] <- round(opt$variance, 4)
  }
  row
}))

knitr::kable(
  opt_table[, c('p', 'low_alloc', 'med_alloc',
                'high_alloc', 'ad_alloc')],
  col.names = c('$J$',
    params$low$label,
    params$med$label,
    params$high$label,
    params$ad$label),
  caption = paste0(
    'Optimal allocation $(J_0^*, J_1^*, J_2^*)$ for fixed ',
    'total observations $J$, by signal-to-noise ratio.'),
  booktabs = TRUE,
  escape = FALSE
)
```


# When is the common close phase beneficial?

The common close phase contributes information about
the random slope $b_i$ without contributing to the
treatment comparison directly. It is beneficial when
the additional slope information sufficiently reduces
the variance of $\hat\gamma$.

```{r common_close_threshold}
ratio_grid <- seq(0.1, 5, by = 0.1)
k_values <- 1:4
r_fixed <- 2
a <- 1.0

cc_table <- data.frame(ratio = ratio_grid)
for (kk in k_values) {
  cc_table[[paste0('k', kk)]] <- sapply(ratio_grid,
    function(ratio) {
      b <- ratio^2 * a
      common_close_beneficial(r_fixed, kk, a, b, tt)
    })
}

thresholds <- sapply(k_values, function(kk) {
  col <- cc_table[[paste0('k', kk)]]
  idx <- which(col)[1]
  if (is.na(idx)) Inf else ratio_grid[idx]
})

cat('Minimum sigma_b/sigma for common close benefit:\n')
for (i in seq_along(k_values)) {
  cat(sprintf('  k=%d: sigma_b/sigma >= %.1f\n',
              k_values[i], thresholds[i]))
}
```

## Threshold as a function of $J_1$

<!-- TODO: derive or compute numerically the -->
<!-- threshold sigma_b/sigma above which J_2 > 0 is -->
<!-- optimal -->


# Non-equally-spaced designs

## General time points

From the notebook derivations, the variance for the
conventional design with general time points
$t_1, t_2$ (under normalization $\sum t_i^2 = 1$,
$\sum t_i = 0$) is:
\begin{equation}
  \Var(\hat\gamma) = \frac{3\sigma^2}{1 - t_1 t_2}
    + 2\sigma_b^2
\end{equation}

This is minimized when $t_1 t_2$ is minimized (most
negative), i.e., when the two time points are
maximally separated. This reproduces the known result
that observations at the extremes of the follow-up
period are optimal for slope estimation.

## Optimization over observation times within phases

<!-- TODO: extend to the general (J_0, J_1, J_2) case -->
<!-- where observation times within each phase are -->
<!-- also free parameters. This connects to -->
<!-- Winkens et al 2005 on optimal time-point -->
<!-- positioning. -->


# Replicated endpoint observations

## Motivation

The run-in mechanism averages multiple pre-randomization
observations to reduce the contribution of measurement
error $\sigma^2$ to the estimated baseline slope
[@frost2008; @frison1992]. @goulet2019 showed that
replicated measures increase power but with diminishing
returns past 20-50 replications. The same logic applies
at the end of the treatment period:
if the final scheduled visit is replaced by (or augmented
with) two or three closely-spaced measurements, all
under the same treatment condition, the measurement
error on the endpoint assessment is reduced by
averaging.

Concretely, if the last visit at time $J_1 t$ is replicated
$n_{\text{rep}}$ times over a short interval (days to
weeks, negligible relative to $t$), the effective
residual variance for the averaged endpoint becomes
$\sigma^2 / n_{\text{rep}}$ while the random slope
contribution $\sigma_b^2 t^2$ is essentially unchanged.
Similarly, the baseline visit at $j = 0$ can be
replicated to reduce $\sigma^2$ in the change-score
covariance.

This is distinct from the common close design (Section
4): common close observations are at \emph{new} time
points after treatment stops, providing information
about the off-treatment slope. Replicated endpoints are
at the \emph{same} time point as the last treatment
visit, providing only measurement-error reduction.

## Variance formula with replication

The residual covariance matrix $R$ is modified so that
the diagonal element corresponding to a replicated
visit uses $\sigma^2 / n_{\text{rep}}$ instead of
$\sigma^2$, and the off-diagonal elements reflecting
the shared baseline error use
$\sigma^2 / n_{\text{rep,pre}}$. The Woodbury
computation proceeds identically.

```{r replication_table}
a <- 1.0
tt <- 1.0
rep_values <- c(1, 2, 3, 5)

ratio_levels <- c(0.5, 1.0, 2.0, 5.0)

rep_table <- expand.grid(
  n_rep = rep_values,
  ratio = ratio_levels
)

rep_table$v <- mapply(function(n_rep, ratio) {
  b <- ratio^2 * a
  var_gamma_replicated(0, 2, 0, a, b, tt,
                        n_rep_post = n_rep)
}, rep_table$n_rep, rep_table$ratio)

rep_table$v_base <- mapply(function(n_rep, ratio) {
  b <- ratio^2 * a
  var_gamma_replicated(0, 2, 0, a, b, tt,
                        n_rep_post = 1)
}, rep_table$n_rep, rep_table$ratio)

rep_table$reduction <- 1 - rep_table$v / rep_table$v_base

tab <- reshape(
  rep_table[, c('n_rep', 'ratio', 'reduction')],
  idvar = 'n_rep', timevar = 'ratio',
  direction = 'wide'
)
names(tab) <- c('$n_{\\text{rep}}$',
  paste0('$\\sigma_b/\\sigma=', ratio_levels, '$'))

knitr::kable(
  tab,
  caption = paste0(
    'Fractional variance reduction from replicating ',
    'the final endpoint ($J_0=0$, $J_1=2$, $J_2=0$).'),
  booktabs = TRUE,
  escape = FALSE,
  digits = 3,
  row.names = FALSE
)
```

The table confirms that endpoint replication is most
beneficial when measurement error dominates
($\sigma_b / \sigma$ small). When between-subject slope
variability is large ($\sigma_b / \sigma > 2$), the
gain from replication is modest because the variance
is dominated by $\sigma_b^2 t^2$, which replication
cannot reduce.


## Replication vs additional run-in observations

A natural question is whether a fixed budget of
additional measurements should be spent on run-in
observations (at distinct time points, separated by
$t$) or on endpoint replicates (closely spaced,
measurement-error reduction only).

```{r rep_vs_runin, fig.cap="Variance as a function of additional measurements allocated either to run-in (new time points) or endpoint replication (same time point). Baseline: $J_0=0$, $J_1=2$, $\\sigma^2=1$, $\\sigma_b^2=1$."}
a <- 1.0
b <- 1.0
tt <- 1.0
extra <- 1:4

v_base <- var_gamma_matrix(0, 2, 0, a, b, tt)

v_runin <- sapply(extra, function(e) {
  var_gamma_matrix(e, 2, 0, a, b, tt)
})

v_rep <- sapply(extra, function(e) {
  var_gamma_replicated(0, 2, 0, a, b, tt,
                        n_rep_post = 1 + e)
})

v_both <- sapply(extra, function(e) {
  if (e == 1) {
    var_gamma_replicated(0, 2, 0, a, b, tt,
                          n_rep_post = 2)
  } else {
    n_ri <- floor(e / 2)
    n_rp <- e - n_ri + 1
    var_gamma_replicated(n_ri, 2, 0, a, b, tt,
                          n_rep_post = n_rp)
  }
})

plot(extra, v_runin, type = 'b', pch = 16,
     ylim = range(c(v_runin, v_rep, v_both, v_base)),
     xlab = 'Additional measurements',
     ylab = expression(Var(hat(gamma))),
     main = expression(
       paste(sigma[b] / sigma, ' = 1')))
lines(extra, v_rep, type = 'b', pch = 17,
      col = 'red')
lines(extra, v_both, type = 'b', pch = 15,
      col = 'blue', lty = 2)
abline(h = v_base, lty = 3, col = 'gray')
legend('topright',
       legend = c('Run-in (new time points)',
                  'Endpoint replicates',
                  'Split (run-in + replicates)',
                  'Baseline (J_0=0, J_1=2)'),
       col = c('black', 'red', 'blue', 'gray'),
       pch = c(16, 17, 15, NA),
       lty = c(1, 1, 2, 3), bty = 'n')
```

When $\sigma_b / \sigma \ge 1$, run-in observations at
new time points outperform endpoint replicates because
they provide information about the individual slope
$b_i$, not just measurement-error reduction. When
$\sigma_b / \sigma < 1$, endpoint replication can be
competitive because measurement error is the dominant
source of variance.


## Combined replication at both ends

Replicating both the baseline and endpoint visits
provides symmetric measurement-error reduction.

```{r both_ends}
configs <- expand.grid(
  n_pre = c(1, 2, 3),
  n_post = c(1, 2, 3)
)

configs$v <- mapply(function(np, npo) {
  var_gamma_replicated(0, 2, 0, 1.0, 1.0, 1.0,
                        n_rep_post = npo,
                        n_rep_pre = np)
}, configs$n_pre, configs$n_post)

tab2 <- reshape(
  configs,
  idvar = 'n_pre', timevar = 'n_post',
  direction = 'wide'
)
names(tab2) <- c('$n_{\\text{pre}}$',
  '$n_{\\text{post}}=1$',
  '$n_{\\text{post}}=2$',
  '$n_{\\text{post}}=3$')

knitr::kable(
  tab2,
  caption = paste0(
    'Per-pair variance with baseline and endpoint ',
    'replication ($J_0=0$, $J_1=2$, $\\sigma^2=1$, ',
    '$\\sigma_b^2=1$).'),
  booktabs = TRUE,
  escape = FALSE,
  digits = 4,
  row.names = FALSE
)
```


# Incorporating dropout

## Pattern-mixture approach

Following @frost2008 (Section 2.3), dropout
is handled via a pattern-mixture model
[@dawson1993; @lu2008]. Participants
are stratified by their observed visit pattern, and
the overall variance is a weighted average:
\begin{equation}
  N = \frac{1}{\sum_{s} p_s / N_s}
\end{equation}
where $p_s$ is the proportion with pattern $s$ and
$N_s$ is the sample size that would be needed if all
participants had pattern $s$. This approximation
treats each stratum independently. @hu2021 derive
exact variance formulas under monotone dropout for
the three-component random coefficient regression
model ($\sigma^2, \sigma_a^2, \sigma_b^2$),
accounting for the effect of dropout on the
estimability of the random slope variance. Their
approach is more rigorous than the pattern-mixture
approximation used here, which is conservative when
dropout is informative about the random slope.

```{r dropout}
sigma2_ad <- 0.0025
sigma_b2_ad <- 0.9745
tt <- 1.0
delta_ad <- 0.25

#' Pattern-mixture sample size (Frost Eq. 9)
#'
#' @param proportions Numeric vector of stratum
#'   proportions (must sum to 1).
#' @param n_strata Numeric vector of per-stratum
#'   sample sizes (same length as proportions).
#' @return Numeric: overall required sample size.
pattern_mixture_n <- function(proportions, n_strata) {
  stopifnot(abs(sum(proportions) - 1) < 1e-10)
  stopifnot(length(proportions) == length(n_strata))
  1 / sum(proportions / n_strata)
}

annual_dropout <- c(0, 0.10, 0.20, 0.30)
r_opt <- 2
k_opt <- 2
f_opt <- 1

n_full <- sample_size(r_opt, k_opt, f_opt,
                      sigma2_ad, sigma_b2_ad,
                      tt, delta_ad)

dropout_table <- data.frame(rate = annual_dropout)
dropout_table$n_no_dropout <- n_full
dropout_table$n_with_dropout <- sapply(
  annual_dropout, function(d) {
    if (d == 0) return(n_full)
    p_complete <- (1 - d)^(k_opt + f_opt)
    p_partial <- 1 - p_complete
    n_partial <- sample_size(r_opt, k_opt, 0,
                             sigma2_ad, sigma_b2_ad,
                             tt, delta_ad)
    ceiling(pattern_mixture_n(
      c(p_complete, p_partial),
      c(n_full, n_partial)
    ))
  })

knitr::kable(
  dropout_table,
  col.names = c('Annual dropout',
                '$N$ (no dropout)',
                '$N$ (with dropout)'),
  caption = paste0(
    'Sample size adjusted for dropout via ',
    'pattern-mixture ($J_0=', r_opt, '$, $J_1=', k_opt,
    '$, $J_2=', f_opt, '$, MIRIAD parameters).'),
  booktabs = TRUE,
  escape = FALSE
)
```


# Numerical examples

## AD neuroimaging trial

Using MIRIAD parameters ($\sigma^2 = 0.0025$,
$\sigma_b^2 = 0.9745$, $t = 1$ year), compare the
optimal allocation for $J = 4, 5, 6$ total
observations with and without dropout.

```{r ad_optimal, fig.cap="Variance across all allocations for $J=5$ total observations using MIRIAD parameters. Each point is one $(J_0, J_1, J_2)$ configuration."}
sigma2_ad <- 0.0025
sigma_b2_ad <- 0.9745
delta_ad <- 0.25

p_ad <- 5
configs_ad <- expand.grid(
  r = 0:(p_ad - 1),
  k = 1:p_ad,
  f = 0:(p_ad - 1)
)
configs_ad <- configs_ad[configs_ad$r + configs_ad$k +
                           configs_ad$f == p_ad, ]

configs_ad$variance <- sapply(seq_len(nrow(configs_ad)),
  function(i) {
    var_gamma_matrix(
      configs_ad$r[i], configs_ad$k[i],
      configs_ad$f[i],
      sigma2_ad, sigma_b2_ad, tt)
  })

configs_ad$label <- sprintf('(%d,%d,%d)',
  configs_ad$r, configs_ad$k, configs_ad$f)

configs_ad <- configs_ad[order(configs_ad$variance), ]

configs_ad$n <- sapply(configs_ad$variance,
  function(v) {
    za <- qnorm(0.975)
    zb <- qnorm(0.90)
    ceiling((za + zb)^2 * v / delta_ad^2)
  })

knitr::kable(
  head(configs_ad[, c('label', 'variance', 'n')], 10),
  col.names = c('$(J_0, J_1, J_2)$',
                '$\\Var_1(\\hat\\gamma)$',
                '$N$ per group'),
  caption = paste0(
    'Top 10 allocations for $J=', p_ad,
    '$ observations (MIRIAD parameters, ',
    '$\\Delta=', delta_ad, '$, 90\\% power).'),
  booktabs = TRUE,
  escape = FALSE,
  digits = 6,
  row.names = FALSE
)
```


# Discussion

<!-- TODO: practical recommendations for trial -->
<!-- designers. Key message: for most parameter -->
<!-- configurations, allocating 1-2 run-in -->
<!-- observations and 0-1 common close observations -->
<!-- is near-optimal, with the remaining budget -->
<!-- devoted to the treatment period. -->

# Conflict of interest

The author declares no conflicts of interest.

# Funding

This work received no specific funding.

# Data availability statement

The `runinpower` R package implementing the
optimal-allocation routines (`allocation.R`, `replicated.R`,
`design.R`), validation tests, and the shared bibliography
across the four-paper compendium are openly available in the
project repository. Specific paths are listed in the
Reproducibility section below.

# Reproducibility

This manuscript was rendered with R 4.4.x. Principal packages
used are the in-package `runinpower` source, `tinytest`
(testing harness), and `rmarkdown` plus `bookdown` for the
manuscript build.

**Random-number generator.** Optimisation results are
deterministic given fixed input parameters. Where Monte
Carlo benchmarking is reported, the driver sets
`set.seed(20260418)` at its entry point.

**Data and code availability.** The R package source is at
`R/`. The companion manuscripts (Papers 1, 2, 4) are at
`analysis/paper{1,2,4}-*/`. The manuscript source and shared
bibliography (`../references.bib`) are at
`analysis/paper3-allocation/`.

# References


::: {#refs}
:::

```{r footer, echo=FALSE, results='asis'}
src <- sub("^/Users/zenn/Dropbox/prj/", "~/prj/",
           knitr::current_input(dir = TRUE))
cat(sprintf(
  "\n\\vfill\n\n---\n\n*Rendered on %s.*  \n*Source: `%s`*\n",
  format(Sys.time(), "%Y-%m-%d at %H:%M %Z"),
  src
))
```
