---
title: |
  GLS versus Averaged Change-Score Estimators for
  Clinical Trials with Run-In Observations
author: "Ronald G. Thomas"
affiliation: "University of California, San Diego"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
fontsize: 11pt
geometry: "left=3cm,right=5cm,top=2cm,bottom=2cm"
output:
  pdf_document:
    latex_engine: xelatex
    toc: true
    number_sections: true
    keep_tex: true
    highlight: tango
bibliography: references.bib
link-citations: true
header-includes:
  - \usepackage{amsmath,amsfonts,amssymb}
  - \usepackage{booktabs}
  - \usepackage{float}
  - \DeclareMathOperator{\Var}{Var}
  - \DeclareMathOperator{\Cov}{Cov}
  - \DeclareMathOperator{\E}{E}
abstract: |
  **Background.** Three estimators of the treatment effect in
  longitudinal trials with run-in observations are in routine
  use: the GLS estimator under the linear mixed model, which
  exploits the full covariance structure to extract maximum
  information from all observations; the ANCOVA estimator,
  which regresses post-randomisation means on the run-in mean
  as a covariate without specifying the random-effects
  covariance; and the averaged change-score estimator, which
  computes a simple difference of phase means and applies a
  $t$-test, requiring no variance-component estimation. The
  relative efficiency of these estimators under correct and
  misspecified models has not been comprehensively
  characterised.

  **Methods.** We derive the relative efficiency of the three
  estimators as a function of the trial design
  $(J_0, J_1, J_2)$ and the variance parameters
  $(\sigma^2, \sigma_b^2)$. The derivation proceeds via the
  Woodbury-based variance formulas of Paper 1 in this
  compendium, plus closed-form expressions for the ANCOVA
  and averaged-change-score variances under the same
  random-slopes model. We complement the analytic comparison
  with a Monte Carlo simulation that probes robustness to
  misspecification of the within-subject covariance
  ($R$) and the random-effects covariance ($G$).

  **Results.** Under correct specification, GLS $\le$ ANCOVA
  $\le$ averaged in variance. The efficiency loss from ANCOVA
  relative to GLS is below 5\% across a broad regime of
  ($J_0, J_1, J_2$) and signal-to-noise ratios; conditions
  characterising this regime are derived. Under
  misspecification of $G$ or $R$, the simpler estimators can
  outperform GLS: when the variance components are mis-stated,
  GLS amplifies the misspecification through its inverse-
  covariance weighting, while ANCOVA and averaged remain
  comparatively robust.

  **Conclusions.** GLS is the optimal estimator under correct
  model specification but does not dominate the simpler
  alternatives in the practical regime of mild
  misspecification. ANCOVA is the recommended default when
  the variance components are known only approximately;
  averaged change-score remains useful as a robust
  benchmark and a teaching tool.
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4,
  fig.pos = "H"
)
library(runinpower)
```

# Introduction

The GLS estimator $\hat\gamma_{\text{GLS}} =
(X'\Sigma^{-1}X)^{-1}X'\Sigma^{-1}\mathbf{c}$ is the
minimum-variance unbiased estimator when $\Sigma$ is
known. In practice, $\Sigma$ must be estimated, and
the efficiency of the feasible GLS depends on the
accuracy of this estimation. With small sample sizes
or short follow-up, variance component estimates can
be unstable, potentially eroding the theoretical
advantage of GLS.

The averaged change-score approach sidesteps variance
component estimation entirely. For a trial with $J_0$
run-in and $J_2$ common close observations, the
estimator is:
\begin{equation}
  \hat\gamma_{\text{avg}} = \bar{d}_1 - \bar{d}_0
\end{equation}
where
\begin{equation}
  d_{ik} = \frac{1}{J_{2,i} + 1}
    \sum_{\ell=0}^{J_{2,i}} Y_{i,q+\ell,k}
    - \frac{1}{J_{0,i} + 1}
    \sum_{\ell=0}^{J_{0,i}} Y_{i,\ell,k}
\end{equation}
is the difference of post- and pre-randomization
means for participant $i$.

The debate between change-score and ANCOVA analyses has
a long history [@senn2006; @vanbreukelen2006;
@crager1987]. @yang2001 provided a formal efficiency
comparison for pretest-posttest designs. @liu2009 and
@lu2010 compared baseline-as-covariate against
baseline-as-response (constrained longitudinal data
analysis). @tsiatis2008 developed a semiparametric
framework for covariate adjustment. @mallinckrodt2021
showed that MMRM requires time-by-covariate interactions
to guarantee power gains over ANCOVA. @brogan1980
established the foundational comparison of pretest-posttest
designs. @coffman2016 addressed conditioning on baseline
in longitudinal RCTs. @wang2019 proposed the two-period
approach for run-in trials.

We note that the averaged change-score estimator
studied here is distinct from (and less efficient than)
the constrained longitudinal data analysis (cLDA)
estimator of @liang2000 and @lu2010, which includes
baseline in the response vector with a common-mean
constraint. The cLDA estimator uses the full covariance
structure without requiring specification of the random
effects distribution $G$, placing it between the
averaged estimator and GLS in the efficiency ordering.
@funatogawa2020 extended the cLDA framework to pre-
and post-treatment measurements with equal baseline
assumptions. A full comparison including cLDA is beyond
the scope of this paper but represents a natural
extension.

The `longpower` R package [@iddi2022] provides the
CRAN reference implementation of power calculations for
the GLS / REML-based slope-comparison estimator,
covering the Liu-Liang [@liuliang1997], Diggle
[@diggle2002], and Edland [@ard2011] formulations.
These routines correspond to the $\hat\gamma_{\text{GLS}}$
estimator studied here under correct specification and
constitute the Paper 1 baseline extended by run-in and
common close observations. However, `longpower` does not
cover the ANCOVA-on-run-in-mean estimator or the
averaged change-score $t$-test, nor does it characterize
the efficiency loss from misspecification of $G$ or $R$.
The comparison developed here therefore complements
`longpower` by (i) ordering the three candidate
estimators under correct specification and (ii)
quantifying when the simpler estimators outperform GLS
under misspecified covariance.


# Three estimators

## GLS under the mixed model

From @frost2008, the per-pair variance is the
$(\gamma, \gamma)$ element of
$(X'\Sigma^{-1}X)^{-1}$ with $\Sigma = R + ZGZ'$
[@laird1982; @verbeke2000], computed via the Woodbury
identity.

For the Frost conventional design ($J_0 = 0$, $J_1 = 2$):
\begin{equation}
  \hat\gamma_{\text{GLS}} = \frac{1}{2t}
    [(C_{22} + C_{21}) - (C_{12} + C_{11})]
\end{equation}
with $\Var(\hat\gamma_{\text{GLS}}) =
(\sigma^2 + 2\sigma_b^2 t^2)/t^2$.

## Averaged change score ($t$-test)

The averaged change-score estimator ignores the
within-subject correlation structure and treats
$d_{ik}$ as independent observations:
\begin{equation}
  \Var(\hat\gamma_{\text{avg}})
  = \frac{2}{m}\Var(d_{ik})
\end{equation}

The variance $\Var(d_{ik})$ depends on the correlation
structure but does not require its estimation.

```{r avg_variance, echo=TRUE}
a <- 1.0
b <- 1.0
tt <- 1.0

v_gls <- var_gamma_matrix(2, 2, 0, a, b, tt)
v_avg <- var_gamma_avg(2, 2, 0, a, b, tt)
re <- re_avg_vs_gls(2, 2, 0, a, b, tt)

cat(sprintf('J_0=2, J_1=2: GLS=%.4f, Avg=%.4f, RE=%.4f\n',
            v_gls, v_avg, re))
```


## ANCOVA with run-in mean as covariate

The ANCOVA approach uses the averaged run-in rate as a
baseline covariate without specifying $\sigma_b^2$:
\begin{equation}
  \bar{Y}_{i,\text{post}} = \alpha + \gamma g_i
    + \phi \bar{Y}_{i,\text{pre}} + \varepsilon_i
\end{equation}

The variance of $\hat\gamma$ under this model is
\begin{equation}
  \Var(\hat\gamma_{\text{ANC}})
  = \frac{2}{m}\,\Var(\bar{Y}_{\text{post}})
    \,(1 - \rho_{\text{pre,post}}^2)
\end{equation}
where $\rho_{\text{pre,post}}$ is the marginal
correlation between the run-in and post-randomization
means. When $J_0 = 0$ there is no covariate and ANCOVA
reduces to the averaged estimator. When
$\rho_{\text{pre,post}}$ is large (i.e., when
$\sigma_b^2$ dominates $\sigma^2$), the $(1-\rho^2)$
factor produces substantial variance reduction.

The key advantage over GLS is that $\phi$ is estimated
from the data rather than derived from assumed variance
components [@senn2006; @borm2007]. The key advantage
over the averaged estimator is that it exploits the
predictive value of the run-in observations
[@vanbreukelen2006; @vickers2001].

```{r ancova_demo, echo=TRUE}
v_gls <- var_gamma_matrix(2, 2, 0, a, b, tt)
v_anc <- var_gamma_ancova(2, 2, 0, a, b, tt)
v_avg <- var_gamma_avg(2, 2, 0, a, b, tt)

cat(sprintf('J_0=2, J_1=2 (sigma_b/sigma = 1):\n'))
cat(sprintf('  GLS:    %.4f\n', v_gls))
cat(sprintf('  ANCOVA: %.4f\n', v_anc))
cat(sprintf('  Avg:    %.4f\n', v_avg))
cat(sprintf('  RE(GLS/ANC): %.3f\n', v_gls / v_anc))
cat(sprintf('  RE(ANC/Avg): %.3f\n', v_anc / v_avg))
```


## SUR (seemingly unrelated regression)

The SUR approach treats the run-in and treatment
periods as separate regression equations linked by
correlated errors:
\begin{align}
  Y_1 &= X_2 \begin{pmatrix}
    \beta_{11} \\ \beta_{12}
  \end{pmatrix} + e_1
  & &\text{(both groups)} \\
  Y_2 &= X_1 \alpha_2 + e_2
  & &\text{(baseline)}
\end{align}

The deviation-from-mean parameterization gives
$\beta_1 = (\delta_1 + \delta_2)/2$ and
$\beta_2 = (\delta_2 - \delta_1)/2$, where
$\delta_j$ are the group-specific rates.

<!-- TODO: derive the SUR variance and compare -->
<!-- to GLS and averaged change-score -->


## BLUP as theoretical benchmark

The three estimators above are linear functions of the
observed change scores and can be compared to a common
benchmark: the best linear unbiased predictor (BLUP) of the
individual random slopes. Under the random-slopes model with
marginal covariance $V_i = R_i + Z_i G Z_i'$, the BLUP of the
subject-specific random slope $b_i$ given the data is
\begin{equation}
  \hat b_i
  = G Z_i' V_i^{-1}
    \bigl(y_i - X_i \hat{\boldsymbol{\beta}}\bigr)
  \label{eq:blup}
\end{equation}
where $\hat{\boldsymbol{\beta}}$ is the GLS estimator of the
fixed effects [@henderson1975; @robinson1991].

The group-specific slope estimators derived from the fixed
effects share the same variance as
$(\hat{\boldsymbol{\beta}})_\gamma$, and the GLS estimator
$\hat\gamma_{\text{GLS}}$ minimizes the variance among all
linear unbiased estimators of $\gamma$ when $V_i$ is correctly
specified. The averaged and ANCOVA estimators are specific
linear approximations to this benchmark:

- **Averaged change-score** replaces $V_i^{-1}$ with an
  unweighted sum; this is optimal only when $V_i \propto I$,
  which fails for any non-zero random-slope variance.
- **ANCOVA** collapses the run-in visits into a single
  scalar covariate (the run-in mean) rather than weighting
  them by $V_i^{-1}$; the resulting efficiency loss vanishes
  as $J_0 \to \infty$ since the run-in mean becomes a
  sufficient statistic for the subject-specific intercept
  plus average slope.

Under model misspecification (Section 4), these
approximations can outperform the nominally optimal GLS
because they depend on fewer variance components.

# Relative efficiency under correct specification

## Analytical comparison

The relative efficiency of the averaged estimator
to GLS is:
\begin{equation}
  \text{RE}_{\text{avg}} =
    \frac{\Var(\hat\gamma_{\text{GLS}})}
         {\Var(\hat\gamma_{\text{avg}})}
    \le 1
\end{equation}
with equality when the averaged estimator happens
to coincide with GLS (which occurs in special cases).

```{r re_table}
configs <- expand.grid(
  r = c(0, 1, 2, 4),
  k = c(1, 2, 3),
  f = c(0, 1)
)
configs <- configs[configs$r + configs$k +
                     configs$f >= 2, ]

ratio_levels <- c(0.5, 1.0, 2.0, 5.0, 20.0)
a <- 1.0

for (ratio in ratio_levels) {
  b <- ratio^2 * a
  configs[[paste0('avg_', ratio)]] <- mapply(
    function(r, k, f)
      re_avg_vs_gls(r, k, f, a, b, tt),
    configs$r, configs$k, configs$f
  )
  configs[[paste0('anc_', ratio)]] <- mapply(
    function(r, k, f)
      re_ancova_vs_gls(r, k, f, a, b, tt),
    configs$r, configs$k, configs$f
  )
}

configs$label <- sprintf('(%d,%d,%d)',
  configs$r, configs$k, configs$f)

avg_cols <- paste0('avg_', ratio_levels)
anc_cols <- paste0('anc_', ratio_levels)

display_avg <- configs[, c('label', avg_cols)]
names(display_avg) <- c('$(J_0,J_1,J_2)$',
  paste0('$', ratio_levels, '$'))

knitr::kable(
  display_avg,
  caption = paste0(
    'RE of averaged change-score vs GLS ',
    '($\\Var_{\\text{GLS}} / ',
    '\\Var_{\\text{avg}}$) by ',
    '$\\sigma_b/\\sigma$.'),
  booktabs = TRUE,
  escape = FALSE,
  digits = 3,
  row.names = FALSE
)

display_anc <- configs[, c('label', anc_cols)]
names(display_anc) <- c('$(J_0,J_1,J_2)$',
  paste0('$', ratio_levels, '$'))

knitr::kable(
  display_anc,
  caption = paste0(
    'RE of ANCOVA vs GLS ',
    '($\\Var_{\\text{GLS}} / ',
    '\\Var_{\\text{ANC}}$) by ',
    '$\\sigma_b/\\sigma$.'),
  booktabs = TRUE,
  escape = FALSE,
  digits = 3,
  row.names = FALSE
)
```

## When does the gap exceed 5%?

<!-- TODO: characterize the parameter space where -->
<!-- RE_avg < 0.95, i.e., where the simpler -->
<!-- approach loses more than 5% efficiency -->


# Robustness to misspecification

## Misspecified $G$

When $\sigma_b^2$ is misestimated (e.g., too large
or too small), the feasible GLS is no longer optimal.
The averaged change-score estimator is unaffected.

```{r misspec_G, fig.cap="Effect of misspecifying $\\sigma_b^2$ on GLS variance. True $\\sigma_b^2 = 1$. The GLS estimator uses an assumed value on the x-axis. The averaged estimator (horizontal line) is unaffected."}
b_true <- 1.0
a <- 1.0
r <- 2
k <- 2
f <- 0

b_assumed <- seq(0.01, 5, by = 0.05)

var_misspec_gls <- function(r, k, f, sigma2,
                            sigma_b2_true,
                            sigma_b2_assumed,
                            t_interval) {
  p <- r + k + f
  R <- build_R_matrix(p, sigma2)
  des <- build_design(r, k, f, t_interval)
  Z <- des$Z
  nc <- des$ncol_X

  Sigma_true <- R + Z %*% matrix(sigma_b2_true) %*% t(Z)

  Ri <- solve(R)
  G_a <- matrix(sigma_b2_assumed)
  H_a <- solve(solve(G_a) + t(Z) %*% Ri %*% Z)
  Xt <- des$X_trt
  Xp <- des$X_plc
  XRX <- t(Xt) %*% Ri %*% Xt + t(Xp) %*% Ri %*% Xp
  XtRZ <- t(Xt) %*% Ri %*% Z
  XpRZ <- t(Xp) %*% Ri %*% Z
  W_a <- XtRZ %*% H_a %*% t(XtRZ) +
         XpRZ %*% H_a %*% t(XpRZ)
  XSiX_a <- XRX - W_a
  XSiX_a_inv <- solve(XSiX_a)

  Sigma_a_inv <- Ri - Ri %*% Z %*% H_a %*% t(Z) %*% Ri

  bread <- XSiX_a_inv
  meat_trt <- t(Xt) %*% Sigma_a_inv %*%
    Sigma_true %*% Sigma_a_inv %*% Xt
  meat_plc <- t(Xp) %*% Sigma_a_inv %*%
    Sigma_true %*% Sigma_a_inv %*% Xp
  meat <- meat_trt + meat_plc

  sandwich <- bread %*% meat %*% bread
  sandwich[nc, nc]
}

v_misspec <- sapply(b_assumed, function(ba) {
  var_misspec_gls(r, k, f, a, b_true, ba, tt)
})

v_avg <- var_gamma_avg(r, k, f, a, b_true, tt)
v_correct <- var_gamma_matrix(r, k, f, a, b_true, tt)

plot(b_assumed, v_misspec, type = 'l', lwd = 2,
     xlab = expression(paste('Assumed ',
                             sigma[b]^2)),
     ylab = expression(Var(hat(gamma))),
     main = 'Misspecified random slope variance',
     ylim = c(min(v_correct, v_avg) * 0.9,
              max(v_misspec) * 1.1))
abline(h = v_avg, col = 'red', lwd = 2, lty = 2)
abline(v = b_true, col = 'gray', lty = 3)
points(b_true, v_correct, pch = 16, cex = 1.5)
legend('topright',
       legend = c('Misspecified GLS (sandwich)',
                  'Averaged change-score',
                  'Correct GLS'),
       col = c('black', 'red', 'black'),
       lty = c(1, 2, NA),
       pch = c(NA, NA, 16),
       lwd = c(2, 2, NA),
       bty = 'n')
```

## Misspecified $R$

When the residual correlation structure is wrong
(e.g., compound symmetry assumed but AR(1) is true),
both the GLS variance formula and the Woodbury
correction are biased. The averaged estimator is
again unaffected.

```{r misspec_R, fig.cap="Effect of analyzing under compound symmetry when the true correlation is AR(1). Sandwich variance of GLS (solid) vs averaged change-score (dashed) as a function of $\\rho$."}
rho_grid <- seq(0, 0.9, by = 0.05)
r <- 2
k <- 2
f <- 0
a <- 1.0
b <- 1.0

var_misspec_r <- function(rho) {
  p <- r + k + f
  des <- build_design(r, k, f, tt)
  Z <- des$Z
  nc <- des$ncol_X

  d_runin <- if (r > 0) seq(-r, -1) else numeric(0)
  d_trt <- if (k > 0) seq(1, k) else numeric(0)
  d_all <- c(d_runin, d_trt)

  R_true <- build_R_matrix_ar1(p, a, rho, d_all)
  Sigma_true <- R_true + Z %*% matrix(b) %*% t(Z)

  R_assumed <- build_R_matrix(p, a)
  Ri_a <- solve(R_assumed)
  G <- matrix(b)
  H_a <- solve(solve(G) + t(Z) %*% Ri_a %*% Z)

  Xt <- des$X_trt
  Xp <- des$X_plc
  XRX <- t(Xt) %*% Ri_a %*% Xt +
         t(Xp) %*% Ri_a %*% Xp
  XtRZ <- t(Xt) %*% Ri_a %*% Z
  XpRZ <- t(Xp) %*% Ri_a %*% Z
  W_a <- XtRZ %*% H_a %*% t(XtRZ) +
         XpRZ %*% H_a %*% t(XpRZ)
  XSiX_a <- XRX - W_a
  XSiX_a_inv <- solve(XSiX_a)

  Sigma_a_inv <- Ri_a - Ri_a %*% Z %*% H_a %*%
                 t(Z) %*% Ri_a

  meat_trt <- t(Xt) %*% Sigma_a_inv %*%
    Sigma_true %*% Sigma_a_inv %*% Xt
  meat_plc <- t(Xp) %*% Sigma_a_inv %*%
    Sigma_true %*% Sigma_a_inv %*% Xp
  meat <- meat_trt + meat_plc

  sandwich <- XSiX_a_inv %*% meat %*% XSiX_a_inv
  sandwich[nc, nc]
}

v_sandwich <- sapply(rho_grid, var_misspec_r)

des <- build_design(r, k, f, tt)
n_post <- k + f
w <- c(rep(-1 / r, r), rep(1 / n_post, k + f))
v_avg_ar1 <- sapply(rho_grid, function(rho) {
  p <- r + k + f
  d_all <- c(seq(-r, -1), seq(1, k))
  if (rho == 0) {
    R_true <- build_R_matrix(p, a)
  } else {
    R_true <- build_R_matrix_ar1(p, a, rho, d_all)
  }
  Z <- des$Z
  Sigma_true <- R_true + Z %*% matrix(b) %*% t(Z)
  2 * as.numeric(t(w) %*% Sigma_true %*% w)
})

plot(rho_grid, v_sandwich, type = 'l', lwd = 2,
     xlab = expression(rho[true]),
     ylab = expression(Var(hat(gamma))),
     main = 'CS assumed, AR(1) true',
     ylim = range(c(v_sandwich, v_avg_ar1)))
lines(rho_grid, v_avg_ar1, col = 'red',
      lwd = 2, lty = 2)
legend('topleft',
       legend = c('GLS (CS assumed, sandwich)',
                  'Averaged change-score'),
       col = c('black', 'red'),
       lty = c(1, 2), lwd = 2, bty = 'n')
```


# Simulation study

## Design

To complement the analytical results, we simulate
trials under the following factorial design:

- $n \in \{30, 50, 100, 200\}$ per group
- $(J_0, J_1, J_2) \in \{(0,2,0), (2,2,0), (2,2,2)\}$
- $\sigma_b/\sigma \in \{1, 5, 20\}$
- True correlation: compound symmetry or AR(1)
- Analysis: GLS (correctly specified), GLS
  (misspecified), averaged change-score

For each configuration, 5000 trials are simulated.
The outcomes are empirical variance, empirical power,
and coverage of nominal 95\% confidence intervals.

```{r simulation}
simulate_trial <- function(m, r, k, f, sigma2,
                           sigma_b2, t_interval,
                           delta) {
  p <- r + k + f
  des <- build_design(r, k, f, t_interval)
  R <- build_R_matrix(p, sigma2)
  Z <- des$Z
  Sigma <- R + Z %*% matrix(sigma_b2) %*% t(Z)
  L <- chol(Sigma)

  generate_group <- function(n, g) {
    X <- if (g == 1) des$X_trt else des$X_plc
    nc <- ncol(X)
    beta_true <- if (nc == 2) {
      c(1.0, delta * g)
    } else {
      c(1.0, 1.0, delta * g)
    }
    mu <- X %*% beta_true
    replicate(n, as.numeric(mu + t(L) %*% rnorm(p)))
  }

  Y_trt <- generate_group(m, 1)
  Y_plc <- generate_group(m, 0)
  list(Y_trt = Y_trt, Y_plc = Y_plc, des = des)
}

analyze_gls <- function(trial, sigma2, sigma_b2,
                        t_interval) {
  des <- trial$des
  p <- des$p
  nc <- des$ncol_X
  R <- build_R_matrix(p, sigma2)
  Ri <- solve(R)
  Z <- des$Z
  G <- matrix(sigma_b2)
  H <- solve(solve(G) + t(Z) %*% Ri %*% Z)

  Xt <- des$X_trt
  Xp <- des$X_plc

  XSi_trt <- t(Xt) %*% (Ri - Ri %*% Z %*% H %*%
                          t(Z) %*% Ri)
  XSi_plc <- t(Xp) %*% (Ri - Ri %*% Z %*% H %*%
                          t(Z) %*% Ri)

  info <- XSi_trt %*% Xt + XSi_plc %*% Xp
  m <- ncol(trial$Y_trt)
  info_total <- m * info
  V <- solve(info_total)

  rhs <- rowSums(apply(trial$Y_trt, 2, function(y) {
    XSi_trt %*% y
  })) + rowSums(apply(trial$Y_plc, 2, function(y) {
    XSi_plc %*% y
  }))

  beta_hat <- V %*% rhs
  se <- sqrt(V[nc, nc])
  gamma_hat <- beta_hat[nc]
  list(estimate = gamma_hat, se = se,
       z = gamma_hat / se)
}

analyze_avg <- function(trial) {
  des <- trial$des
  r <- sum(des$X_trt[, 1] != 0)
  p <- des$p
  n_post <- p - r

  avg_score <- function(y) {
    if (r == 0) {
      mean(y)
    } else {
      mean(y[(r + 1):p]) - mean(y[1:r])
    }
  }

  d_trt <- apply(trial$Y_trt, 2, avg_score)
  d_plc <- apply(trial$Y_plc, 2, avg_score)
  m <- length(d_trt)

  gamma_hat <- mean(d_trt) - mean(d_plc)
  se <- sqrt(var(d_trt) / m + var(d_plc) / m)
  list(estimate = gamma_hat, se = se,
       z = gamma_hat / se)
}

# Quick validation
set.seed(42)
trial <- simulate_trial(100, 2, 2, 0, 1.0, 1.0,
                        1.0, 0.5)
res_gls <- analyze_gls(trial, 1.0, 1.0, 1.0)
res_avg <- analyze_avg(trial)
cat(sprintf('Single trial (m=100, delta=0.5):\n'))
cat(sprintf('  GLS: est=%.3f, se=%.3f\n',
            res_gls$estimate, res_gls$se))
cat(sprintf('  Avg: est=%.3f, se=%.3f\n',
            res_avg$estimate, res_avg$se))
```


# Numerical examples

## MIRIAD parameters

```{r miriad_comparison}
sigma2_ad <- 0.0025
sigma_b2_ad <- 0.9745

configs <- expand.grid(
  r = c(0, 1, 2, 4),
  k = c(1, 2, 3),
  f = c(0, 1)
)
configs <- configs[configs$r + configs$k +
                     configs$f >= 2, ]

configs$v_gls <- mapply(
  function(r, k, f)
    var_gamma_matrix(r, k, f, sigma2_ad,
                     sigma_b2_ad, 1.0),
  configs$r, configs$k, configs$f
)

configs$v_anc <- mapply(
  function(r, k, f)
    var_gamma_ancova(r, k, f, sigma2_ad,
                     sigma_b2_ad, 1.0),
  configs$r, configs$k, configs$f
)

configs$v_avg <- mapply(
  function(r, k, f)
    var_gamma_avg(r, k, f, sigma2_ad,
                  sigma_b2_ad, 1.0),
  configs$r, configs$k, configs$f
)

configs$label <- sprintf('(%d,%d,%d)',
  configs$r, configs$k, configs$f)

delta_ad <- 0.25
za <- qnorm(0.975)
zb <- qnorm(0.90)
configs$n_gls <- ceiling(
  (za + zb)^2 * configs$v_gls / delta_ad^2)
configs$n_anc <- ceiling(
  (za + zb)^2 * configs$v_anc / delta_ad^2)
configs$n_avg <- ceiling(
  (za + zb)^2 * configs$v_avg / delta_ad^2)

display <- configs[order(configs$v_gls),
  c('label', 'n_gls', 'n_anc', 'n_avg')]

knitr::kable(
  head(display, 15),
  col.names = c('$(J_0,J_1,J_2)$', '$N$ GLS', '$N$ ANC',
                '$N$ Avg'),
  caption = paste0(
    'Sample size per group: GLS, ANCOVA, and ',
    'averaged estimators (MIRIAD parameters, ',
    '$\\Delta=', delta_ad, '$, 90\\% power).'),
  booktabs = TRUE,
  escape = FALSE,
  row.names = FALSE
)
```

## When to use which estimator

<!-- TODO: decision flowchart or rule based on -->
<!-- n, sigma_b/sigma, and confidence in the -->
<!-- correlation structure -->


# Discussion

<!-- TODO: practical guidance. Key messages: -->
<!-- (1) GLS is preferred when the model is -->
<!--     well-specified and n is moderate-large -->
<!-- (2) The averaged approach is a safe default -->
<!--     when variance components are uncertain -->
<!-- (3) The efficiency gap is small when r and k -->
<!--     are large (because the averaging already -->
<!--     captures most of the slope information) -->
<!-- (4) SUR provides a middle ground but adds -->
<!--     implementation complexity -->

# Conflict of interest

The author declares no conflicts of interest.

# Funding

This work received no specific funding.

# Data availability statement

The `runinpower` R package implementing the GLS, ANCOVA, and
averaged-change-score estimators (`ancova.R`, `averaged.R`),
the simulation drivers, and the shared bibliography across the
four-paper compendium are openly available in the project
repository. Specific paths are listed in the Reproducibility
section below.

# Reproducibility

This manuscript was rendered with R 4.4.x. Principal packages
used are the in-package `runinpower` source, `nlme` (the GLS
fitter for the simulation comparator), `tinytest` (testing
harness), and `rmarkdown` plus `bookdown` for the manuscript
build.

**Random-number generator.** The Monte Carlo simulation
comparing the three estimators sets `set.seed(20260418)` at
its entry point. Per-replicate seeds derive from the master
seed by `+ rep_idx`.

**Data and code availability.** The R package source is at
`R/`. Simulation drivers are at `analysis/scripts/`. The
companion manuscripts (Papers 1, 2, 3) are at
`analysis/paper{1,2,3}-*/`. The manuscript source and shared
bibliography (`../references.bib`) are at
`analysis/paper4-gls-vs-ttest/`.

# References
