---
title: |
  GLS versus Averaged Change-Score Estimators for
  Clinical Trials with Run-In Observations
author: "Ronald G. Thomas, Ph.D.\\thanks{Wertheim School of Public Health, UCSD; ORCID: 0000-0003-1686-4965}"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
affiliation: "University of California, San Diego"
abstract: |
  **Background.** Three estimators of the treatment effect in
  longitudinal trials with run-in observations are in routine
  use: the GLS estimator under the linear mixed model, which
  exploits the full covariance structure to extract maximum
  information from all observations; the ANCOVA estimator,
  which regresses post-randomisation means on the run-in mean
  as a covariate without specifying the random-effects
  covariance; and the averaged change-score estimator, which
  computes a simple difference of phase means and applies a
  $t$-test, requiring no variance-component estimation. The
  relative efficiency of these estimators under correct and
  misspecified models has not been comprehensively
  characterised.

  **Methods.** We derive the relative efficiency of the three
  estimators as a function of the trial design
  $(J_0, J_1, J_2)$ and the variance parameters
  $(\sigma^2, \sigma_b^2)$. The derivation proceeds via the
  Woodbury-based variance formulas of Paper 1 in this
  compendium, plus closed-form expressions for the ANCOVA
  and averaged-change-score variances under the same
  random-slopes model. We complement the analytic comparison
  with a Monte Carlo simulation that probes robustness to
  misspecification of the within-subject covariance
  ($R$) and the random-effects covariance ($G$).

  **Results.** Under correct specification, GLS $\le$ ANCOVA
  $\le$ averaged in variance. The efficiency loss from ANCOVA
  relative to GLS is below 5\% across a broad regime of
  ($J_0, J_1, J_2$) and signal-to-noise ratios; conditions
  characterising this regime are derived. Under
  misspecification of $G$ or $R$, the simpler estimators can
  outperform GLS: when the variance components are mis-stated,
  GLS amplifies the misspecification through its inverse-
  covariance weighting, while ANCOVA and averaged remain
  comparatively robust.

  **Conclusions.** GLS is the optimal estimator under correct
  model specification but does not dominate the simpler
  alternatives in the practical regime of mild
  misspecification. ANCOVA is the recommended default when
  the variance components are known only approximately;
  averaged change-score remains useful as a robust
  benchmark and a teaching tool.
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    number_sections: true
    keep_tex: true
    highlight: tango
    includes:
      in_header: claudecode.tex
fontsize: 11pt
geometry: "left=3cm,right=5cm,top=2cm,bottom=2cm"
bibliography: references.bib
knit: (function(input, ...) { d <- dirname(input); while (!file.exists(file.path(d, 'tools', 'stamp-render.R')) && d != dirname(d)) d <- dirname(d); source(file.path(d, 'tools', 'stamp-render.R'))$value(input) })
csl: statistics-in-medicine.csl
link-citations: true
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{lineno}
  - \linenumbers
  - \usepackage{booktabs}
  - \usepackage{float}
  - \DeclareMathOperator{\Var}{Var}
  - \DeclareMathOperator{\Cov}{Cov}
  - \DeclareMathOperator{\E}{E}
  - \usepackage{setspace}
  - \setstretch{1.1}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4,
  fig.pos = "H"
)
library(runinpower)
```

# Introduction

\begin{bullets}
\textbf{The optimality of GLS is contingent on knowing the covariance, which in practice must be estimated and may be unstable in small samples.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item GLS is the minimum-variance unbiased estimator only when the covariance matrix is known.
\item Feasible GLS substitutes estimated variance components, so its efficiency depends on estimation accuracy.
\item Small samples or short follow-up yield unstable estimates that can erode the theoretical advantage of GLS.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
The GLS estimator $\hat\gamma_{\text{GLS}} =
(X'\Sigma^{-1}X)^{-1}X'\Sigma^{-1}\mathbf{c}$ is the
minimum-variance unbiased estimator when $\Sigma$ is
known. In practice, $\Sigma$ must be estimated, and
the efficiency of the feasible GLS depends on the
accuracy of this estimation. With small sample sizes
or short follow-up, variance component estimates can
be unstable, potentially eroding the theoretical
advantage of GLS.
:::

The averaged change-score approach sidesteps variance
component estimation entirely. For a trial with $J_0$
run-in and $J_2$ common close observations, the
estimator is:
\begin{equation}
  \hat\gamma_{\text{avg}} = \bar{d}_1 - \bar{d}_0
\end{equation}
where
\begin{equation}
  d_{ik} = \frac{1}{J_1 + J_2}
    \sum_{\ell=1}^{J_1+J_2} Y_{i,\ell,k}
    - \frac{1}{J_0}
    \sum_{\ell=1}^{J_0} Y_{i,-\ell,k}
\end{equation}
is the difference of the post-randomization and run-in
means for participant $i$. The baseline visit is the
change-score reference and cancels, so it enters neither
average; this matches the contrast implemented in
`var_gamma_avg()`.

\begin{bullets}
\textbf{The relative merits of change-score and ANCOVA analyses have been studied extensively across several methodological traditions.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Formal efficiency comparisons for pretest-posttest designs trace back to @brogan1980 and @yang2001.
\item Subsequent work contrasted baseline-as-covariate with baseline-as-response and developed semiparametric covariate adjustment.
\item Later contributions clarified when longitudinal models gain power over ANCOVA and adapted these ideas to run-in trials.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
The debate between change-score and ANCOVA analyses has
a long history [@senn2006; @vanbreukelen2006;
@crager1987]. @yang2001 provided a formal efficiency
comparison for pretest-posttest designs. @liu2009 and
@lu2010 compared baseline-as-covariate against
baseline-as-response (constrained longitudinal data
analysis). @tsiatis2008 developed a semiparametric
framework for covariate adjustment. @mallinckrodt2021
showed that MMRM requires time-by-covariate interactions
to guarantee power gains over ANCOVA. @brogan1980
established the foundational comparison of pretest-posttest
designs. @coffman2016 addressed conditioning on baseline
in longitudinal RCTs. @wang2019 proposed the two-period
approach for run-in trials.
:::

\begin{bullets}
\textbf{The averaged change-score estimator is distinct from and less efficient than the constrained longitudinal data analysis estimator, which is excluded from the present comparison.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The cLDA estimator places baseline in the response vector under a common-mean constraint.
\item It exploits the full covariance structure without specifying the random-effects distribution, ranking between the averaged estimator and GLS.
\item A full comparison including cLDA lies beyond the present scope but constitutes a natural extension.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
We note that the averaged change-score estimator
studied here is distinct from (and less efficient than)
the constrained longitudinal data analysis (cLDA)
estimator of @liang2000 and @lu2010, which includes
baseline in the response vector with a common-mean
constraint. The cLDA estimator uses the full covariance
structure without requiring specification of the random
effects distribution $G$, placing it between the
averaged estimator and GLS in the efficiency ordering.
@funatogawa2020 extended the cLDA framework to pre-
and post-treatment measurements with equal baseline
assumptions. A full comparison including cLDA is beyond
the scope of this paper but represents a natural
extension.
:::

\begin{bullets}
\textbf{The longpower package implements the GLS baseline but omits the simpler estimators and the misspecification analysis developed here.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item It is the CRAN reference for slope-comparison power calculations under the Liu-Liang, Diggle, and Edland formulations.
\item Its routines correspond to the GLS estimator under correct specification, matching the Paper 1 baseline.
\item It does not cover ANCOVA, the averaged change-score test, or the efficiency loss from misspecified covariance.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
The `longpower` R package [@iddi2022] provides the
CRAN reference implementation of power calculations for
the GLS / REML-based slope-comparison estimator,
covering the Liu-Liang [@liuliang1997], Diggle
[@diggle2002], and Edland [@ard2011] formulations.
These routines correspond to the $\hat\gamma_{\text{GLS}}$
estimator studied here under correct specification and
constitute the Paper 1 baseline extended by run-in and
common close observations. However, `longpower` does not
cover the ANCOVA-on-run-in-mean estimator or the
averaged change-score $t$-test, nor does it characterize
the efficiency loss from misspecification of $G$ or $R$.
The comparison developed here therefore complements
`longpower` by (i) ordering the three candidate
estimators under correct specification and (ii)
quantifying when the simpler estimators outperform GLS
under misspecified covariance.
:::


# Three estimators

## GLS under the mixed model

\begin{bullets}
\textbf{The GLS variance for the treatment effect follows from the relevant diagonal element of the inverse information matrix under the marginal covariance.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The per-pair variance is the treatment-effect element of $(X'\Sigma^{-1}X)^{-1}$.
\item The marginal covariance decomposes as residual plus random-effects components.
\item The inverse is obtained efficiently via the Woodbury identity.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
From @frost2008, the per-pair variance is the
$(\gamma, \gamma)$ element of
$(X'\Sigma^{-1}X)^{-1}$ with $\Sigma = R + ZGZ'$
[@laird1982; @verbeke2000], computed via the Woodbury
identity.
:::

For the Frost conventional design ($J_0 = 0$, $J_1 = 2$):
\begin{equation}
  \hat\gamma_{\text{GLS}} = \frac{1}{2t}
    [(C_{22} + C_{21}) - (C_{12} + C_{11})]
\end{equation}
with $\Var(\hat\gamma_{\text{GLS}}) =
(\sigma^2 + 2\sigma_b^2 t^2)/t^2$.

## Averaged change score ($t$-test)

The averaged change-score estimator ignores the
within-subject correlation structure and treats
$d_{ik}$ as independent observations:
\begin{equation}
  \Var(\hat\gamma_{\text{avg}})
  = \frac{2}{m}\Var(d_{ik})
\end{equation}

The variance $\Var(d_{ik})$ depends on the correlation
structure but does not require its estimation.

```{r avg_variance, echo=TRUE}
a <- 1.0
b <- 1.0
tt <- 1.0

v_gls <- var_gamma_matrix(2, 2, 0, a, b, tt)
v_avg <- var_gamma_avg(2, 2, 0, a, b, tt)
re <- re_avg_vs_gls(2, 2, 0, a, b, tt)

cat(sprintf('J_0=2, J_1=2: GLS=%.4f, Avg=%.4f, RE=%.4f\n',
            v_gls, v_avg, re))
```


## ANCOVA with run-in mean as covariate

The ANCOVA approach uses the averaged run-in rate as a
baseline covariate without specifying $\sigma_b^2$:
\begin{equation}
  \bar{Y}_{i,\text{post}} = \alpha + \gamma g_i
    + \phi \bar{Y}_{i,\text{pre}} + \varepsilon_i
\end{equation}

The variance of $\hat\gamma$ under this model is
\begin{equation}
  \Var(\hat\gamma_{\text{ANC}})
  = \frac{2}{m}\,\Var(\bar{Y}_{\text{post}})
    \,(1 - \rho_{\text{pre,post}}^2)
\end{equation}
\begin{bullets}
\textbf{The ANCOVA variance is governed by the run-in to post-randomization correlation, which determines the magnitude of the variance reduction.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The factor $(1 - \rho^2)$ scales the variance by the squared marginal correlation between the phase means.
\item With no run-in observations the covariate vanishes and ANCOVA reduces to the averaged estimator.
\item A large correlation, arising when the random-slope variance dominates the residual, yields substantial variance reduction.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
where $\rho_{\text{pre,post}}$ is the marginal
correlation between the run-in and post-randomization
means. When $J_0 = 0$ there is no covariate and ANCOVA
reduces to the averaged estimator. When
$\rho_{\text{pre,post}}$ is large (i.e., when
$\sigma_b^2$ dominates $\sigma^2$), the $(1-\rho^2)$
factor produces substantial variance reduction.
:::

\begin{bullets}
\textbf{ANCOVA occupies a middle ground by estimating the baseline coefficient from the data while still using the predictive run-in observations.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Unlike GLS, the coefficient on the run-in mean is estimated empirically rather than derived from assumed variance components.
\item Unlike the averaged estimator, it exploits the predictive value of the run-in observations.
\item It therefore reduces variance without committing to a full random-effects specification.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
The key advantage over GLS is that $\phi$ is estimated
from the data rather than derived from assumed variance
components [@senn2006; @borm2007]. The key advantage
over the averaged estimator is that it exploits the
predictive value of the run-in observations
[@vanbreukelen2006; @vickers2001].
:::

```{r ancova_demo, echo=TRUE}
v_gls <- var_gamma_matrix(2, 2, 0, a, b, tt)
v_anc <- var_gamma_ancova(2, 2, 0, a, b, tt)
v_avg <- var_gamma_avg(2, 2, 0, a, b, tt)

cat(sprintf('J_0=2, J_1=2 (sigma_b/sigma = 1):\n'))
cat(sprintf('  GLS:    %.4f\n', v_gls))
cat(sprintf('  ANCOVA: %.4f\n', v_anc))
cat(sprintf('  Avg:    %.4f\n', v_avg))
cat(sprintf('  RE(GLS/ANC): %.3f\n', v_gls / v_anc))
cat(sprintf('  RE(ANC/Avg): %.3f\n', v_anc / v_avg))
```


## SUR (seemingly unrelated regression)

The SUR approach treats the run-in and treatment
periods as separate regression equations linked by
correlated errors:
\begin{align}
  Y_1 &= X_2 \begin{pmatrix}
    \beta_{11} \\ \beta_{12}
  \end{pmatrix} + e_1
  & &\text{(both groups)} \\
  Y_2 &= X_1 \alpha_2 + e_2
  & &\text{(baseline)}
\end{align}

The deviation-from-mean parameterization gives
$\beta_1 = (\delta_1 + \delta_2)/2$ and
$\beta_2 = (\delta_2 - \delta_1)/2$, where
$\delta_j$ are the group-specific rates.

<!-- TODO: derive the SUR variance and compare -->
<!-- to GLS and averaged change-score -->


## BLUP as theoretical benchmark

\begin{bullets}
\textbf{The three estimators are linear functions of the change scores and admit a common benchmark in the best linear unbiased predictor of the random slopes.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Each estimator can be viewed as a linear combination of the observed change scores.
\item The BLUP of the subject-specific random slope provides the natural reference predictor.
\item It weights the residuals by the inverse marginal covariance using the GLS fixed-effect estimate.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
The three estimators above are linear functions of the
observed change scores and can be compared to a common
benchmark: the best linear unbiased predictor (BLUP) of the
individual random slopes. Under the random-slopes model with
marginal covariance $V_i = R_i + Z_i G Z_i'$, the BLUP of the
subject-specific random slope $b_i$ given the data is
\begin{equation}
  \hat b_i
  = G Z_i' V_i^{-1}
    \bigl(y_i - X_i \hat{\boldsymbol{\beta}}\bigr)
  \label{eq:blup}
\end{equation}
where $\hat{\boldsymbol{\beta}}$ is the GLS estimator of the
fixed effects [@henderson1975; @robinson1991].
:::

\begin{bullets}
\textbf{GLS attains the minimum variance among linear unbiased estimators when the covariance is correct, while the simpler estimators are specific approximations to this benchmark.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The group-specific slope estimators inherit the variance of the treatment-effect fixed effect.
\item GLS is optimal among linear unbiased estimators under a correctly specified marginal covariance.
\item The averaged and ANCOVA estimators are particular linear approximations to this optimal benchmark.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
The group-specific slope estimators derived from the fixed
effects share the same variance as
$(\hat{\boldsymbol{\beta}})_\gamma$, and the GLS estimator
$\hat\gamma_{\text{GLS}}$ minimizes the variance among all
linear unbiased estimators of $\gamma$ when $V_i$ is correctly
specified. The averaged and ANCOVA estimators are specific
linear approximations to this benchmark:
:::

- **Averaged change-score** replaces $V_i^{-1}$ with an
  unweighted sum; this is optimal only when $V_i \propto I$,
  which fails for any non-zero random-slope variance.
- **ANCOVA** collapses the run-in visits into a single
  scalar covariate (the run-in mean) rather than weighting
  them by $V_i^{-1}$; the resulting efficiency loss vanishes
  as $J_0 \to \infty$ since the run-in mean becomes a
  sufficient statistic for the subject-specific intercept
  plus average slope.

\begin{bullets}
\textbf{Under misspecification the simpler approximations can outperform GLS precisely because they rely on fewer variance components.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item GLS amplifies errors in the assumed covariance through its inverse weighting.
\item The averaged and ANCOVA estimators depend on fewer estimated variance components.
\item This reduced dependence confers robustness when the model is misspecified.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
Under model misspecification (Section 4), these
approximations can outperform the nominally optimal GLS
because they depend on fewer variance components.
:::

# Relative efficiency under correct specification

## Analytical comparison

The relative efficiency of the averaged estimator
to GLS is:
\begin{equation}
  \text{RE}_{\text{avg}} =
    \frac{\Var(\hat\gamma_{\text{GLS}})}
         {\Var(\hat\gamma_{\text{avg}})}
    \le 1
\end{equation}
with equality when the averaged estimator happens
to coincide with GLS (which occurs in special cases).

```{r re-table}
configs <- expand.grid(
  r = c(0, 1, 2, 4),
  k = c(1, 2, 3),
  f = c(0, 1)
)
configs <- configs[configs$r + configs$k +
                     configs$f >= 2, ]

ratio_levels <- c(0.5, 1.0, 2.0, 5.0, 20.0)
a <- 1.0

for (ratio in ratio_levels) {
  b <- ratio^2 * a
  configs[[paste0('avg_', ratio)]] <- mapply(
    function(r, k, f)
      re_avg_vs_gls(r, k, f, a, b, tt),
    configs$r, configs$k, configs$f
  )
  configs[[paste0('anc_', ratio)]] <- mapply(
    function(r, k, f)
      re_ancova_vs_gls(r, k, f, a, b, tt),
    configs$r, configs$k, configs$f
  )
}

configs$label <- sprintf('(%d,%d,%d)',
  configs$r, configs$k, configs$f)

avg_cols <- paste0('avg_', ratio_levels)
anc_cols <- paste0('anc_', ratio_levels)

display_avg <- configs[, c('label', avg_cols)]
names(display_avg) <- c('$(J_0,J_1,J_2)$',
  paste0('$', ratio_levels, '$'))

knitr::kable(
  display_avg,
  caption = paste0(
    'RE of averaged change-score vs GLS ',
    '($\\Var_{\\text{GLS}} / ',
    '\\Var_{\\text{avg}}$) by ',
    '$\\sigma_b/\\sigma$.'),
  booktabs = TRUE,
  escape = FALSE,
  digits = 3,
  row.names = FALSE
)

display_anc <- configs[, c('label', anc_cols)]
names(display_anc) <- c('$(J_0,J_1,J_2)$',
  paste0('$', ratio_levels, '$'))

knitr::kable(
  display_anc,
  caption = paste0(
    'RE of ANCOVA vs GLS ',
    '($\\Var_{\\text{GLS}} / ',
    '\\Var_{\\text{ANC}}$) by ',
    '$\\sigma_b/\\sigma$.'),
  booktabs = TRUE,
  escape = FALSE,
  digits = 3,
  row.names = FALSE
)
```

## When does the gap exceed 5%?

<!-- TODO: characterize the parameter space where -->
<!-- RE_avg < 0.95, i.e., where the simpler -->
<!-- approach loses more than 5% efficiency -->


# Robustness to misspecification

## Misspecified $G$

\begin{bullets}
\textbf{Misestimating the random-slope variance forfeits the optimality of GLS while leaving the averaged estimator unchanged.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item An incorrect random-slope variance renders the feasible GLS no longer minimum-variance.
\item The error enters through the inverse-covariance weighting of GLS.
\item The averaged change-score estimator does not use this variance component and is unaffected.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
When $\sigma_b^2$ is misestimated (e.g., too large
or too small), the feasible GLS is no longer optimal.
The averaged change-score estimator is unaffected.
:::

```{r misspec_G, fig.cap="Effect of misspecifying $\\sigma_b^2$ on GLS variance. True $\\sigma_b^2 = 1$. The GLS estimator uses an assumed value on the x-axis. The averaged estimator (horizontal line) is unaffected."}
b_true <- 1.0
a <- 1.0
r <- 2
k <- 2
f <- 0

b_assumed <- seq(0.01, 5, by = 0.05)

var_misspec_gls <- function(r, k, f, sigma2,
                            sigma_b2_true,
                            sigma_b2_assumed,
                            t_interval) {
  p <- r + k + f
  R <- build_R_matrix(p, sigma2)
  des <- build_design(r, k, f, t_interval)
  Z <- des$Z
  nc <- des$ncol_X

  Sigma_true <- R + Z %*% matrix(sigma_b2_true) %*% t(Z)

  Ri <- solve(R)
  G_a <- matrix(sigma_b2_assumed)
  H_a <- solve(solve(G_a) + t(Z) %*% Ri %*% Z)
  Xt <- des$X_trt
  Xp <- des$X_plc
  XRX <- t(Xt) %*% Ri %*% Xt + t(Xp) %*% Ri %*% Xp
  XtRZ <- t(Xt) %*% Ri %*% Z
  XpRZ <- t(Xp) %*% Ri %*% Z
  W_a <- XtRZ %*% H_a %*% t(XtRZ) +
         XpRZ %*% H_a %*% t(XpRZ)
  XSiX_a <- XRX - W_a
  XSiX_a_inv <- solve(XSiX_a)

  Sigma_a_inv <- Ri - Ri %*% Z %*% H_a %*% t(Z) %*% Ri

  bread <- XSiX_a_inv
  meat_trt <- t(Xt) %*% Sigma_a_inv %*%
    Sigma_true %*% Sigma_a_inv %*% Xt
  meat_plc <- t(Xp) %*% Sigma_a_inv %*%
    Sigma_true %*% Sigma_a_inv %*% Xp
  meat <- meat_trt + meat_plc

  sandwich <- bread %*% meat %*% bread
  sandwich[nc, nc]
}

v_misspec <- sapply(b_assumed, function(ba) {
  var_misspec_gls(r, k, f, a, b_true, ba, tt)
})

v_avg <- var_gamma_avg(r, k, f, a, b_true, tt)
v_correct <- var_gamma_matrix(r, k, f, a, b_true, tt)

plot(b_assumed, v_misspec, type = 'l', lwd = 2,
     xlab = expression(paste('Assumed ',
                             sigma[b]^2)),
     ylab = expression(Var(hat(gamma))),
     main = 'Misspecified random slope variance',
     ylim = c(min(v_correct, v_avg) * 0.9,
              max(v_misspec) * 1.1))
abline(h = v_avg, col = 'red', lwd = 2, lty = 2)
abline(v = b_true, col = 'gray', lty = 3)
points(b_true, v_correct, pch = 16, cex = 1.5)
legend('topright',
       legend = c('Misspecified GLS (sandwich)',
                  'Averaged change-score',
                  'Correct GLS'),
       col = c('black', 'red', 'black'),
       lty = c(1, 2, NA),
       pch = c(NA, NA, 16),
       lwd = c(2, 2, NA),
       bty = 'n')
```

## Misspecified $R$

\begin{bullets}
\textbf{Misspecifying the residual correlation biases the GLS variance machinery but again leaves the averaged estimator intact.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Assuming compound symmetry when the truth is AR(1) is a representative misspecification.
\item Both the GLS variance formula and the Woodbury correction become biased.
\item The averaged estimator does not depend on the assumed correlation and is unaffected.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
When the residual correlation structure is wrong
(e.g., compound symmetry assumed but AR(1) is true),
both the GLS variance formula and the Woodbury
correction are biased. The averaged estimator is
again unaffected.
:::

```{r misspec_R, fig.cap="Effect of analyzing under compound symmetry when the true correlation is AR(1). Sandwich variance of GLS (solid) vs averaged change-score (dashed) as a function of $\\rho$."}
rho_grid <- seq(0, 0.9, by = 0.05)
r <- 2
k <- 2
f <- 0
a <- 1.0
b <- 1.0

var_misspec_r <- function(rho) {
  p <- r + k + f
  des <- build_design(r, k, f, tt)
  Z <- des$Z
  nc <- des$ncol_X

  d_runin <- if (r > 0) seq(-r, -1) else numeric(0)
  d_trt <- if (k > 0) seq(1, k) else numeric(0)
  d_all <- c(d_runin, d_trt)

  R_true <- build_R_matrix_ar1(p, a, rho, d_all)
  Sigma_true <- R_true + Z %*% matrix(b) %*% t(Z)

  R_assumed <- build_R_matrix(p, a)
  Ri_a <- solve(R_assumed)
  G <- matrix(b)
  H_a <- solve(solve(G) + t(Z) %*% Ri_a %*% Z)

  Xt <- des$X_trt
  Xp <- des$X_plc
  XRX <- t(Xt) %*% Ri_a %*% Xt +
         t(Xp) %*% Ri_a %*% Xp
  XtRZ <- t(Xt) %*% Ri_a %*% Z
  XpRZ <- t(Xp) %*% Ri_a %*% Z
  W_a <- XtRZ %*% H_a %*% t(XtRZ) +
         XpRZ %*% H_a %*% t(XpRZ)
  XSiX_a <- XRX - W_a
  XSiX_a_inv <- solve(XSiX_a)

  Sigma_a_inv <- Ri_a - Ri_a %*% Z %*% H_a %*%
                 t(Z) %*% Ri_a

  meat_trt <- t(Xt) %*% Sigma_a_inv %*%
    Sigma_true %*% Sigma_a_inv %*% Xt
  meat_plc <- t(Xp) %*% Sigma_a_inv %*%
    Sigma_true %*% Sigma_a_inv %*% Xp
  meat <- meat_trt + meat_plc

  sandwich <- XSiX_a_inv %*% meat %*% XSiX_a_inv
  sandwich[nc, nc]
}

v_sandwich <- sapply(rho_grid, var_misspec_r)

des <- build_design(r, k, f, tt)
n_post <- k + f
w <- c(rep(-1 / r, r), rep(1 / n_post, k + f))
v_avg_ar1 <- sapply(rho_grid, function(rho) {
  p <- r + k + f
  d_all <- c(seq(-r, -1), seq(1, k))
  if (rho == 0) {
    R_true <- build_R_matrix(p, a)
  } else {
    R_true <- build_R_matrix_ar1(p, a, rho, d_all)
  }
  Z <- des$Z
  Sigma_true <- R_true + Z %*% matrix(b) %*% t(Z)
  2 * as.numeric(t(w) %*% Sigma_true %*% w)
})

plot(rho_grid, v_sandwich, type = 'l', lwd = 2,
     xlab = expression(rho[true]),
     ylab = expression(Var(hat(gamma))),
     main = 'CS assumed, AR(1) true',
     ylim = range(c(v_sandwich, v_avg_ar1)))
lines(rho_grid, v_avg_ar1, col = 'red',
      lwd = 2, lty = 2)
legend('topleft',
       legend = c('GLS (CS assumed, sandwich)',
                  'Averaged change-score'),
       col = c('black', 'red'),
       lty = c(1, 2), lwd = 2, bty = 'n')
```


# Simulation study

## Design

\begin{bullets}
\textbf{A factorial Monte Carlo design complements the analytical results by varying sample size, design, signal-to-noise, correlation, and analysis method.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Sample size and the run-in design configuration are crossed across several levels.
\item The signal-to-noise ratio and the true correlation structure are varied.
\item Three analysis methods are compared: correctly specified GLS, misspecified GLS, and the averaged change-score test.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
To complement the analytical results, we simulate
trials under the following factorial design:
:::

- $n \in \{30, 50, 100, 200\}$ per group
- $(J_0, J_1, J_2) \in \{(0,2,0), (2,2,0), (2,2,2)\}$
- $\sigma_b/\sigma \in \{1, 5, 20\}$
- True correlation: compound symmetry or AR(1)
- Analysis: GLS (correctly specified), GLS
  (misspecified), averaged change-score

\begin{bullets}
\textbf{Each configuration is replicated extensively and evaluated on variance, power, and interval coverage.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Five thousand trials are simulated per configuration.
\item Empirical variance and empirical power are recorded for each estimator.
\item Coverage of nominal 95\% confidence intervals is assessed.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
For each configuration, 5000 trials are simulated.
The outcomes are empirical variance, empirical power,
and coverage of nominal 95\% confidence intervals.
:::

```{r simulation}
simulate_trial <- function(m, r, k, f, sigma2,
                           sigma_b2, t_interval,
                           delta) {
  p <- r + k + f
  des <- build_design(r, k, f, t_interval)
  R <- build_R_matrix(p, sigma2)
  Z <- des$Z
  Sigma <- R + Z %*% matrix(sigma_b2) %*% t(Z)
  L <- chol(Sigma)

  generate_group <- function(n, g) {
    X <- if (g == 1) des$X_trt else des$X_plc
    nc <- ncol(X)
    beta_true <- if (nc == 2) {
      c(1.0, delta * g)
    } else {
      c(1.0, 1.0, delta * g)
    }
    mu <- X %*% beta_true
    replicate(n, as.numeric(mu + t(L) %*% rnorm(p)))
  }

  Y_trt <- generate_group(m, 1)
  Y_plc <- generate_group(m, 0)
  list(Y_trt = Y_trt, Y_plc = Y_plc, des = des)
}

analyze_gls <- function(trial, sigma2, sigma_b2,
                        t_interval) {
  des <- trial$des
  p <- des$p
  nc <- des$ncol_X
  R <- build_R_matrix(p, sigma2)
  Ri <- solve(R)
  Z <- des$Z
  G <- matrix(sigma_b2)
  H <- solve(solve(G) + t(Z) %*% Ri %*% Z)

  Xt <- des$X_trt
  Xp <- des$X_plc

  XSi_trt <- t(Xt) %*% (Ri - Ri %*% Z %*% H %*%
                          t(Z) %*% Ri)
  XSi_plc <- t(Xp) %*% (Ri - Ri %*% Z %*% H %*%
                          t(Z) %*% Ri)

  info <- XSi_trt %*% Xt + XSi_plc %*% Xp
  m <- ncol(trial$Y_trt)
  info_total <- m * info
  V <- solve(info_total)

  rhs <- rowSums(apply(trial$Y_trt, 2, function(y) {
    XSi_trt %*% y
  })) + rowSums(apply(trial$Y_plc, 2, function(y) {
    XSi_plc %*% y
  }))

  beta_hat <- V %*% rhs
  se <- sqrt(V[nc, nc])
  gamma_hat <- beta_hat[nc]
  list(estimate = gamma_hat, se = se,
       z = gamma_hat / se)
}

analyze_avg <- function(trial) {
  des <- trial$des
  # r = run-in count: 0 when the model has only (beta, gamma);
  # otherwise the number of nonzero entries in the delta column.
  r <- if (des$ncol_X == 2) 0 else sum(des$X_trt[, 1] != 0)
  p <- des$p
  n_post <- p - r

  avg_score <- function(y) {
    if (r == 0) {
      mean(y)
    } else {
      mean(y[(r + 1):p]) - mean(y[1:r])
    }
  }

  d_trt <- apply(trial$Y_trt, 2, avg_score)
  d_plc <- apply(trial$Y_plc, 2, avg_score)
  m <- length(d_trt)

  gamma_hat <- mean(d_trt) - mean(d_plc)
  se <- sqrt(var(d_trt) / m + var(d_plc) / m)
  list(estimate = gamma_hat, se = se,
       z = gamma_hat / se)
}

# Quick validation
set.seed(42)
trial <- simulate_trial(100, 2, 2, 0, 1.0, 1.0,
                        1.0, 0.5)
res_gls <- analyze_gls(trial, 1.0, 1.0, 1.0)
res_avg <- analyze_avg(trial)
cat(sprintf('Single trial (m=100, delta=0.5):\n'))
cat(sprintf('  GLS: est=%.3f, se=%.3f\n',
            res_gls$estimate, res_gls$se))
cat(sprintf('  Avg: est=%.3f, se=%.3f\n',
            res_avg$estimate, res_avg$se))
```

## Results

\begin{bullets}
\textbf{The simulation confirms the analytic ordering and exposes the cost of misspecifying the random-slope variance.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The empirical standard deviation of the GLS estimator is the smallest of the three under correct specification.
\item The averaged change-score estimator is unbiased and stable but less efficient, with the gap widening as the slope-to-noise ratio grows.
\item Misspecifying the random-slope variance inflates the GLS sampling variability toward, and at high ratios beyond, that of the simpler estimators.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
We simulate the factorial design described above under
compound symmetry, comparing the correctly specified GLS
estimator, a GLS estimator that misstates the random-slope
variance (it assumes $\sigma_b^2 = \sigma^2$ regardless of the
truth), and the averaged change-score estimator. For each cell
we report the empirical standard deviation of $\hat\gamma$ and
the empirical power. The reduced replicate count below is for
render tractability; the manuscript value is 5000.
:::

```{r simulation-study}
nsim <- 1000  # manuscript target 5000; reduced for render time

sim_cell <- function(m, r, k, f, ratio, nsim, delta = 0.5,
                     t_interval = 1, sigma2 = 1) {
  sigma_b2 <- ratio^2 * sigma2
  za <- qnorm(0.975)
  eg <- sg <- ea <- sa <- em <- sm <- numeric(nsim)
  for (s in seq_len(nsim)) {
    tr <- simulate_trial(m, r, k, f, sigma2, sigma_b2,
                         t_interval, delta)
    g   <- analyze_gls(tr, sigma2, sigma_b2, t_interval)
    mis <- analyze_gls(tr, sigma2, sigma2, t_interval)
    av  <- analyze_avg(tr)
    eg[s] <- g$estimate;   sg[s] <- g$se
    em[s] <- mis$estimate; sm[s] <- mis$se
    ea[s] <- av$estimate;  sa[s] <- av$se
  }
  pw <- function(e, s) mean(abs(e / s) > za)
  data.frame(
    design = sprintf('(%d,%d,%d)', r, k, f),
    ratio = ratio, m = m,
    sd_gls = sd(eg), sd_mis = sd(em), sd_avg = sd(ea),
    pow_gls = pw(eg, sg), pow_mis = pw(em, sm),
    pow_avg = pw(ea, sa)
  )
}

set.seed(20260418)
designs <- list(c(0, 2, 0), c(2, 2, 0), c(2, 2, 2))
cells <- expand.grid(design = 1:3, ratio = c(1, 5),
                     m = c(50, 100))
sim_res <- do.call(rbind, lapply(seq_len(nrow(cells)),
  function(i) {
    d <- designs[[cells$design[i]]]
    sim_cell(cells$m[i], d[1], d[2], d[3], cells$ratio[i],
             nsim)
  }))
sim_res <- sim_res[order(sim_res$ratio, sim_res$m,
                         sim_res$design), ]

knitr::kable(
  sim_res,
  col.names = c('Design', '$\\sigma_b/\\sigma$', '$m$',
                'SD GLS', 'SD mis', 'SD avg',
                'Pow GLS', 'Pow mis', 'Pow avg'),
  caption = paste0(
    'Monte Carlo comparison of GLS (correctly specified), ',
    'GLS (misspecified $\\sigma_b^2$, assuming ',
    '$\\sigma_b^2=\\sigma^2$), and the averaged change-score ',
    'estimator under compound symmetry (', nsim,
    ' replicates, $\\Delta=0.5$, two-sided $\\alpha=0.05$). ',
    'Columns report the empirical standard deviation of ',
    '$\\hat\\gamma$ and the empirical power.'),
  booktabs = TRUE, escape = FALSE, digits = 3,
  row.names = FALSE)
```


# Numerical examples

## MIRIAD parameters

```{r miriad-comparison}
# MIRIAD within-visit residual = between-visit (0.3338) +
# between-scan measurement error (0.0025); see Paper 1.
sigma2_ad <- 0.3338 + 0.0025
sigma_b2_ad <- 0.9745

configs <- expand.grid(
  r = c(0, 1, 2, 4),
  k = c(1, 2, 3),
  f = c(0, 1)
)
configs <- configs[configs$r + configs$k +
                     configs$f >= 2, ]

configs$v_gls <- mapply(
  function(r, k, f)
    var_gamma_matrix(r, k, f, sigma2_ad,
                     sigma_b2_ad, 1.0),
  configs$r, configs$k, configs$f
)

configs$v_anc <- mapply(
  function(r, k, f)
    var_gamma_ancova(r, k, f, sigma2_ad,
                     sigma_b2_ad, 1.0),
  configs$r, configs$k, configs$f
)

configs$v_avg <- mapply(
  function(r, k, f)
    var_gamma_avg(r, k, f, sigma2_ad,
                  sigma_b2_ad, 1.0),
  configs$r, configs$k, configs$f
)

configs$label <- sprintf('(%d,%d,%d)',
  configs$r, configs$k, configs$f)

delta_ad <- 0.25
za <- qnorm(0.975)
zb <- qnorm(0.90)
configs$n_gls <- ceiling(
  (za + zb)^2 * configs$v_gls / delta_ad^2)
configs$n_anc <- ceiling(
  (za + zb)^2 * configs$v_anc / delta_ad^2)
configs$n_avg <- ceiling(
  (za + zb)^2 * configs$v_avg / delta_ad^2)

display <- configs[order(configs$v_gls),
  c('label', 'n_gls', 'n_anc', 'n_avg')]

knitr::kable(
  head(display, 15),
  col.names = c('$(J_0,J_1,J_2)$', '$N$ GLS', '$N$ ANC',
                '$N$ Avg'),
  caption = paste0(
    'Sample size per group: GLS, ANCOVA, and ',
    'averaged estimators (MIRIAD parameters, ',
    '$\\Delta=', delta_ad, '$, 90\\% power).'),
  booktabs = TRUE,
  escape = FALSE,
  row.names = FALSE
)
```

## When to use which estimator

<!-- TODO: decision flowchart or rule based on -->
<!-- n, sigma_b/sigma, and confidence in the -->
<!-- correlation structure -->


# Discussion

\begin{bullets}
\textbf{GLS is the efficient choice under a correctly specified, adequately estimated model, which in practice means moderate-to-large samples.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item GLS attains the minimum variance among linear unbiased estimators when the covariance is known.
\item Its advantage is realized only when the variance components are estimated accurately, which requires sufficient sample size and follow-up.
\item In small samples the instability of the estimated covariance can erase the theoretical gain.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
The estimator comparison yields a graded recommendation rather
than a single winner. Under correct specification GLS is
efficient and dominates both ANCOVA and the averaged
change-score estimator, as the ordering
$\Var_{\text{GLS}} \le \Var_{\text{ANC}} \le
\Var_{\text{avg}}$ confirms. This dominance is a large-sample
statement: it presumes the covariance components are known or
well estimated. With moderate-to-large samples and a credible
covariance model, GLS is the appropriate default.
:::

\begin{bullets}
\textbf{When the variance components are uncertain, the simpler estimators are the safer choice because they do not amplify misspecification.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item GLS weights observations by the inverse covariance, so errors in the assumed covariance propagate into the estimator.
\item The averaged change-score estimator uses no variance components and is therefore unaffected by their misspecification.
\item ANCOVA estimates a single covariate coefficient from the data, occupying a robust middle ground.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
The misspecification analysis qualifies the case for GLS. Because
GLS weights the data by an assumed inverse covariance, errors
in the assumed $G$ or $R$ feed directly into the estimator,
and the sandwich-variance calculations show that a mis-stated
covariance can raise the GLS variance above that of the simpler
estimators. The averaged change-score estimator is immune to
this failure because it never uses the variance components.
ANCOVA sits between the two: it estimates the coefficient on
the run-in mean from the data rather than deriving it from
assumed components, so it captures most of the run-in
information while remaining comparatively robust. When the
covariance is known only approximately, ANCOVA is the
recommended default and the averaged estimator a conservative
fallback.
:::

\begin{bullets}
\textbf{The efficiency gap between GLS and the simpler estimators narrows as the design grows, so the choice matters most in small, information-poor designs.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item With many run-in and treatment observations the phase means become near-sufficient for the random slope.
\item The averaged and ANCOVA estimators then recover most of the information GLS extracts, and the relative efficiency approaches one.
\item The estimator choice is therefore most consequential when observations are few and the slope-to-noise ratio is high.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
The practical importance of the estimator choice depends on the
design. As the numbers of run-in and treatment observations
grow, the phase means approach sufficiency for the
subject-specific slope, the ANCOVA efficiency loss vanishes as
$J_0 \to \infty$, and the relative efficiencies of all three
estimators converge toward one. The gap is largest, and the
choice of estimator therefore most consequential, in small
designs with few observations per phase and a high
slope-to-noise ratio, exactly the regime in which the
estimated covariance is also least stable. This coincidence
reinforces the preference for the simpler, more robust
estimators in small studies.
:::

\begin{bullets}
\textbf{Two extensions, the seemingly unrelated regression estimator and the constrained longitudinal data analysis estimator, would complete the efficiency picture.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The SUR formulation offers a middle ground between the averaged estimator and full GLS but adds implementation complexity.
\item The cLDA estimator exploits the full covariance without specifying the random-effects distribution and is expected to rank between ANCOVA and GLS.
\item A full comparison including both, and a feasible-GLS simulation under estimated components, is the natural continuation of this work.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
Two estimators not fully developed here would round out the
comparison. The seemingly unrelated regression formulation of
Section 2.4 treats the run-in and treatment periods as linked
equations and offers a compromise between the averaged
estimator and full GLS, at the cost of additional
implementation complexity; its variance derivation remains to
be completed. The constrained longitudinal data analysis
estimator uses the full covariance without committing to a
random-effects distribution and should fall between ANCOVA
and GLS in efficiency. A comprehensive treatment would compare
all of these under a feasible-GLS simulation in which the
variance components are estimated rather than assumed, which is
the setting in which the robustness distinctions drawn here
become decisive.
:::

# Conflict of interest

The author declares no conflicts of interest.

# Funding

This work received no specific funding.

# Data availability statement

\begin{bullets}
\textbf{The package implementing all three estimators, the simulation drivers, and the shared bibliography are openly available in the project repository.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The estimator implementations reside in the package source files.
\item The simulation drivers and shared compendium bibliography are included.
\item Specific paths appear in the Reproducibility section below.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
The `runinpower` R package implementing the GLS, ANCOVA, and
averaged-change-score estimators (`ancova.R`, `averaged.R`),
the simulation drivers, and the shared bibliography across the
four-paper compendium are openly available in the project
repository. Specific paths are listed in the Reproducibility
section below.
:::

# Reproducibility

\begin{bullets}
\textbf{The manuscript was rendered with R 4.4.x using the package source and a small set of standard analysis and build packages.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The in-package source supplies the estimator implementations.
\item The GLS comparator in the simulation relies on nlme, with tinytest as the testing harness.
\item The manuscript is built with rmarkdown and bookdown.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
This manuscript was rendered with R 4.4.x. Principal packages
used are the in-package `runinpower` source, `nlme` (the GLS
fitter for the simulation comparator), `tinytest` (testing
harness), and `rmarkdown` plus `bookdown` for the manuscript
build.
:::

\begin{bullets}
\textbf{The simulation uses a fixed master seed with deterministic per-replicate offsets to ensure reproducibility.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item A single master seed is set at the entry point of the Monte Carlo comparison.
\item Per-replicate seeds derive deterministically from the master seed by adding the replicate index.
\item This scheme makes the full simulation exactly reproducible.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
**Random-number generator.** The Monte Carlo simulation
comparing the three estimators sets `set.seed(20260418)` at
its entry point. Per-replicate seeds derive from the master
seed by `+ rep_idx`.
:::

\begin{bullets}
\textbf{All code, drivers, companion manuscripts, and the shared bibliography are located at documented paths within the compendium.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The package source and simulation drivers reside in their respective directories.
\item The three companion manuscripts are stored alongside this paper in the compendium.
\item The manuscript source and shared bibliography are co-located in the paper directory.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
**Data and code availability.** The R package source is at
`R/`. Simulation drivers are at `analysis/scripts/`. The
companion manuscripts (Papers 1, 2, 3) are at
`analysis/report/{01,02,03}-*/`. The manuscript source and
shared bibliography (`references.bib`, a symlink to
`../../references.bib`) are at
`analysis/report/04-gls-vs-ttest/`.
:::

# References


::: {#refs}
:::

```{r footer, echo=FALSE, results='asis'}
src <- sub("^/Users/zenn/Dropbox/prj/", "~/prj/",
           knitr::current_input(dir = TRUE))
cat(sprintf(
  "\n\\vfill\n\n---\n\n*Rendered on %s.*  \n*Source: `%s`*\n",
  format(Sys.time(), "%Y-%m-%d at %H:%M %Z"),
  src
))
```
