---
title: |
  GLS versus Averaged Change-Score Estimators for
  Clinical Trials with Run-In Observations
author: "Ronald G. Thomas, Ph.D.\\thanks{Wertheim School of Public Health, UCSD; ORCID: 0000-0003-1686-4965}"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
affiliation: "University of California, San Diego"
abstract: |
  **Background.** Three estimators of the treatment effect in
  longitudinal trials with run-in observations are in routine
  use: the GLS estimator under the linear mixed model, which
  exploits the full covariance structure to extract maximum
  information from all observations; the ANCOVA estimator,
  which regresses post-randomization means on the run-in mean
  as a covariate without specifying the random-effects
  covariance; and the averaged change-score estimator, which
  computes a simple difference of phase means and applies a
  $t$-test, requiring no variance-component estimation. The
  relative efficiency of these estimators under correct and
  misspecified models has not been comprehensively
  characterized.

  **Methods.** We derive the variance of the three
  estimators as a function of the trial design
  $(J_0, J_1, J_2)$ and the variance parameters
  $(\sigma^2, \sigma_b^2)$. The derivation proceeds via the
  Woodbury-based variance formulas of Paper 1 in this
  compendium, plus closed-form expressions for the ANCOVA
  and averaged-change-score variances under the same
  random-slopes model. Robustness to misspecification of the
  within-subject covariance ($R$) and the random-effects
  covariance ($G$) is assessed analytically through sandwich
  variances, and a Monte Carlo simulation under compound
  symmetry evaluates the estimators alongside a GLS analysis
  that misstates the random-slope variance.

  **Results.** All comparisons are made on the common
  estimand scale, with the averaged and ANCOVA contrasts
  rescaled to unbiasedness for the treatment effect
  $\gamma$. Under correct specification, GLS $\le$ ANCOVA
  $\le$ averaged in variance, with no violations over a
  115-cell grid of designs and signal-to-noise ratios. The
  efficiency penalty of the simpler estimators is
  substantial in run-in designs with multiple
  treatment-period visits: at
  $(J_0, J_1, J_2) = (2, 2, 0)$ the relative efficiency of
  ANCOVA versus GLS is 0.64 to 0.81 across signal-to-noise
  ratios, an efficiency loss of 19\% to 36\%. Relative
  efficiency above 0.95 occurs only in single-post-visit
  run-in designs ($J_1 = 1$, $J_2 = 0$) and in designs
  without run-in or common close ($J_0 = 0$, $J_2 = 0$) at
  high signal-to-noise. In the MIRIAD-based example the
  ANCOVA sample size exceeds the GLS sample size by up to
  a factor of roughly 2.6 among run-in designs, and the
  averaged sample size by up to a factor of roughly 45.
  Under the $G$ and $R$ misspecifications studied, GLS
  never loses its advantage over the simpler estimators:
  the misspecified-GLS sandwich variance peaks at 5.3
  against an averaged-estimator variance of 8.9 at the
  reference design.

  **Conclusions.** GLS is the efficient estimator under
  correct specification, and its advantage survives the
  covariance misspecifications examined here. The averaged
  change-score estimator remains useful as a fully
  prespecified robust benchmark and a teaching tool. The
  open robustness question, not studied here, is whether
  feasible GLS with variance components estimated from
  small samples retains this advantage.
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    number_sections: true
    keep_tex: true
    highlight: tango
    includes:
      in_header: claudecode.tex
fontsize: 11pt
geometry: "left=3cm,right=5cm,top=2cm,bottom=2cm"
bibliography: references.bib
knit: (function(input, ...) { d <- dirname(input); while (!file.exists(file.path(d, 'tools', 'stamp-render.R')) && d != dirname(d)) d <- dirname(d); source(file.path(d, 'tools', 'stamp-render.R'))$value(input) })
csl: statistics-in-medicine.csl
link-citations: true
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{lineno}
  - \linenumbers
  - \usepackage{booktabs}
  - \usepackage{float}
  - \DeclareMathOperator{\Var}{Var}
  - \DeclareMathOperator{\Cov}{Cov}
  - \DeclareMathOperator{\E}{E}
  - \usepackage{setspace}
  - \setstretch{1.1}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4,
  fig.pos = "H"
)
library(runinpower)
```

# Introduction

\begin{bullets}
\textbf{The optimality of GLS is contingent on knowing the covariance, which in practice must be estimated and may be unstable in small samples.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item GLS is the best linear unbiased estimator (BLUE) only when the covariance matrix is known.
\item Feasible GLS substitutes estimated variance components, so its efficiency depends on estimation accuracy.
\item Small samples or short follow-up yield unstable estimates that can erode the theoretical advantage of GLS.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
The GLS estimator $\hat\gamma_{\text{GLS}} =
(X'\Sigma^{-1}X)^{-1}X'\Sigma^{-1}\mathbf{c}$ is the
best linear unbiased estimator (minimum variance among
linear unbiased estimators) when $\Sigma$ is
known. In practice, $\Sigma$ must be estimated, and
the efficiency of the feasible GLS depends on the
accuracy of this estimation. With small sample sizes
or short follow-up, variance component estimates can
be unstable, potentially eroding the theoretical
advantage of GLS.
:::

The averaged change-score approach sidesteps variance
component estimation entirely. For a trial with $J_0$
run-in and $J_2$ common close observations, the
estimator is:
\begin{equation}
  \hat\gamma_{\text{avg}} = \bar{d}_1 - \bar{d}_0
\end{equation}
where
\begin{equation}
  d_{ik} = \frac{1}{J_1 + J_2}
    \sum_{\ell=1}^{J_1+J_2} Y_{i,\ell,k}
    - \frac{1}{J_0}
    \sum_{\ell=1}^{J_0} Y_{i,-\ell,k}
\end{equation}
is the difference of the post-randomization and run-in
means for participant $i$. The baseline visit is the
change-score reference and cancels, so it enters neither
average; this matches the contrast implemented in
`var_gamma_avg()`. Throughout, the code and figure
parameters use $(r, k, f)$ for the design counts, with
the correspondence $J_0 = r$ (run-in), $J_1 = k$
(treatment period), and $J_2 = f$ (common close); the
prose and table headers use $(J_0, J_1, J_2)$.

\begin{bullets}
\textbf{The relative merits of change-score and ANCOVA analyses have been studied extensively across several methodological traditions.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Formal efficiency comparisons for pretest-posttest designs trace back to @brogan1980 and @yang2001.
\item @frison1992 and @frison1997 established the efficiency ordering among post-treatment, change-score, and ANCOVA analyses with multiple baseline measurements, the closest antecedent of the present comparison.
\item Subsequent work contrasted baseline-as-covariate with baseline-as-response and developed semiparametric covariate adjustment.
\item Later contributions clarified when longitudinal models gain power over ANCOVA and adapted these ideas to run-in trials.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
The debate between change-score and ANCOVA analyses has
a long history [@senn2006; @vanbreukelen2006;
@crager1987]. @yang2001 provided a formal efficiency
comparison for pretest-posttest designs. The closest
antecedent of the present work is @frison1992, who
derived the relative efficiency of post-treatment,
change-score, and ANCOVA analyses when several baseline
measurements are available, and @frison1997, who
extended the treatment to repeated post-treatment
measurements. Their ordering, that ANCOVA on the mean of
the available baselines dominates the change score which
in turn dominates the raw post-treatment comparison, is
the qualitative result that the ordering established in
Section 3.1 specializes. What is new here is not the
ordering itself but the closed-form relative
efficiencies in the run-in random-slope parameterization:
the efficiency margins are obtained as explicit
functions of the design counts $(J_0, J_1, J_2)$ and the
signal-to-noise ratio $\sigma_b/\sigma$, they cover a
post-treatment common close phase absent from the
Frison-Pocock setting, and the estimand is a difference
of slopes rather than a difference of means. Within the
run-in literature, @hu2021 derived GLS power formulas
for trials with pre-randomization observations, which
the present comparison takes as the efficient benchmark
against which the simpler estimators are measured.
@liu2009 and
@lu2010 compared baseline-as-covariate against
baseline-as-response (constrained longitudinal data
analysis). @tsiatis2008 developed a semiparametric
framework for covariate adjustment. @mallinckrodt2021
showed that MMRM requires time-by-covariate interactions
to guarantee power gains over ANCOVA. @brogan1980
established the foundational comparison of pretest-posttest
designs. @coffman2016 addressed conditioning on baseline
in longitudinal RCTs. @wang2019 proposed the two-period
approach for run-in trials.
:::

\begin{bullets}
\textbf{The averaged change-score estimator is distinct from and less efficient than the constrained longitudinal data analysis estimator, which is excluded from the present comparison.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The cLDA estimator places baseline in the response vector under a common-mean constraint.
\item It exploits the full covariance structure without specifying the random-effects distribution, ranking between the averaged estimator and GLS.
\item A full comparison including cLDA lies beyond the present scope but constitutes a natural extension.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
We note that the averaged change-score estimator
studied here is distinct from (and less efficient than)
the constrained longitudinal data analysis (cLDA)
estimator of @liang2000 and @lu2010, which includes
baseline in the response vector with a common-mean
constraint. The cLDA estimator uses the full covariance
structure without requiring specification of the random
effects distribution $G$, placing it between the
averaged estimator and GLS in the efficiency ordering.
@funatogawa2020 extended the cLDA framework to pre-
and post-treatment measurements with equal baseline
assumptions. A full comparison including cLDA is beyond
the scope of this paper but represents a natural
extension.
:::

\begin{bullets}
\textbf{The longpower package implements the GLS baseline but omits the simpler estimators and the misspecification analysis developed here.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item It is the CRAN reference for slope-comparison power calculations under the Liu-Liang, Diggle, and Edland formulations.
\item Its routines correspond to the GLS estimator under correct specification, matching the Paper 1 baseline.
\item It does not cover ANCOVA, the averaged change-score test, or the efficiency loss from misspecified covariance.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
The `longpower` R package [@iddi2022] provides the
CRAN reference implementation of power calculations for
the GLS / REML-based slope-comparison estimator,
covering the Liu-Liang [@liuliang1997], Diggle
[@diggle2002], and Edland [@ard2011] formulations.
These routines correspond to the $\hat\gamma_{\text{GLS}}$
estimator studied here under correct specification and
constitute the Paper 1 baseline extended by run-in and
common close observations. However, `longpower` does not
cover the ANCOVA-on-run-in-mean estimator or the
averaged change-score $t$-test, nor does it characterize
the efficiency loss from misspecification of $G$ or $R$.
The comparison developed here therefore complements
`longpower` by (i) ordering the three candidate
estimators under correct specification and (ii)
quantifying the efficiency margin that GLS retains
under the misspecifications of $G$ and $R$ studied
below.
:::


# Three estimators

## GLS under the mixed model

\begin{bullets}
\textbf{The GLS variance for the treatment effect follows from the relevant diagonal element of the inverse information matrix under the marginal covariance.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The per-pair variance is the treatment-effect element of $(X'\Sigma^{-1}X)^{-1}$.
\item The marginal covariance decomposes as residual plus random-effects components.
\item The inverse is obtained efficiently via the Woodbury identity.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
From @frost2008, the per-pair variance is the
$(\gamma, \gamma)$ element of
$(X'\Sigma^{-1}X)^{-1}$ with $\Sigma = R + ZGZ'$
[@laird1982; @verbeke2000], computed via the Woodbury
identity.
:::

For the Frost conventional design ($J_0 = 0$, $J_1 = 2$):
\begin{equation}
  \hat\gamma_{\text{GLS}} = \frac{1}{2t}
    [(C_{22} + C_{21}) - (C_{12} + C_{11})]
\end{equation}
with $\Var(\hat\gamma_{\text{GLS}}) =
(\sigma^2 + 2\sigma_b^2 t^2)/t^2$.

## Averaged change score ($t$-test)

The averaged change-score estimator ignores the
within-subject correlation structure and treats
$d_{ik}$ as independent observations:
\begin{equation}
  \Var(\hat\gamma_{\text{avg}})
  = \frac{2}{m}\Var(d_{ik})
\end{equation}

The variance $\Var(d_{ik})$ depends on the correlation
structure but does not require its estimation. As an
estimator of $\gamma$, the contrast is divided by the
design constant $c$ derived in Section 3.1, and all
variances, relative efficiencies, and sample sizes
reported in this paper are on that common $\gamma$
scale.

```{r avg_variance, echo=TRUE}
a <- 1.0
b <- 1.0
tt <- 1.0

# gamma-scale rescaling constant (Section 3.1):
# mean accumulated effect over post-randomization visits
cfac <- function(r, k, f, t_interval = 1) {
  t_interval * (k * (k + 1) / 2 + k * f) / (k + f)
}

v_gls <- var_gamma_matrix(2, 2, 0, a, b, tt)
v_avg <- var_gamma_avg(2, 2, 0, a, b, tt) /
  cfac(2, 2, 0, tt)^2
re <- v_gls / v_avg

cat(sprintf('J_0=2, J_1=2: GLS=%.4f, Avg=%.4f, RE=%.4f\n',
            v_gls, v_avg, re))
```


## ANCOVA with run-in mean as covariate

The ANCOVA approach uses the averaged run-in rate as a
baseline covariate without specifying $\sigma_b^2$:
\begin{equation}
  \bar{Y}_{i,\text{post}} = \alpha + \gamma g_i
    + \phi \bar{Y}_{i,\text{pre}} + \varepsilon_i
\end{equation}

The variance of $\hat\gamma$ under this model is
\begin{equation}
  \Var(\hat\gamma_{\text{ANC}})
  = \frac{2}{m}\,\Var(\bar{Y}_{\text{post}})
    \,(1 - \rho_{\text{pre,post}}^2)
\end{equation}
\begin{bullets}
\textbf{The ANCOVA variance is governed by the run-in to post-randomization correlation, which determines the magnitude of the variance reduction.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The factor $(1 - \rho^2)$ scales the variance by the squared marginal correlation between the phase means.
\item With no run-in observations the covariate vanishes and ANCOVA reduces to the averaged estimator.
\item A large correlation, arising when the random-slope variance dominates the residual, yields substantial variance reduction.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
where $\rho_{\text{pre,post}}$ is the marginal
correlation between the run-in and post-randomization
means. When $J_0 = 0$ there is no covariate and ANCOVA
reduces to the averaged estimator. When
$\rho_{\text{pre,post}}$ is large (i.e., when
$\sigma_b^2$ dominates $\sigma^2$), the $(1-\rho^2)$
factor produces substantial variance reduction.
:::

\begin{bullets}
\textbf{ANCOVA occupies a middle ground by estimating the baseline coefficient from the data while still using the predictive run-in observations.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Unlike GLS, the coefficient on the run-in mean is estimated empirically rather than derived from assumed variance components.
\item Unlike the averaged estimator, it exploits the predictive value of the run-in observations.
\item It therefore reduces variance without committing to a full random-effects specification.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
The key advantage over GLS is that $\phi$ is estimated
from the data rather than derived from assumed variance
components [@senn2006; @borm2007]. The key advantage
over the averaged estimator is that it exploits the
predictive value of the run-in observations
[@vanbreukelen2006; @vickers2001].
:::

```{r ancova_demo, echo=TRUE}
c220 <- cfac(2, 2, 0, tt)
v_gls <- var_gamma_matrix(2, 2, 0, a, b, tt)
v_anc <- var_gamma_ancova(2, 2, 0, a, b, tt) / c220^2
v_avg <- var_gamma_avg(2, 2, 0, a, b, tt) / c220^2

cat(sprintf('J_0=2, J_1=2 (sigma_b/sigma = 1):\n'))
cat(sprintf('  GLS:    %.4f\n', v_gls))
cat(sprintf('  ANCOVA: %.4f\n', v_anc))
cat(sprintf('  Avg:    %.4f\n', v_avg))
cat(sprintf('  RE(GLS/ANC): %.3f\n', v_gls / v_anc))
cat(sprintf('  RE(ANC/Avg): %.3f\n', v_anc / v_avg))
```


## SUR (seemingly unrelated regression)

The SUR approach treats the run-in and treatment
periods as separate regression equations linked by
correlated errors:
\begin{align}
  Y_1 &= X_2 \begin{pmatrix}
    \beta_{11} \\ \beta_{12}
  \end{pmatrix} + e_1
  & &\text{(both groups)} \\
  Y_2 &= X_1 \alpha_2 + e_2
  & &\text{(baseline)}
\end{align}

The deviation-from-mean parameterization gives
$\beta_1 = (\delta_1 + \delta_2)/2$ and
$\beta_2 = (\delta_2 - \delta_1)/2$, where
$\delta_j$ are the group-specific rates.

The SUR variance derivation and its comparison to the
GLS and averaged estimators are not developed in this
paper; the subsection is retained as a specification
only and is flagged as future work in the Discussion.


## BLUP as theoretical benchmark

\begin{bullets}
\textbf{The three estimators are linear functions of the change scores and admit a common benchmark in the best linear unbiased predictor of the random slopes.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Each estimator can be viewed as a linear combination of the observed change scores.
\item The BLUP of the subject-specific random slope provides the natural reference predictor.
\item It weights the residuals by the inverse marginal covariance using the GLS fixed-effect estimate.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
The three estimators above are linear functions of the
observed change scores and can be compared to a common
benchmark: the best linear unbiased predictor (BLUP) of the
individual random slopes. Under the random-slopes model with
marginal covariance $V_i = R_i + Z_i G Z_i'$, the BLUP of the
subject-specific random slope $b_i$ given the data is
\begin{equation}
  \hat b_i
  = G Z_i' V_i^{-1}
    \bigl(y_i - X_i \hat{\boldsymbol{\beta}}\bigr)
  \label{eq:blup}
\end{equation}
where $\hat{\boldsymbol{\beta}}$ is the GLS estimator of the
fixed effects [@henderson1975; @robinson1991].
:::

\begin{bullets}
\textbf{GLS attains the minimum variance among linear unbiased estimators when the covariance is correct, while the simpler estimators are specific approximations to this benchmark.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The group-specific slope estimators inherit the variance of the treatment-effect fixed effect.
\item GLS is optimal among linear unbiased estimators under a correctly specified marginal covariance.
\item The averaged and ANCOVA estimators are particular linear approximations to this optimal benchmark.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
The group-specific slope estimators derived from the fixed
effects share the same variance as
$(\hat{\boldsymbol{\beta}})_\gamma$, and the GLS estimator
$\hat\gamma_{\text{GLS}}$ minimizes the variance among all
linear unbiased estimators of $\gamma$ when $V_i$ is correctly
specified. The averaged and ANCOVA estimators are specific
linear approximations to this benchmark:
:::

- **Averaged change-score** replaces $V_i^{-1}$ with an
  unweighted sum; this is optimal only when $V_i \propto I$,
  which fails for any non-zero random-slope variance.
- **ANCOVA** collapses the run-in visits into a single
  scalar covariate (the run-in mean) rather than weighting
  them by $V_i^{-1}$; the resulting efficiency loss vanishes
  as $J_0 \to \infty$ since the run-in mean becomes a
  sufficient statistic for the subject-specific intercept
  plus average slope.

\begin{bullets}
\textbf{Misspecification erodes but, in the settings studied here, does not eliminate the GLS efficiency advantage over the simpler approximations.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item GLS propagates errors in the assumed covariance through its inverse weighting.
\item The averaged and ANCOVA estimators depend on fewer assumed variance components.
\item Section 4 quantifies the resulting erosion; in the misspecifications examined, GLS retains a large margin.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
Section 4 examines whether model misspecification erodes
the nominal optimality of GLS relative to these
approximations, which depend on fewer variance
components. In the $G$ and $R$ misspecifications studied
there, the GLS variance increases but remains below
that of the averaged estimator; whether feasible GLS with
estimated components loses its advantage in small samples
is not studied here.
:::

# Relative efficiency under correct specification

## Analytical comparison

The relative efficiency of the averaged estimator
to GLS is:
\begin{equation}
  \text{RE}_{\text{avg}} =
    \frac{\Var(\hat\gamma_{\text{GLS}})}
         {\Var(\hat\gamma_{\text{avg}}/c)}
    \le 1
\end{equation}
with equality when the averaged estimator happens
to coincide with GLS (which occurs in special cases).

The ordering asserted above can be stated and proved
directly. One preliminary is needed. As defined, the
averaged and ANCOVA contrasts average the accumulated
treatment effect over the post-randomization visits, so
their expectation is $c\gamma$ rather than $\gamma$,
where
\begin{equation}
  c = t\,\frac{J_1(J_1 + 1)/2 + J_1 J_2}{J_1 + J_2}
\end{equation}
is the mean of the accumulated-effect design column over
the post-randomization visits ($c = 1.5t$ for the
$(2,2,0)$ design; $c = t$ when $J_1 = 1$, $J_2 = 0$).
Dividing either contrast by $c$ yields an unbiased
estimator of $\gamma$ with variance scaled by $1/c^2$.

**Proposition.** Under the random-slopes model with
correctly specified $\Sigma$,
$\Var(\hat\gamma_{\text{GLS}}) \le
\Var(\hat\gamma_{\text{ANC}}/c) \le
\Var(\hat\gamma_{\text{avg}}/c)$.

*Proof.* Each rescaled contrast is a linear unbiased
estimator of $\gamma$, and $\hat\gamma_{\text{GLS}}$ is
the best linear unbiased estimator of $\gamma$ when
$\Sigma$ is known (Gauss-Markov), which gives the first
inequality. For the second, both contrasts have the form
(post-randomization mean) $- \phi \times$ (run-in mean):
the population ANCOVA coefficient minimizes the residual
variance over $\phi$, while the averaged estimator fixes
$\phi = 1$, so the ANCOVA contrast variance is no larger;
dividing both by $c^2$ preserves the inequality. (The
feasible ANCOVA estimates $\phi$ and carries the usual
finite-sample inflation.) $\square$

The ordering itself is not new. @frison1992 established
the same qualitative ranking for repeated-measures
designs with several baseline measurements, and
@frison1997 extended it to repeated post-treatment
measurements; the argument above is the specialization
of that result to the run-in random-slope model, where
the run-in mean plays the role of their baseline mean.
The contribution of the following sections is the
closed-form magnitude of the two gaps as functions of
$(J_0, J_1, J_2)$ and $\sigma_b/\sigma$, which the
Frison-Pocock parameterization does not deliver for a
slope estimand or for designs with a common close phase.

Numerically, the ordering holds with zero violations
over the 115-cell grid of the tables below. All tables,
figures, and simulation results in this paper are on the
equal-estimand $\gamma$ scale: the averaged and ANCOVA
contrast variances are divided by $c^2$ before any
comparison. For reference, the raw phase-mean contrasts
carry variances larger by the factor $c^2$ ($c^2 = 2.25$
at the $(2,2,0)$ design), so raw-contrast comparisons
would overstate the penalty of the simpler estimators.

```{r re-table}
configs <- expand.grid(
  r = c(0, 1, 2, 4),
  k = c(1, 2, 3),
  f = c(0, 1)
)
configs <- configs[configs$r + configs$k +
                     configs$f >= 2, ]

ratio_levels <- c(0.5, 1.0, 2.0, 5.0, 20.0)
a <- 1.0

for (ratio in ratio_levels) {
  b <- ratio^2 * a
  # gamma-scale RE: multiply the raw contrast RE by c^2
  configs[[paste0('avg_', ratio)]] <- mapply(
    function(r, k, f)
      re_avg_vs_gls(r, k, f, a, b, tt) *
        cfac(r, k, f, tt)^2,
    configs$r, configs$k, configs$f
  )
  configs[[paste0('anc_', ratio)]] <- mapply(
    function(r, k, f)
      re_ancova_vs_gls(r, k, f, a, b, tt) *
        cfac(r, k, f, tt)^2,
    configs$r, configs$k, configs$f
  )
}

configs$label <- sprintf('(%d,%d,%d)',
  configs$r, configs$k, configs$f)

avg_cols <- paste0('avg_', ratio_levels)
anc_cols <- paste0('anc_', ratio_levels)

display_avg <- configs[, c('label', avg_cols)]
names(display_avg) <- c('$(J_0,J_1,J_2)$',
  paste0('$', ratio_levels, '$'))

knitr::kable(
  display_avg,
  caption = paste0(
    'RE of averaged change-score vs GLS on the ',
    '$\\gamma$ scale ($\\Var_{\\text{GLS}} / ',
    '\\Var(\\hat\\gamma_{\\text{avg}}/c)$) by ',
    '$\\sigma_b/\\sigma$.'),
  booktabs = TRUE,
  escape = FALSE,
  digits = 3,
  row.names = FALSE
)

```

```{r re-table-anc}
display_anc <- configs[, c('label', anc_cols)]
names(display_anc) <- c('$(J_0,J_1,J_2)$',
  paste0('$', ratio_levels, '$'))

knitr::kable(
  display_anc,
  caption = paste0(
    'RE of ANCOVA vs GLS on the $\\gamma$ scale ',
    '($\\Var_{\\text{GLS}} / ',
    '\\Var(\\hat\\gamma_{\\text{ANC}}/c)$) by ',
    '$\\sigma_b/\\sigma$.'),
  booktabs = TRUE,
  escape = FALSE,
  digits = 3,
  row.names = FALSE
)
```

## When does the gap exceed 5%?

```{r gap-5pct}
gap <- do.call(rbind, lapply(ratio_levels, function(rr) {
  data.frame(
    ratio = rr,
    r = configs$r, k = configs$k, f = configs$f,
    re_anc = configs[[paste0('anc_', rr)]],
    re_avg = configs[[paste0('avg_', rr)]]
  )
}))
n_cells <- nrow(gap)
hi_anc <- gap[gap$re_anc >= 0.95, ]
n_hi_anc <- nrow(hi_anc)
n_k1 <- sum(hi_anc$k == 1 & hi_anc$f == 0)
n_r0 <- sum(hi_anc$r == 0 & hi_anc$f == 0 &
              hi_anc$k >= 2)
stopifnot(n_hi_anc == n_k1 + n_r0)
n_hi_avg <- sum(gap$re_avg >= 0.95)
re_220 <- gap$re_anc[gap$r == 2 & gap$k == 2 & gap$f == 0]
```

The answer, computed over the full grid of the preceding
tables (all on the $\gamma$ scale), is: in most cells.
Of the `r n_cells` design-by-ratio cells, `r n_hi_anc`
have ANCOVA relative efficiency of at least 0.95 (an
efficiency loss below 5%), and they form exactly two
families. The first (`r n_k1` cells) comprises
single-post-visit run-in designs with no common close
($J_1 = 1$, $J_2 = 0$): the $(1,1,0)$ design at every
tabulated ratio, plus $(2,1,0)$ at
$\sigma_b/\sigma = 0.5$; with a single post-randomization
visit there is no within-period slope information for GLS
to exploit. The second (`r n_r0` cells) comprises
designs without run-in or common close ($J_0 = 0$,
$J_2 = 0$, $J_1 \in \{2, 3\}$) at
$\sigma_b/\sigma \ge 2$, where ANCOVA reduces to the
averaged estimator and the rescaled phase mean is nearly
efficient. In run-in designs with $J_1 \ge 2$, the
regime that motivates this paper, the loss is
substantial: at the representative $(2,2,0)$ design the
relative efficiency ranges from
`r sprintf('%.2f', max(re_220))` at
$\sigma_b/\sigma = 0.5$ down to
`r sprintf('%.2f', min(re_220))` at
$\sigma_b/\sigma = 20$, an efficiency loss of 19% to
36%. The averaged estimator reaches 0.95 only in the
second family (`r n_hi_avg` of `r n_cells` cells).


# Robustness to misspecification

## Misspecified $G$

\begin{bullets}
\textbf{Misestimating the random-slope variance forfeits the optimality of GLS while leaving the averaged estimator unchanged.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item An incorrect random-slope variance renders the assumed-covariance GLS no longer minimum-variance.
\item The error enters through the inverse-covariance weighting of GLS.
\item The averaged change-score estimator's definition does not use this variance component, so its variance does not depend on the assumed value.
\item In the sweep below, however, the misspecified GLS variance peaks at 5.30 while the averaged variance on the $\gamma$ scale is 8.89: the GLS margin is eroded, not eliminated.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
When $\sigma_b^2$ is misestimated (e.g., too large
or too small), the GLS analysis that assumes it is no
longer optimal. Its sampling variance is then the
sandwich form
$(X'\Sigma_a^{-1}X)^{-1}
 X'\Sigma_a^{-1}\Sigma\Sigma_a^{-1}X
 (X'\Sigma_a^{-1}X)^{-1}$,
with $\Sigma_a$ the assumed and $\Sigma$ the true
covariance, which is the model-robust variance of
@liang1986 evaluated at a known working covariance
rather than an estimated one. The averaged change-score
estimator's definition does not depend on the assumed
value, so its variance is constant across the sweep. Note the scale of
the comparison: over assumed $\sigma_b^2 \in [0.01, 5]$
with true $\sigma_b^2 = 1$, the misspecified GLS
sandwich variance never exceeds 5.30, still below the
averaged-estimator variance of 8.89 on the $\gamma$
scale. The studied misspecification erodes but does not
overturn the GLS advantage.
:::

```{r misspec_G, fig.cap="Effect of misspecifying $\\sigma_b^2$ on GLS variance. True $\\sigma_b^2 = 1$. The GLS estimator uses an assumed value on the x-axis. The averaged estimator's definition does not use $\\sigma_b^2$, so its $\\gamma$-scale variance (horizontal line) does not depend on the assumed value. The misspecified GLS variance remains below the averaged variance throughout."}
b_true <- 1.0
a <- 1.0
r <- 2
k <- 2
f <- 0

b_assumed <- seq(0.01, 5, by = 0.05)

var_misspec_gls <- function(r, k, f, sigma2,
                            sigma_b2_true,
                            sigma_b2_assumed,
                            t_interval) {
  p <- r + k + f
  R <- build_R_matrix(p, sigma2)
  des <- build_design(r, k, f, t_interval)
  Z <- des$Z
  nc <- des$ncol_X

  Sigma_true <- R + Z %*% matrix(sigma_b2_true) %*% t(Z)

  Ri <- solve(R)
  G_a <- matrix(sigma_b2_assumed)
  H_a <- solve(solve(G_a) + t(Z) %*% Ri %*% Z)
  Xt <- des$X_trt
  Xp <- des$X_plc
  XRX <- t(Xt) %*% Ri %*% Xt + t(Xp) %*% Ri %*% Xp
  XtRZ <- t(Xt) %*% Ri %*% Z
  XpRZ <- t(Xp) %*% Ri %*% Z
  W_a <- XtRZ %*% H_a %*% t(XtRZ) +
         XpRZ %*% H_a %*% t(XpRZ)
  XSiX_a <- XRX - W_a
  XSiX_a_inv <- solve(XSiX_a)

  Sigma_a_inv <- Ri - Ri %*% Z %*% H_a %*% t(Z) %*% Ri

  bread <- XSiX_a_inv
  meat_trt <- t(Xt) %*% Sigma_a_inv %*%
    Sigma_true %*% Sigma_a_inv %*% Xt
  meat_plc <- t(Xp) %*% Sigma_a_inv %*%
    Sigma_true %*% Sigma_a_inv %*% Xp
  meat <- meat_trt + meat_plc

  sandwich <- bread %*% meat %*% bread
  sandwich[nc, nc]
}

v_misspec <- sapply(b_assumed, function(ba) {
  var_misspec_gls(r, k, f, a, b_true, ba, tt)
})

v_avg <- var_gamma_avg(r, k, f, a, b_true, tt) /
  cfac(r, k, f, tt)^2
v_correct <- var_gamma_matrix(r, k, f, a, b_true, tt)

plot(b_assumed, v_misspec, type = 'l', lwd = 2,
     xlab = expression(paste('Assumed ',
                             sigma[b]^2)),
     ylab = expression(Var(hat(gamma))),
     main = 'Misspecified random slope variance',
     ylim = c(min(v_correct, v_avg) * 0.9,
              max(v_misspec, v_avg) * 1.1))
abline(h = v_avg, col = 'red', lwd = 2, lty = 2)
abline(v = b_true, col = 'gray', lty = 3)
points(b_true, v_correct, pch = 16, cex = 1.5)
legend('topright',
       legend = c('Misspecified GLS (sandwich)',
                  'Averaged change-score',
                  'Correct GLS'),
       col = c('black', 'red', 'black'),
       lty = c(1, 2, NA),
       pch = c(NA, NA, 16),
       lwd = c(2, 2, NA),
       bty = 'n')
```

## Misspecified $R$

\begin{bullets}
\textbf{Misspecifying the residual correlation biases the GLS variance machinery but again leaves the averaged estimator intact.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Assuming compound symmetry when the truth is AR(1) is a representative misspecification.
\item Both the GLS variance formula and the Woodbury correction become biased.
\item The averaged estimator's definition does not depend on the assumed correlation, although its $\gamma$-scale variance still changes with the true $R$ (8.4 to 9.1 over the plotted range).
\item The misspecified GLS sandwich variance stays four- to elevenfold below the averaged variance across the sweep.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
When the residual correlation structure is wrong
(e.g., compound symmetry assumed but AR(1) is true),
both the GLS variance formula and the Woodbury
correction are biased. The averaged estimator's
definition does not depend on the assumed model, though
its $\gamma$-scale variance does change with the true
$R$ (8.4 to 9.1 across $\rho \in [0, 0.9]$ in the
figure). The GLS sandwich variance ranges from 0.74 to
2.16 over the same sweep, so the studied $R$
misspecification leaves GLS four to eleven times more
efficient than the averaged estimator throughout.
:::

```{r misspec_R, fig.cap="Effect of analyzing under compound symmetry when the true correlation is AR(1). Sandwich variance of GLS (solid) vs averaged change-score on the $\\gamma$ scale (dashed) as a function of $\\rho$. The averaged estimator's definition does not use the assumed correlation, but its variance varies with the true $R$; GLS remains at least fourfold more efficient at every $\\rho$."}
rho_grid <- seq(0, 0.9, by = 0.05)
r <- 2
k <- 2
f <- 0
a <- 1.0
b <- 1.0

var_misspec_r <- function(rho) {
  p <- r + k + f
  des <- build_design(r, k, f, tt)
  Z <- des$Z
  nc <- des$ncol_X

  d_runin <- if (r > 0) seq(-r, -1) else numeric(0)
  d_trt <- if (k > 0) seq(1, k) else numeric(0)
  d_all <- c(d_runin, d_trt)

  R_true <- build_R_matrix_ar1(p, a, rho, d_all)
  Sigma_true <- R_true + Z %*% matrix(b) %*% t(Z)

  R_assumed <- build_R_matrix(p, a)
  Ri_a <- solve(R_assumed)
  G <- matrix(b)
  H_a <- solve(solve(G) + t(Z) %*% Ri_a %*% Z)

  Xt <- des$X_trt
  Xp <- des$X_plc
  XRX <- t(Xt) %*% Ri_a %*% Xt +
         t(Xp) %*% Ri_a %*% Xp
  XtRZ <- t(Xt) %*% Ri_a %*% Z
  XpRZ <- t(Xp) %*% Ri_a %*% Z
  W_a <- XtRZ %*% H_a %*% t(XtRZ) +
         XpRZ %*% H_a %*% t(XpRZ)
  XSiX_a <- XRX - W_a
  XSiX_a_inv <- solve(XSiX_a)

  Sigma_a_inv <- Ri_a - Ri_a %*% Z %*% H_a %*%
                 t(Z) %*% Ri_a

  meat_trt <- t(Xt) %*% Sigma_a_inv %*%
    Sigma_true %*% Sigma_a_inv %*% Xt
  meat_plc <- t(Xp) %*% Sigma_a_inv %*%
    Sigma_true %*% Sigma_a_inv %*% Xp
  meat <- meat_trt + meat_plc

  sandwich <- XSiX_a_inv %*% meat %*% XSiX_a_inv
  sandwich[nc, nc]
}

v_sandwich <- sapply(rho_grid, var_misspec_r)

des <- build_design(r, k, f, tt)
n_post <- k + f
w <- c(rep(-1 / r, r), rep(1 / n_post, k + f))
v_avg_ar1 <- sapply(rho_grid, function(rho) {
  p <- r + k + f
  d_all <- c(seq(-r, -1), seq(1, k))
  if (rho == 0) {
    R_true <- build_R_matrix(p, a)
  } else {
    R_true <- build_R_matrix_ar1(p, a, rho, d_all)
  }
  Z <- des$Z
  Sigma_true <- R_true + Z %*% matrix(b) %*% t(Z)
  2 * as.numeric(t(w) %*% Sigma_true %*% w) /
    cfac(r, k, f, tt)^2
})

plot(rho_grid, v_sandwich, type = 'l', lwd = 2,
     xlab = expression(rho[true]),
     ylab = expression(Var(hat(gamma))),
     main = 'CS assumed, AR(1) true',
     ylim = range(c(v_sandwich, v_avg_ar1)))
lines(rho_grid, v_avg_ar1, col = 'red',
      lwd = 2, lty = 2)
legend('topleft',
       legend = c('GLS (CS assumed, sandwich)',
                  'Averaged change-score'),
       col = c('black', 'red'),
       lty = c(1, 2), lwd = 2, bty = 'n')
```


# Simulation study

## Design

\begin{bullets}
\textbf{A factorial Monte Carlo design complements the analytical results by varying sample size, design, signal-to-noise, and analysis method under a compound-symmetric truth.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Sample size and the run-in design configuration are crossed.
\item The signal-to-noise ratio is varied; the true correlation is compound symmetric throughout.
\item Three analysis methods are compared: correctly specified GLS, misspecified GLS, and the averaged change-score test.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
To complement the analytical results, we simulate
trials under the following factorial design:
:::

- $m \in \{50, 100\}$ per group
- $(J_0, J_1, J_2) \in \{(0,2,0), (2,2,0), (2,2,2)\}$
- $\sigma_b/\sigma \in \{1, 5\}$
- True correlation: compound symmetry (an AR(1) truth
  is examined analytically in Section 4 but is not
  simulated)
- Analysis: GLS (correctly specified), GLS
  (misspecified), averaged change-score

\begin{bullets}
\textbf{Each configuration is replicated 1000 times and evaluated on empirical variance and power; coverage is not assessed.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item One thousand trials are simulated per configuration.
\item Empirical standard deviation and empirical power are recorded for each estimator.
\item Coverage of nominal 95\% confidence intervals, empirical size at $\Delta = 0$, larger sample sizes, higher ratios, an AR(1) truth arm, and 5000 replicates remain future work.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
For each configuration, 1000 trials are simulated. The
outcomes are the empirical standard deviation of
$\hat\gamma$ and the empirical power. Coverage of
nominal 95\% confidence intervals and empirical size at
$\Delta = 0$ are not assessed; extending the factorial
to $m \in \{30, 200\}$, $\sigma_b/\sigma = 20$, an
AR(1) truth arm, and 5000 replicates remains future
work. Test statistics use standard normal critical
values rather than $t$; the distinction is immaterial
at $m \ge 50$ per group.
:::

```{r simulation}
simulate_trial <- function(m, r, k, f, sigma2,
                           sigma_b2, t_interval,
                           delta) {
  p <- r + k + f
  des <- build_design(r, k, f, t_interval)
  R <- build_R_matrix(p, sigma2)
  Z <- des$Z
  Sigma <- R + Z %*% matrix(sigma_b2) %*% t(Z)
  L <- chol(Sigma)

  generate_group <- function(n, g) {
    X <- if (g == 1) des$X_trt else des$X_plc
    nc <- ncol(X)
    beta_true <- if (nc == 2) {
      c(1.0, delta * g)
    } else {
      c(1.0, 1.0, delta * g)
    }
    mu <- X %*% beta_true
    replicate(n, as.numeric(mu + t(L) %*% rnorm(p)))
  }

  Y_trt <- generate_group(m, 1)
  Y_plc <- generate_group(m, 0)
  list(Y_trt = Y_trt, Y_plc = Y_plc, des = des)
}

analyze_gls <- function(trial, sigma2, sigma_b2,
                        t_interval) {
  des <- trial$des
  p <- des$p
  nc <- des$ncol_X
  R <- build_R_matrix(p, sigma2)
  Ri <- solve(R)
  Z <- des$Z
  G <- matrix(sigma_b2)
  H <- solve(solve(G) + t(Z) %*% Ri %*% Z)

  Xt <- des$X_trt
  Xp <- des$X_plc

  XSi_trt <- t(Xt) %*% (Ri - Ri %*% Z %*% H %*%
                          t(Z) %*% Ri)
  XSi_plc <- t(Xp) %*% (Ri - Ri %*% Z %*% H %*%
                          t(Z) %*% Ri)

  info <- XSi_trt %*% Xt + XSi_plc %*% Xp
  m <- ncol(trial$Y_trt)
  info_total <- m * info
  V <- solve(info_total)

  rhs <- rowSums(apply(trial$Y_trt, 2, function(y) {
    XSi_trt %*% y
  })) + rowSums(apply(trial$Y_plc, 2, function(y) {
    XSi_plc %*% y
  }))

  beta_hat <- V %*% rhs
  se <- sqrt(V[nc, nc])
  gamma_hat <- beta_hat[nc]
  list(estimate = gamma_hat, se = se,
       z = gamma_hat / se)
}

analyze_avg <- function(trial) {
  des <- trial$des
  # r = run-in count: 0 when the model has only (beta, gamma);
  # otherwise the number of nonzero entries in the delta column.
  r <- if (des$ncol_X == 2) 0 else sum(des$X_trt[, 1] != 0)
  p <- des$p
  n_post <- p - r

  avg_score <- function(y) {
    if (r == 0) {
      mean(y)
    } else {
      mean(y[(r + 1):p]) - mean(y[1:r])
    }
  }

  d_trt <- apply(trial$Y_trt, 2, avg_score)
  d_plc <- apply(trial$Y_plc, 2, avg_score)
  m <- length(d_trt)

  # rescale the contrast to the gamma scale (Section 3.1)
  gcol <- des$X_trt[, des$ncol_X]
  cc <- mean(gcol[(r + 1):p])

  gamma_hat <- (mean(d_trt) - mean(d_plc)) / cc
  se <- sqrt(var(d_trt) / m + var(d_plc) / m) / cc
  list(estimate = gamma_hat, se = se,
       z = gamma_hat / se)
}
```

## Results

\begin{bullets}
\textbf{The simulation confirms the analytic ordering and quantifies the cost of misspecifying the random-slope variance.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The empirical standard deviation of the GLS estimator is the smallest of the three under correct specification.
\item On the $\gamma$ scale the averaged estimator is nearly as efficient as GLS in the no-run-in design, but carries two to three times the GLS standard deviation in the run-in designs at $\sigma_b/\sigma = 1$, growing to eight to fifteen times at $\sigma_b/\sigma = 5$.
\item Misspecifying the random-slope variance at most roughly doubles the GLS standard deviation, which remains well below that of the averaged estimator in every run-in cell.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
We simulate the factorial design described above under
compound symmetry, comparing the correctly specified GLS
estimator, a GLS estimator that misstates the random-slope
variance (it assumes $\sigma_b^2 = \sigma^2$ regardless of the
truth), and the averaged change-score estimator, the latter
rescaled to the $\gamma$ scale by dividing the contrast by
$c$. For each cell we report the empirical standard deviation
of $\hat\gamma$ and the empirical power from 1000 replicates;
the Monte Carlo standard error of a tabulated power estimate
is at most $\sqrt{0.25/1000} \approx 0.016$.
:::

```{r simulation-study}
nsim <- 1000  # extension to 5000 replicates is future work

sim_cell <- function(m, r, k, f, ratio, nsim, delta = 0.5,
                     t_interval = 1, sigma2 = 1) {
  sigma_b2 <- ratio^2 * sigma2
  za <- qnorm(0.975)
  eg <- sg <- ea <- sa <- em <- sm <- numeric(nsim)
  for (s in seq_len(nsim)) {
    tr <- simulate_trial(m, r, k, f, sigma2, sigma_b2,
                         t_interval, delta)
    g   <- analyze_gls(tr, sigma2, sigma_b2, t_interval)
    mis <- analyze_gls(tr, sigma2, sigma2, t_interval)
    av  <- analyze_avg(tr)
    eg[s] <- g$estimate;   sg[s] <- g$se
    em[s] <- mis$estimate; sm[s] <- mis$se
    ea[s] <- av$estimate;  sa[s] <- av$se
  }
  pw <- function(e, s) mean(abs(e / s) > za)
  data.frame(
    design = sprintf('(%d,%d,%d)', r, k, f),
    ratio = ratio, m = m,
    sd_gls = sd(eg), sd_mis = sd(em), sd_avg = sd(ea),
    pow_gls = pw(eg, sg), pow_mis = pw(em, sm),
    pow_avg = pw(ea, sa)
  )
}

set.seed(20260418)
designs <- list(c(0, 2, 0), c(2, 2, 0), c(2, 2, 2))
cells <- expand.grid(design = 1:3, ratio = c(1, 5),
                     m = c(50, 100))
sim_res <- do.call(rbind, lapply(seq_len(nrow(cells)),
  function(i) {
    d <- designs[[cells$design[i]]]
    sim_cell(cells$m[i], d[1], d[2], d[3], cells$ratio[i],
             nsim)
  }))
sim_res <- sim_res[order(sim_res$ratio, sim_res$m,
                         sim_res$design), ]

knitr::kable(
  sim_res,
  col.names = c('Design', '$\\sigma_b/\\sigma$', '$m$',
                'SD GLS', 'SD mis', 'SD avg',
                'Pow GLS', 'Pow mis', 'Pow avg'),
  caption = paste0(
    'Monte Carlo comparison of GLS (correctly specified), ',
    'GLS (misspecified $\\sigma_b^2$, assuming ',
    '$\\sigma_b^2=\\sigma^2$), and the averaged change-score ',
    'estimator under compound symmetry (', nsim,
    ' replicates, $\\Delta=0.5$, two-sided $\\alpha=0.05$). ',
    'Columns report the empirical standard deviation of ',
    '$\\hat\\gamma$ and the empirical power. The averaged ',
    'estimator is rescaled to the $\\gamma$ scale ',
    '(contrast divided by $c$).'),
  booktabs = TRUE, escape = FALSE, digits = 3,
  row.names = FALSE)
```

One feature of the table requires comment. In the
$(0,2,0)$ rows the correctly specified and misspecified
GLS analyses report identical empirical standard
deviations but different powers. With $J_0 = 0$ the
design has two observations and two fixed effects, so
the GLS point estimator does not depend on the assumed
covariance and the two analyses coincide as estimators;
the apparent power advantage of the misspecified
analysis reflects only an anticonservative model-based
standard error, not additional information. More
generally, because empirical size at $\Delta = 0$ and
interval coverage are not tabulated, power comparisons
across variance models in this table are not
interpretable as efficiency comparisons; adding size and
coverage columns is remaining work.


# Numerical examples

## MIRIAD parameters

```{r miriad-comparison}
# MIRIAD within-visit residual = between-visit (0.3338) +
# between-scan measurement error (0.0025); see Paper 1.
sigma2_ad <- 0.3338 + 0.0025
sigma_b2_ad <- 0.9745

configs <- expand.grid(
  r = c(0, 1, 2, 4),
  k = c(1, 2, 3),
  f = c(0, 1)
)
configs <- configs[configs$r + configs$k +
                     configs$f >= 2, ]

configs$v_gls <- mapply(
  function(r, k, f)
    var_gamma_matrix(r, k, f, sigma2_ad,
                     sigma_b2_ad, 1.0),
  configs$r, configs$k, configs$f
)

configs$v_anc <- mapply(
  function(r, k, f)
    var_gamma_ancova(r, k, f, sigma2_ad,
                     sigma_b2_ad, 1.0) /
      cfac(r, k, f, 1.0)^2,
  configs$r, configs$k, configs$f
)

configs$v_avg <- mapply(
  function(r, k, f)
    var_gamma_avg(r, k, f, sigma2_ad,
                  sigma_b2_ad, 1.0) /
      cfac(r, k, f, 1.0)^2,
  configs$r, configs$k, configs$f
)

configs$label <- sprintf('(%d,%d,%d)',
  configs$r, configs$k, configs$f)

delta_ad <- 0.25
za <- qnorm(0.975)
zb <- qnorm(0.90)
configs$n_gls <- ceiling(
  (za + zb)^2 * configs$v_gls / delta_ad^2)
configs$n_anc <- ceiling(
  (za + zb)^2 * configs$v_anc / delta_ad^2)
configs$n_avg <- ceiling(
  (za + zb)^2 * configs$v_avg / delta_ad^2)

display <- configs[order(configs$v_gls),
  c('label', 'n_gls', 'n_anc', 'n_avg')]

knitr::kable(
  head(display, 15),
  col.names = c('$(J_0,J_1,J_2)$', '$N$ GLS', '$N$ ANC',
                '$N$ Avg'),
  caption = paste0(
    'Sample size per group: GLS, ANCOVA, and ',
    'averaged estimators (MIRIAD parameters, ',
    '$\\Delta=', delta_ad, '$, 90\\% power). All ',
    'variances are on the $\\gamma$ scale (ANCOVA ',
    'and averaged contrasts divided by $c$).'),
  booktabs = TRUE,
  escape = FALSE,
  row.names = FALSE
)
```

## When to use which estimator

A decision rule based on $m$, $\sigma_b/\sigma$, and
confidence in the correlation structure requires the
feasible-GLS simulation described in the Discussion and
is deferred to a revision. The results above support
only the narrower statement that, with variance
components taken as known, GLS dominates and the
penalty of the simpler estimators is largest in run-in
designs with $J_1 \ge 2$ and high $\sigma_b/\sigma$.


# Discussion

\begin{bullets}
\textbf{GLS is the efficient choice under a correctly specified, adequately estimated model, which in practice means moderate-to-large samples.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item GLS attains the minimum variance among linear unbiased estimators when the covariance is known.
\item Its advantage is realized only when the variance components are estimated accurately, which requires sufficient sample size and follow-up.
\item In small samples the instability of the estimated covariance can erase the theoretical gain.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
The estimator comparison yields a graded recommendation rather
than a single winner. Under correct specification GLS is
efficient and dominates both ANCOVA and the averaged
change-score estimator, as the ordering
$\Var_{\text{GLS}} \le \Var_{\text{ANC}} \le
\Var_{\text{avg}}$ confirms. This dominance is a large-sample
statement: it presumes the covariance components are known or
well estimated. With moderate-to-large samples and a credible
covariance model, GLS is the appropriate default.
:::

\begin{bullets}
\textbf{The misspecifications studied here erode the GLS advantage without overturning it; the open robustness question concerns feasible GLS with estimated components.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item GLS weights observations by the assumed inverse covariance, so errors in $G$ or $R$ propagate into its sampling variance.
\item In the sweeps of Section 4 the misspecified GLS sandwich variance peaks at 5.30 against a $\gamma$-scale averaged-estimator variance of 8.89, and in the simulation the misspecified GLS standard deviation remains well below the averaged standard deviation in every run-in cell.
\item Whether feasible GLS, with variance components estimated from small samples, retains this margin is not studied here.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
The misspecification analysis qualifies the case for GLS less
than might be expected. Because GLS weights the data by an
assumed inverse covariance, errors in the assumed $G$ or $R$
feed into its sampling variance, and the sandwich-variance
calculations quantify the inflation. In the settings studied,
however, the inflation never closes the gap: the misspecified
GLS variance peaks at 5.30 in the $G$ sweep and at 2.16 in the
$R$ sweep, against $\gamma$-scale averaged-estimator variances
of 8.4 to 9.1 at the same design. The averaged change-score
estimator avoids the
variance components entirely, and ANCOVA estimates the run-in
coefficient from the data, but under the misspecifications
examined neither robustness property is enough to overcome
their efficiency deficit. These calculations take the variance
components of the GLS analysis as fixed inputs; the open and
practically important question is whether feasible GLS, with
components estimated by REML from small samples, keeps its
advantage. That question is not answered here.
:::

\begin{bullets}
\textbf{The efficiency gap between GLS and the simpler estimators narrows as the design grows, so the choice matters most in small, information-poor designs.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item With many run-in and treatment observations the phase means become near-sufficient for the random slope.
\item The averaged and ANCOVA estimators then recover most of the information GLS extracts, and the relative efficiency approaches one.
\item The estimator choice is therefore most consequential when observations are few and the slope-to-noise ratio is high.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
The practical importance of the estimator choice depends on the
design. As the numbers of run-in and treatment observations
grow, the phase means approach sufficiency for the
subject-specific slope, the ANCOVA efficiency loss vanishes as
$J_0 \to \infty$, and the relative efficiencies of all three
estimators converge toward one. The gap is largest, and the
choice of estimator therefore most consequential, in small
designs with few observations per phase and a high
slope-to-noise ratio, exactly the regime in which the
estimated covariance is also least stable. Whether that
instability erodes the feasible-GLS advantage enough to
change the ranking in small studies is the question the
planned feasible-GLS simulation must answer; the
known-components results here cannot.
:::

\begin{bullets}
\textbf{Two extensions, the seemingly unrelated regression estimator and the constrained longitudinal data analysis estimator, would complete the efficiency picture.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The SUR formulation offers a middle ground between the averaged estimator and full GLS but adds implementation complexity.
\item The cLDA estimator exploits the full covariance without specifying the random-effects distribution and is expected to rank between ANCOVA and GLS.
\item A full comparison including both, and a feasible-GLS simulation under estimated components, is the natural continuation of this work.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
Two estimators not fully developed here would round out the
comparison. The seemingly unrelated regression formulation of
Section 2.4 treats the run-in and treatment periods as linked
equations and offers a compromise between the averaged
estimator and full GLS, at the cost of additional
implementation complexity; its variance derivation remains to
be completed. The constrained longitudinal data analysis
estimator uses the full covariance without committing to a
random-effects distribution and should fall between ANCOVA
and GLS in efficiency. A comprehensive treatment would compare
all of these under a feasible-GLS simulation in which the
variance components are estimated rather than assumed, which is
the setting in which the robustness distinctions drawn here
become decisive.
:::

# Conflict of interest

The author declares no conflicts of interest.

# Funding

This work received no specific funding.

# Data availability statement

\begin{bullets}
\textbf{The package implementing all three estimators, the simulation drivers, and the shared bibliography are openly available in the project repository.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The estimator implementations reside in the package source files.
\item The simulation drivers and shared compendium bibliography are included.
\item Specific paths appear in the Reproducibility section below.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
The `runinpower` R package implementing the GLS, ANCOVA, and
averaged-change-score estimators (`ancova.R`, `averaged.R`),
the simulation drivers, and the shared bibliography across the
four-paper compendium are openly available in the project
repository. Specific paths are listed in the Reproducibility
section below.
:::

# Reproducibility

\begin{bullets}
\textbf{The manuscript was rendered with R 4.4.x using the package source and a small set of standard analysis and build packages.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The in-package source supplies the estimator implementations.
\item The GLS comparator in the simulation is implemented directly in the manuscript source (the analyze functions of the simulation chunk); tinytest is the testing harness.
\item The manuscript is built with rmarkdown and bookdown.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
This manuscript was rendered with R 4.4.x. Principal packages
used are the in-package `runinpower` source, `tinytest`
(testing harness), and `rmarkdown` plus `bookdown` for the
manuscript build. The GLS comparator in the simulation is
implemented directly in the manuscript source
(`analyze_gls` in the simulation chunk); no external mixed
model fitter is used.
:::

\begin{bullets}
\textbf{The simulation sets a single fixed seed before the replicate loop, which makes the run exactly reproducible.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item A single seed is set once at the entry point of the Monte Carlo comparison.
\item No per-replicate reseeding is performed; the replicate stream follows from the single seed.
\item Re-running the chunk therefore reproduces the tabulated results exactly.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
**Random-number generator.** The Monte Carlo simulation
comparing the three estimators sets `set.seed(20260418)` once,
immediately before the replicate loop. No per-replicate
reseeding is performed, so the full replicate stream, and
hence the tabulated results, follow deterministically from
that single seed.
:::

\begin{bullets}
\textbf{All code, drivers, companion manuscripts, and the shared bibliography are located at documented paths within the compendium.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The package source and simulation drivers reside in their respective directories.
\item The three companion manuscripts are stored alongside this paper in the compendium.
\item The manuscript source and shared bibliography are co-located in the paper directory.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
**Data and code availability.** The R package source is at
`R/`. Simulation drivers are at `analysis/scripts/`. The
companion manuscripts (Papers 1, 2, 3) are at
`analysis/report/{01,02,03}-*/`. The manuscript source and
shared bibliography (`references.bib`, a symlink to
`../../references.bib`) are at
`analysis/report/04-gls-vs-ttest/`.
:::

# References


::: {#refs}
:::

```{r footer, echo=FALSE, results='asis'}
src <- sub("^/Users/zenn/Dropbox/prj/", "~/prj/",
           knitr::current_input(dir = TRUE))
cat(sprintf(
  "\n\\vfill\n\n---\n\n*Rendered on %s.*  \n*Source: `%s`*\n",
  format(Sys.time(), "%Y-%m-%d at %H:%M %Z"),
  src
))
```
