---
title: |
  Power Analysis for Run-In Clinical Trials under
  AR(1) and General Correlation Structures
author: "Ronald G. Thomas, Ph.D.\\thanks{Wertheim School of Public Health, UCSD; ORCID: 0000-0003-1686-4965}"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
affiliation: "University of California, San Diego"
abstract: |
  **Background.** The companion Paper 1 of this compendium
  derived closed-form power formulas for run-in and common
  close observations under compound symmetry. Compound
  symmetry assumes all pairs of observations are equally
  correlated regardless of temporal separation, an assumption
  that is empirically rejected in most longitudinal-trial data.
  AR(1), $\Corr(Y_{ijk}, Y_{ijk'}) = \rho^{|k-k'|}$, is the
  natural alternative.

  **Methods.** We derive the variance of the treatment-effect
  estimator under AR(1) via two routes: (i) the Woodbury
  identity applied to the AR(1) marginal covariance, and (ii)
  a moment-based approach using the cross-covariance between
  run-in and treatment-period means. The two routes target
  different estimands: the GLS variance of the slope-difference
  effect and the variance of the simple phase-mean difference.
  The GLS derivation accommodates arbitrary $J_0$ run-in
  observations and arbitrary $J_2$ common close observations.

  **Results.** A taxonomy of five correlation cases is
  presented, and change-score correlations are derived in
  closed form under compound symmetry and AR(1). Numerical
  comparisons using parameters from Alzheimer's disease
  neuroimaging trials show that, when the two structures are
  matched at a common nominal correlation parameter $\rho$,
  the AR(1) sample-size estimate is smaller than the
  compound-symmetric estimate (by 4-36\% at $\rho = 0.5$),
  with the deviation growing as $\rho$ increases. A narrow
  anticonservative regime exists at large $J_0$ and small
  $\rho$, where compound symmetry understates the variance
  by at most about 6\%.

  **Conclusions.** At a matched nominal $\rho$, the
  compound-symmetry assumption underlying Paper 1 is generally
  conservative: it overstates the required sample size, and
  increasingly so as the autocorrelation grows. The exception
  is a narrow regime with many run-in observations and low
  autocorrelation, where it is anticonservative by at most
  about 6\%. The AR(1) formulas derived here quantify both
  effects and should be used directly when serial correlation
  is expected; this conclusion is specific to the
  common-nominal-$\rho$ matching convention stated in the
  text.
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    number_sections: true
    keep_tex: true
    highlight: tango
    includes:
      in_header: claudecode.tex
fontsize: 11pt
geometry: "left=3cm,right=3cm,top=2cm,bottom=2cm"
bibliography: references.bib
knit: (function(input, ...) { d <- dirname(input); while (!file.exists(file.path(d, 'tools', 'stamp-render.R')) && d != dirname(d)) d <- dirname(d); source(file.path(d, 'tools', 'stamp-render.R'))$value(input) })
csl: statistics-in-medicine.csl
link-citations: true
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{lineno}
  - \linenumbers
  - \usepackage{booktabs}
  - \usepackage{float}
  - \DeclareMathOperator{\Var}{Var}
  - \DeclareMathOperator{\Cov}{Cov}
  - \DeclareMathOperator{\Corr}{Corr}
  - \DeclareMathOperator{\tr}{tr}
  - \usepackage{setspace}
  - \setstretch{1.1}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4,
  fig.pos = "H"
)
library(runinpower)
```

# Introduction

\begin{bullets}
\textbf{The efficiency of run-in designs is governed by the correlation structure of the repeated measurements.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Under compound symmetry the correlation between change scores is constant.
\item Run-in observations then inform the random slope through simple averaging.
\item Under AR(1) the correlation decays with separation, so run-in information depends on distance from the treatment period.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
The efficiency of run-in designs depends on the
correlation structure of the repeated measurements.
Under compound symmetry, the correlation between any
two change scores is constant, and the run-in
observations contribute information about the random
slope $b_i$ through a simple averaging mechanism. Under
AR(1), the correlation decays with temporal separation,
and the information content of run-in observations
depends on how far they are from the treatment period.
:::

\begin{bullets}
\textbf{This paper extends prior compound-symmetric run-in frameworks to AR(1) residual correlation.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Paper 1 assumed the rank-one-plus-identity change-score covariance from independent errors.
\item Wang (2019) extended run-in to two-period mixed models but kept compound symmetry.
\item We instead adopt AR(1) residual correlation with geometric decay in lag.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
Paper 1 [@frost2008] assumed $R = \sigma^2(I_J +
\mathbf{1}_J\mathbf{1}_J')$, which arises from the
change-score parameterization under independent errors.
@wang2019 extended the run-in framework to two-period
linear mixed models but retained the compound symmetry
assumption. We extend here to AR(1) residual
correlation, where $\Corr(e_{ij}, e_{ij'}) =
\rho^{|j-j'|}$ for $0 < \rho < 1$.
:::

\begin{bullets}
\textbf{Existing literature treats autocorrelation and change-score variation but not the run-in setting.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Winkens (2007) studied optimal designs for autocorrelated data outside the run-in context.
\item Diggle and Verbeke give general treatments of longitudinal correlation structures.
\item Chambless (1993) addressed intra-individual variation in change scores, a concern amplified under AR(1).
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
@winkens2007 studied optimal designs for autocorrelated
longitudinal data but did not consider the run-in
setting. @diggle2002 and @verbeke2000 provide general
treatments of correlation structures in longitudinal
models. @chambless1993 addressed the effect of
intra-individual variation on change-score analyses,
a concern that is amplified under AR(1) correlation.
:::

\begin{bullets}
\textbf{The closed-form AR(1) derivation covers a design case the reference longpower toolkit does not.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Its slope-based closed-form routines assume independent or compound-symmetric residuals.
\item Its \texttt{power.mmrm.ar1()} routine implements the analytic AR(1) formula of Lu, Luo, and Chen (2008), but without run-in or common close observations under a random slope.
\item Our formula supplies that case and collapses to the compound-symmetric structure of Paper 1 as $\rho \to 0$.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
The `longpower` R package [@iddi2022] is the CRAN
reference implementation for longitudinal-trial power
calculations. Its `power.mmrm.ar1()` routine implements
the analytic AR(1) sample-size formula of @lu2008 for
mixed models for repeated measures, and its slope-based
closed-form routines (`edland.linear.power()`,
`diggle.linear.power()`) assume random intercepts and
slopes with independent or compound-symmetric residuals.
None of these routines accommodates run-in or common
close observations under a random-slope model. The
closed-form AR(1) derivation developed here supplies
that case. As $\rho \to 0$ (no serial correlation), the
AR(1) change-score covariance collapses to the
compound-symmetric structure of Paper 1, so the present
formulas contain the Paper 1 results as a special case.
:::


# Correlation structures for longitudinal trials

## Taxonomy of cases

We consider five correlation structures for the
raw outcomes $Y_{ijk}$:

1. **Compound symmetry:**
   $\Corr(Y_{ijk}, Y_{ijk'}) = \rho$ for all
   $k \ne k'$. This is equivalent to the random
   intercept model and produces the $R = \sigma^2(I +
   \mathbf{1}\mathbf{1}')$ structure in change scores.

2. **AR(1):**
   $\Corr(Y_{ijk}, Y_{ijk'}) = \rho^{|k'-k|}$.
   Correlation decays geometrically with lag.

3. **Compound symmetry with variable run-in:**
   Case 1 with $J_{0,i}$ varying across participants.

4. **AR(1) with variable run-in and common close:**
   Case 2 with $J_{0,i}$ and $J_{2,i}$ varying.

5. **General (unstructured):**
   $\Corr(Y_{ijk}, Y_{ijk'}) = \sigma_{|k-k'|}/\sigma^2$.

Throughout, the design quantities $J_0$, $J_1$, and
$J_2$ correspond to the arguments `r`, `k`, and `f` of
the implementing `runinpower` functions.


## Change-score correlations under each structure

Under the consecutive-difference parameterization
($C_j = Y_j - Y_{j-1}$, successive visits separated by
$t$), the correlation between change scores depends on
the raw-outcome correlation structure. Here
$a = \sigma^2$ and $b = \sigma_b^2$. (The GLS machinery
of the next section uses change scores from baseline;
the two parameterizations are related by a nonsingular
linear map, so the GLS variance is unaffected.)

For the run-in design with $J_0 = 2$, $J_1 = 1$, with
visits at times $-2t, -t, 0, t$ and consecutive
differences $C_1, C_2, C_3$:

**Compound symmetry (independent raw errors; the shared
intercept component cancels in differencing):**
\begin{align}
  \Corr(C_1, C_3) &= \frac{bt^2}{2a + bt^2} \\
  \Corr(C_1, C_2) &= \frac{-a + bt^2}{2a + bt^2}
\end{align}

**AR(1):** for consecutive differences at lag
$m = |j - l| \ge 1$,
\begin{align}
  \Var(C_j) &= bt^2 + 2a(1 - \rho) \\
  \Cov(C_j, C_l) &= bt^2 - a\,\rho^{m-1}(1 - \rho)^2
\end{align}
so that
\begin{align}
  \Corr(C_1, C_3) &=
    \frac{bt^2 - a\rho(1 - \rho)^2}
         {bt^2 + 2a(1 - \rho)} \\
  \Corr(C_1, C_2) &=
    \frac{bt^2 - a(1 - \rho)^2}
         {bt^2 + 2a(1 - \rho)}
\end{align}
These expressions hold for arbitrary $(J_0, J_1, J_2)$
because they depend only on the lag $m$; at $\rho = 0$
they reduce to the compound-symmetry formulas above.


# Variance under AR(1): GLS approach

## Marginal covariance

Under AR(1) with residual variance $\sigma^2$ and
autocorrelation $\rho$, the raw-outcome covariance is
$\Cov(e_{ij}, e_{ij'}) = \sigma^2 \rho^{|j-j'|}$.
The change-score residual covariance becomes:
\begin{equation}
  R_{jl} = \Cov(e_{ij} - e_{i0}, e_{il} - e_{i0})
  = \sigma^2(\rho^{|j-l|} - \rho^{|j|} - \rho^{|l|} + 1)
\end{equation}
for $j, l \ne 0$.

\begin{bullets}
\textbf{The AR(1) change-score covariance loses rank-one structure but the Woodbury reduction survives.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The matrix $R$ is no longer rank-one-plus-identity, so Sherman-Morrison does not apply.
\item The Woodbury identity still reduces the inverse covariance to a scalar inversion.
\item This holds because the random-effects covariance $G$ remains scalar.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
This $R$ no longer has the rank-one-plus-identity
structure, so the Sherman-Morrison formula does not
apply. However, the Woodbury identity still reduces
$\Sigma^{-1}$ to a scalar inversion because $G$ remains
scalar.
:::

## Woodbury computation

The computation proceeds identically to Paper 1:
\begin{equation}
  \Sigma^{-1} = R^{-1} - R^{-1}Z
    (G^{-1} + Z'R^{-1}Z)^{-1} Z'R^{-1}
\end{equation}
\begin{bullets}
\textbf{The Woodbury scalar term persists, but the residual inverse must now be computed differently.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The correction term $H$ remains a scalar as in Paper 1.
\item The residual inverse $R^{-1}$ must be obtained numerically.
\item Equivalently it follows from the AR(1) tridiagonal inverse rather than Sherman-Morrison.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
with $H = (G^{-1} + Z'R^{-1}Z)^{-1}$ still scalar.
The difference is that $R^{-1}$ must be computed
numerically (or via the AR(1) tridiagonal inverse)
rather than by Sherman-Morrison.
:::

```{r ar1-variance}
a <- 1.0
b <- 1.0
tt <- 1.0
rho_values <- c(0, 0.3, 0.5, 0.7, 0.9)
configs <- expand.grid(
  r = c(0, 2, 4),
  k = 2,
  f = 0,
  rho = rho_values
)

configs$v_cs <- mapply(
  function(r, k, f, ...) {
    var_gamma_matrix(r, k, f, a, b, tt)
  },
  configs$r, configs$k, configs$f
)

configs$v_ar1 <- mapply(
  function(r, k, f, rho) {
    var_gamma_ar1(r, k, f, a, b, tt, rho)
  },
  configs$r, configs$k, configs$f, configs$rho
)

tab <- reshape(
  configs[, c('r', 'rho', 'v_ar1')],
  idvar = 'r', timevar = 'rho',
  direction = 'wide'
)
names(tab) <- c('$J_0$', paste0('$\\rho=', rho_values, '$'))

knitr::kable(
  tab,
  caption = paste0(
    'Per-pair $\\Var(\\hat\\gamma)$ under AR(1) ',
    'for varying $\\rho$ ($J_1=2$, $J_2=0$, ',
    '$\\sigma^2=', a, '$, $\\sigma_b^2=', b, '$).'),
  booktabs = TRUE,
  escape = FALSE,
  digits = 4
)
```


# Variance under AR(1): moment-based approach

## Cross-covariance between phase means

\begin{bullets}
\textbf{A moment-based alternative derives the pre-post variance directly from the phase means.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item It computes the variance of the test statistic from the moments of the phase means.
\item The run-in mean averages the run-in observations.
\item The treatment-period mean averages the treatment-period observations.
\item The construction follows the summary-measures approach of Frison and Pocock (1992).
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
An alternative to the GLS approach computes the
variance of the pre-post test statistic directly
from the moments of the phase means, following the
summary-measures approach of @frison1992, specialized
here to AR(1) correlation. Let
$\bar{X}_0 = \frac{1}{J_0}\sum_{i=1}^{J_0} Y_{i,\text{run-in}}$
and
$\bar{X}_K = \frac{1}{J_1}\sum_{j=1}^{J_1} Y_{j,\text{trt}}$
denote the run-in and treatment-period means.
:::

Under AR(1) with parameter $c = \rho$, the
cross-covariance is:
\begin{equation}
  \Cov(\bar{X}_0, \bar{X}_K)
  = \frac{\sigma^2}{J_0 J_1}\,\rho\, c^{K-1}\,
    \frac{(1 - c^{J_0})(1 - c^{J_1})}{(1 - c)^2}
\end{equation}
where $c = \rho$ (the symbol $c$ here denotes the
autocorrelation and is unrelated to the
reparameterization $c = \sigma^2 - \sigma_b^2 t^2$ used
in Paper 1) and $K$ is the time-index lag from the last
run-in visit to the first treatment visit on the visit
grid $d = (-J_0, \ldots, -1, 1, \ldots, J_1)$. Because
baseline (time 0) is not itself a phase observation,
adjacent phases have $K = 2$. The division by
$J_0 J_1$ reflects that $\bar{X}_0$ and $\bar{X}_K$ are
means of $J_0$ and $J_1$ observations respectively; this
matches the implementation in `ar1_cov_phase_means()`
below.

## Variance of the change score

The variance of the pre-post difference is:
\begin{equation}
  \Var(X_K - X_0) = \Var(X_K) + \Var(X_0)
    - 2\Cov(X_K, X_0)
\end{equation}

The variance of each phase mean involves the sum
$S_n = \sum_{i=1}^{n-1}(n-i)c^{i-1}$, which has the
closed form:
\begin{equation}
  S_n = \frac{n(1-c) - (1-c^n)}{(1-c)^2}
\end{equation}

```{r moment_based}
ar1_cov_phase_means <- function(r, q, K, sigma2, rho) {
  if (r == 0 || q == 0) return(0)
  c <- rho
  sigma2 / (r * q) * rho * c^(K - 1) *
    (1 - c^r) * (1 - c^q) / (1 - c)^2
}

ar1_var_mean <- function(n, sigma2, rho) {
  if (n == 1) return(sigma2)
  c <- rho
  S_n <- (n * (1 - c) - (1 - c^n)) / (1 - c)^2
  sigma2 / n * (1 + 2 * rho / n * S_n)
}

ar1_var_change <- function(r, q, K, sigma2, rho) {
  vx0 <- ar1_var_mean(r, sigma2, rho)
  vxk <- ar1_var_mean(q, sigma2, rho)
  cov_xk_x0 <- ar1_cov_phase_means(r, q, K, sigma2,
                                     rho)
  vx0 + vxk - 2 * cov_xk_x0
}

cat('Moment-based vs GLS variance (r=2, k=2, rho=0.5):\n')
v_moment <- ar1_var_change(r = 2, q = 2, K = 2,
                           sigma2 = a, rho = 0.5)
v_gls0 <- var_gamma_ar1(2, 2, 0, a, 0, tt, 0.5)
v_gls <- var_gamma_ar1(2, 2, 0, a, b, tt, 0.5)
cat(sprintf('  Moment-based phase-mean difference: %.4f\n',
            v_moment))
cat(sprintf('  GLS slope difference (sigma_b2=0):  %.4f\n',
            v_gls0))
cat(sprintf('  GLS slope difference (sigma_b2=%.0f):  %.4f\n',
            b, v_gls))
cat('  (Different estimands; see next section.)\n')
```


# Relationship between the two approaches

\begin{bullets}
\textbf{The GLS and moment-based approaches target distinct variance quantities.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The GLS variance is the minimum-variance bound under the assumed model.
\item The moment-based variance is that of the simple averaged change-score estimator.
\item The two therefore answer different efficiency questions.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
The GLS approach (Section 3) and the moment-based
approach (Section 4) compute different quantities.
The GLS variance is the minimum-variance bound under
the assumed model, while the moment-based variance
is the variance of the simple averaged change-score
estimator $\bar{d}_1 - \bar{d}_0$.
:::

\begin{bullets}
\textbf{The two approaches do not coincide even without a random slope.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item At $\sigma_b^2 = 0$ the GLS estimator still weights observations through the AR(1) residual covariance and targets the slope difference.
\item The moment-based quantity is the variance of the unweighted phase-mean difference, a different estimand.
\item Numerically, at $J_0 = 2$, $J_1 = 2$, $\rho = 0.5$, $\sigma^2 = 1$, $\sigma_b^2 = 0$: GLS 0.636 versus moment-based 1.219.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
The two approaches do not coincide even when
$\sigma_b^2 = 0$. At that boundary the GLS estimator
still weights the change scores through the AR(1)
residual covariance and estimates the slope-difference
parameter $\gamma$, whereas the moment-based quantity is
the variance of the unweighted phase-mean difference
$\bar{X}_K - \bar{X}_0$. With $J_0 = 2$, $J_1 = 2$,
$\rho = 0.5$, $\sigma^2 = 1$, and $\sigma_b^2 = 0$, the
GLS variance is 0.636 while the moment-based variance
is 1.219. When $\sigma_b^2 > 0$ the GLS estimator
additionally exploits the random-slope structure of
$\Sigma = R + ZGZ'$.
:::

\begin{bullets}
\textbf{The moment-based computation serves the simple pre-post estimator, not as a bound on the GLS variance.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item It requires no specification of the random-slope variance.
\item It is the exact variance of the unweighted pre-post difference an analyst might report without fitting a mixed model.
\item Because the estimands differ, its ratio to the GLS variance is descriptive, not an efficiency bound.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
The practical value of the moment-based computation is
that it requires no specification of $\sigma_b^2$ and
gives the exact variance of the simple pre-post
difference that an analyst might report without fitting
a mixed model. It is not a bound on the GLS variance: at
the parameters above the moment-based value (1.219) lies
between the GLS variance at $\sigma_b^2 = 0$ (0.636) and
at $\sigma_b^2 = 1$ (1.666). Because the two quantities
are variances of different estimands, their ratio
describes the cost of abandoning the mixed model rather
than an efficiency ordering within it. The
relative-efficiency question is taken up in Paper 4.
:::


# Robustness to correlation misspecification

The practical question is whether a trial designed
under compound symmetry loses substantial efficiency
when the true correlation is AR(1), and vice versa.
Throughout this section the two structures are matched
at a common nominal correlation parameter $\rho$; the
comparison, and its direction, depend on this matching
convention.

```{r robustness, fig.cap="Ratio of AR(1) to compound-symmetry variance at a matched nominal $\\rho$ for three run-in configurations ($J_1=2$, $J_2=0$, $\\sigma^2=1$, $\\sigma_b^2=1$). The ratio remains near or below one except at $J_0=4$ and small $\\rho$, where it peaks near 1.06."}
rho_grid <- seq(0, 0.95, by = 0.05)
r_values <- c(0, 2, 4)

plot(NULL, xlim = c(0, 1), ylim = c(0, 1.25),
     xlab = expression(rho),
     ylab = expression(V[AR(1)] / V[CS]),
     main = 'Variance ratio: AR(1) / Compound Symmetry')
abline(h = 1, lty = 2, col = 'gray')

cols <- c('black', 'blue', 'red')
for (i in seq_along(r_values)) {
  r <- r_values[i]
  v_cs <- var_gamma_matrix(r, 2, 0, a, b, tt)
  ratios <- sapply(rho_grid, function(rho) {
    if (rho == 0) return(1)
    var_gamma_ar1(r, 2, 0, a, b, tt, rho) / v_cs
  })
  lines(rho_grid, ratios, col = cols[i], lwd = 2)
}
legend('bottomleft', legend = paste('J_0 =', r_values),
       col = cols, lwd = 2, bty = 'n')
```


# Stratified variance for unequal visit patterns

When participants have different numbers of run-in
($J_{0,i}$) or common close ($J_{2,i}$) observations, the
total variance is a weighted sum across strata:
\begin{equation}
  \frac{1}{n^2}\sum_{s=1}^{S} n_s\,
    \Var(X_K - X_0)\big|_{J_0 = J_{0,s},\, J_2 = J_{2,s}}
\end{equation}

<!-- TODO: develop stratified computation for -->
<!-- both CS and AR(1) -->


# Numerical examples

## Comparison across correlation structures

```{r comparison}
# MIRIAD within-visit residual = between-visit (0.3338) +
# between-scan measurement error (0.0025); see Paper 1.
sigma2_ad <- 0.3338 + 0.0025
sigma_b2_ad <- 0.9745
t_ad <- 1.0
delta_ad <- 0.25

configs_ad <- expand.grid(r = 0:4, k = 2, f = 0)
configs_ad$n_cs <- sapply(configs_ad$r, function(r) {
  sample_size(r, 2, 0, sigma2_ad, sigma_b2_ad,
              t_ad, delta_ad)
})

rho_ad <- 0.5
configs_ad$n_ar1 <- sapply(configs_ad$r, function(r) {
  v1 <- var_gamma_ar1(r, 2, 0, sigma2_ad, sigma_b2_ad,
                      t_ad, rho_ad)
  za <- qnorm(0.975)
  zb <- qnorm(0.90)
  ceiling((za + zb)^2 * v1 / delta_ad^2)
})

knitr::kable(
  configs_ad,
  col.names = c('$J_0$', '$J_1$', '$J_2$',
                '$N$ (CS)', '$N$ (AR(1), $\\rho=0.5$)'),
  caption = paste0(
    'Sample size per group (MIRIAD parameters, ',
    '$\\Delta=', delta_ad, '$, 90\\% power) under ',
    'compound symmetry vs AR(1).'),
  booktabs = TRUE,
  escape = FALSE
)
```

At the matched nominal $\rho = 0.5$, the AR(1) sample
sizes are smaller than the compound-symmetric ones at
every $J_0$, by 3.6\% ($J_0 = 0$) to 36.4\%
($J_0 = 1$). Positive serial correlation reduces the
residual variance of each change score
($R_{jj} = 2\sigma^2(1 - \rho^{|j|}) < 2\sigma^2$), so
the compound-symmetric calculation, which corresponds
to independent raw errors, is conservative here.

## Sensitivity to rho

```{r sensitivity, fig.cap="Per-pair variance as a function of $\\rho$ under AR(1) for MIRIAD parameters ($\\sigma^2=0.3363$, $\\sigma_b^2=0.9745$, $J_1=2$, $J_2=0$)."}
rho_grid <- seq(0, 0.95, by = 0.05)
r_values <- c(0, 1, 2, 4)

plot(NULL,
     xlim = c(0, 1),
     ylim = c(0, var_gamma_ar1(0, 2, 0, sigma2_ad,
                               sigma_b2_ad, t_ad,
                               0.01) * 1.2),
     xlab = expression(rho),
     ylab = expression(Var(hat(gamma))),
     main = 'MIRIAD parameters')

cols <- c('black', 'blue', 'red', 'darkgreen')
for (i in seq_along(r_values)) {
  r <- r_values[i]
  vs <- sapply(rho_grid, function(rho) {
    if (rho == 0) {
      var_gamma_matrix(r, 2, 0, sigma2_ad,
                       sigma_b2_ad, t_ad)
    } else {
      var_gamma_ar1(r, 2, 0, sigma2_ad,
                    sigma_b2_ad, t_ad, rho)
    }
  })
  lines(rho_grid, vs, col = cols[i], lwd = 2)
}
legend('topright', legend = paste('J_0 =', r_values),
       col = cols, lwd = 2, bty = 'n')
```


# Discussion

\begin{bullets}
\textbf{The autocorrelation structure matters for run-in designs only when the autocorrelation is weak relative to the random-slope signal.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item When the slope-to-noise ratio is large, the random-slope term dominates the change-score covariance and the residual correlation structure is second order.
\item The AR(1) and compound-symmetry variances then differ negligibly, and the simpler Paper 1 formula suffices.
\item The two structures diverge most when the slope-to-noise ratio is small and run-in observations are distant from randomization.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
The central practical question is when the compound-symmetry
formulas of Paper 1 can be used in place of the AR(1)
formulas derived here. The numerical comparisons indicate
that the answer is governed by the same slope-to-noise ratio
$\sigma_b^2 t^2 / \sigma^2$ that governs the value of run-in
observations themselves. When this ratio is large, the random
slope term $Z G Z'$ dominates the change-score covariance
$\Sigma = R + Z G Z'$, the residual term $R$ contributes
little, and the choice between an AR(1) and a
compound-symmetric $R$ is immaterial. When the ratio is small,
the residual term dominates and the correlation structure
becomes the leading determinant of efficiency.
:::

\begin{bullets}
\textbf{At a matched nominal $\rho$, compound symmetry is generally conservative, with a narrow anticonservative corner.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Positive serial correlation reduces the residual variance of every change score, so the AR(1) variance falls below the compound-symmetric value for most designs.
\item The discrepancy grows with $\rho$; near $\rho = 0.95$ the AR(1) variance can fall below one tenth of the compound-symmetric value.
\item The exception is large $J_0$ at small $\rho$, where distant run-in observations lose information and compound symmetry understates the variance by at most about 6\%.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
The direction of the error is important for trial design.
When the two structures are matched at a common nominal
$\rho$, positive serial correlation reduces the residual
variance of every change score
($R_{jj} = 2\sigma^2(1 - \rho^{|j|})$), so the AR(1)
variance lies below the compound-symmetric value for most
configurations and the compound-symmetric sample size is
conservative. In the MIRIAD example at $\rho = 0.5$ the
AR(1) sample sizes are 4 to 36 percent smaller than the
compound-symmetric ones, and the gap widens as $\rho$
grows. The exception is a narrow regime with many run-in
observations and low autocorrelation ($J_0 = 4$,
$\rho \approx 0.2$), where correlation decay makes
distant run-in observations less informative than
compound symmetry credits them to be; there the
compound-symmetric formula understates the variance, by
at most about 6 percent in the configurations examined.
Both statements are specific to the
common-nominal-$\rho$ matching convention; matching the
structures on, for example, a common lag-one
change-score correlation would alter the comparison.
:::

\begin{bullets}
\textbf{The numerical comparisons give a practical decision rule for when the simpler formula is safe.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item At a matched nominal $\rho$, the compound-symmetry formula of Paper 1 is a conservative and convenient default for most designs.
\item With many run-in observations and low expected autocorrelation, the AR(1) formula should be used directly; the compound-symmetric formula can understate the variance by up to about 6\% there.
\item When $\rho$ is uncertain, evaluating the AR(1) formula over a plausible range of $\rho$ and taking the largest variance is a safe default.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
For applied use we recommend the following. At a matched
nominal $\rho$, the compound-symmetry formula of Paper 1
is conservative for most designs and can serve as a
convenient default; the AR(1) formula then refines the
sample size downward. When the design places many run-in
observations before randomization and the expected
autocorrelation is low, the AR(1) formula should be used
directly, because the compound-symmetric formula can
understate the variance by up to about 6 percent in that
regime. When the correlation structure cannot be pinned
down from prior data, evaluating the AR(1) formula over
a plausible range of $\rho$ and taking the largest
resulting variance yields a safe sample size. The
moment-based computation of Section 4 addresses the
simple pre-post estimator and is not a bound on the GLS
variance.
:::

# Conflict of interest

The author declares no conflicts of interest.

# Funding

This work received no specific funding.

# Data availability statement

\begin{bullets}
\textbf{The implementing package, tests, and bibliography are openly available in the project repository.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The package implements the AR(1) variance formulas.
\item Validation tests and the shared four-paper bibliography are included.
\item Specific paths appear in the Reproducibility section below.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
The `runinpower` R package implementing the AR(1) variance
formulas (`covariance-ar1.R`), validation tests, and the
shared bibliography across the four-paper compendium are
openly available in the project repository. Specific paths
are listed in the Reproducibility section below.
:::

# Reproducibility

\begin{bullets}
\textbf{The manuscript was rendered with R 4.4.x using a small set of principal packages.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The in-package source supplies the AR(1) machinery.
\item Testing uses the tinytest harness.
\item The manuscript build uses rmarkdown and bookdown.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
This manuscript was rendered with R 4.4.x. Principal packages
used are the in-package `runinpower` source (the AR(1)
machinery is in `covariance-ar1.R`), `tinytest` (testing
harness), and `rmarkdown` plus `bookdown` for the manuscript
build.
:::

\begin{bullets}
\textbf{All reported quantities are deterministic.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Given fixed inputs the variance formulas exercise no random-number generator.
\item No Monte Carlo computation appears in this manuscript.
\item Re-running the chunks therefore reproduces every table and figure exactly.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
**Random-number generator.** The AR(1) variance formulas
are deterministic given fixed input parameters; no
random-number generation is used anywhere in this
manuscript, and re-running the code chunks reproduces
every table and figure exactly.
:::

\begin{bullets}
\textbf{All source, companion manuscripts, and bibliography reside at documented repository paths.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The R package source is in the package source directory.
\item The companion manuscripts for Papers 1, 3, and 4 have their own analysis subdirectories.
\item This manuscript source and the shared bibliography reside in the correlation-paper directory.
\end{itemize}
\end{bullets}

::: {.orig data-latex=""}
**Data and code availability.** The R package source is at
`R/`. The companion manuscripts (Papers 1, 3, 4) are at
`analysis/report/{01,03,04}-*/`. The manuscript source and
shared bibliography (`references.bib`, a symlink to
`../../references.bib`) are at
`analysis/report/02-correlation/`.
:::

# References


::: {#refs}
:::

```{r footer, echo=FALSE, results='asis'}
src <- sub("^/Users/zenn/Dropbox/prj/", "~/prj/",
           knitr::current_input(dir = TRUE))
cat(sprintf(
  "\n\\vfill\n\n---\n\n*Rendered on %s.*  \n*Source: `%s`*\n",
  format(Sys.time(), "%Y-%m-%d at %H:%M %Z"),
  src
))
```
