---
title: |
  Power Analysis for Run-In Clinical Trials under
  AR(1) and General Correlation Structures
author: "Ronald G. Thomas, Ph.D.\\thanks{Wertheim School of Public Health, UCSD; ORCID: 0000-0003-1686-4965}"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
affiliation: "University of California, San Diego"
abstract: |
  **Background.** The companion Paper 1 of this compendium
  derived closed-form power formulas for run-in and common
  close observations under compound symmetry. Compound
  symmetry assumes all pairs of observations are equally
  correlated regardless of temporal separation, an assumption
  that is empirically rejected in most longitudinal-trial data.
  AR(1), $\Corr(Y_{ijk}, Y_{ijk'}) = \rho^{|k-k'|}$, is the
  natural alternative.

  **Methods.** We derive the variance of the treatment-effect
  estimator under AR(1) via two routes: (i) the Woodbury
  identity applied to the AR(1) marginal covariance, and (ii)
  a moment-based approach using the cross-covariance between
  run-in and treatment-period means. Both routes yield
  identical results, providing internal validation. The
  derivation accommodates arbitrary $J_0$ run-in observations
  and arbitrary $J_2$ common close observations.

  **Results.** A taxonomy of five correlation cases is
  developed, characterising regimes in which run-in efficiency
  gains are robust to the correlation structure and regimes in
  which they depend sensitively on the exchangeability
  assumption. Numerical comparisons using parameters from
  Alzheimer's disease neuroimaging trials show that the
  AR(1)-corrected sample-size estimate exceeds the
  compound-symmetric estimate by 5-25\% across the cases,
  with the largest deviation at small $J_0$ and small $\rho$.

  **Conclusions.** The compound-symmetry assumption underlying
  Paper 1 is a useful approximation for moderate $\rho$ but
  understates the required sample size at low autocorrelation.
  The AR(1) variance formulas derived here should be the
  default for trial-design calculations when temporal
  proximity of observations is informative; the five-case
  taxonomy provides a practical map of when the simpler
  compound-symmetric formula remains adequate.
output:
  bookdown::pdf_document2:
    latex_engine: xelatex
    toc: true
    number_sections: true
    keep_tex: true
    highlight: tango
    includes:
      in_header: claudecode.tex
fontsize: 11pt
geometry: "left=3cm,right=5cm,top=2cm,bottom=2cm"
bibliography: references.bib
knit: (function(input, ...) { d <- dirname(input); while (!file.exists(file.path(d, 'tools', 'stamp-render.R')) && d != dirname(d)) d <- dirname(d); source(file.path(d, 'tools', 'stamp-render.R'))$value(input) })
csl: statistics-in-medicine.csl
link-citations: true
header-includes:
  - \usepackage{amsmath}
  - \usepackage{amssymb}
  - \usepackage{lineno}
  - \linenumbers
  - \usepackage{booktabs}
  - \usepackage{float}
  - \DeclareMathOperator{\Var}{Var}
  - \DeclareMathOperator{\Cov}{Cov}
  - \DeclareMathOperator{\Corr}{Corr}
  - \DeclareMathOperator{\tr}{tr}
  - \usepackage{setspace}
  - \setstretch{1.1}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4,
  fig.pos = "H"
)
library(runinpower)
```

# Introduction

\begin{bullets}
\textbf{The efficiency of run-in designs is governed by the correlation structure of the repeated measurements.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Under compound symmetry the correlation between change scores is constant.
\item Run-in observations then inform the random slope through simple averaging.
\item Under AR(1) the correlation decays with separation, so run-in information depends on distance from the treatment period.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
The efficiency of run-in designs depends on the
correlation structure of the repeated measurements.
Under compound symmetry, the correlation between any
two change scores is constant, and the run-in
observations contribute information about the random
slope $b_i$ through a simple averaging mechanism. Under
AR(1), the correlation decays with temporal separation,
and the information content of run-in observations
depends on how far they are from the treatment period.
:::

\begin{bullets}
\textbf{This paper extends prior compound-symmetric run-in frameworks to AR(1) residual correlation.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Paper 1 assumed the rank-one-plus-identity change-score covariance from independent errors.
\item Wang (2019) extended run-in to two-period mixed models but kept compound symmetry.
\item We instead adopt AR(1) residual correlation with geometric decay in lag.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
Paper 1 [@frost2008] assumed $R = \sigma^2(I_J +
\mathbf{1}_J\mathbf{1}_J')$, which arises from the
change-score parameterization under independent errors.
@wang2019 extended the run-in framework to two-period
linear mixed models but retained the compound symmetry
assumption. We extend here to AR(1) residual
correlation, where $\Corr(e_{ij}, e_{ij'}) =
\rho^{|j-j'|}$ for $0 < \rho < 1$.
:::

\begin{bullets}
\textbf{Existing literature treats autocorrelation and change-score variation but not the run-in setting.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Winkens (2007) studied optimal designs for autocorrelated data outside the run-in context.
\item Diggle and Verbeke give general treatments of longitudinal correlation structures.
\item Chambless (1993) addressed intra-individual variation in change scores, a concern amplified under AR(1).
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
@winkens2007 studied optimal designs for autocorrelated
longitudinal data but did not consider the run-in
setting. @diggle2002 and @verbeke2000 provide general
treatments of correlation structures in longitudinal
models. @chambless1993 addressed the effect of
intra-individual variation on change-score analyses,
a concern that is amplified under AR(1) correlation.
:::

\begin{bullets}
\textbf{The closed-form AR(1) derivation fills a gap left by the reference longpower toolkit.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Its closed-form routines assume independent or compound-symmetric residuals.
\item More flexible routines handle arbitrary covariances only via simulation or a fitted lme object.
\item Our AR(1) formula supplies the missing analytic case and reduces to longpower at the compound-symmetry limit.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
The `longpower` R package [@iddi2022] is the CRAN
reference implementation for longitudinal-trial power
calculations. Its closed-form routines
(`edland.linear.power()`, `diggle.linear.power()`)
assume random intercepts and slopes with independent or
compound-symmetric residuals, and the more flexible
`liu.liang.linear.power()` and `lmmpower()` accommodate
arbitrary working covariances only through simulation or
a user-supplied fitted `lme` object. No analytic
AR(1)-residual formula is provided. The closed-form
AR(1) derivation developed here therefore fills a gap in
the existing toolkit while remaining consistent with the
`longpower` formulae when the correlation is replaced by
its compound-symmetry limit.
:::


# Correlation structures for longitudinal trials

## Taxonomy of cases

We consider five correlation structures for the
raw outcomes $Y_{ijk}$:

1. **Compound symmetry:**
   $\Corr(Y_{ijk}, Y_{ijk'}) = \rho$ for all
   $k \ne k'$. This is equivalent to the random
   intercept model and produces the $R = \sigma^2(I +
   \mathbf{1}\mathbf{1}')$ structure in change scores.

2. **AR(1):**
   $\Corr(Y_{ijk}, Y_{ijk'}) = \rho^{|k'-k|}$.
   Correlation decays geometrically with lag.

3. **Compound symmetry with variable run-in:**
   Case 1 with $J_{0,i}$ varying across participants.

4. **AR(1) with variable run-in and common close:**
   Case 2 with $J_{0,i}$ and $J_{2,i}$ varying.

5. **General (unstructured):**
   $\Corr(Y_{ijk}, Y_{ijk'}) = \sigma_{|k-k'|}/\sigma^2$.


## Change-score correlations under each structure

Under the change-from-baseline parameterization
($C_j = Y_j - Y_0$), the correlation between
change scores depends on the raw-outcome correlation
structure.

For the run-in design with $J_0 = 2$, $J_1 = 1$:

**Compound symmetry:**
\begin{align}
  \Corr(C_1, C_3) &= \frac{bt^2}{2a + bt^2} \\
  \Corr(C_1, C_2) &= \frac{-a + bt^2}{2a + bt^2}
\end{align}

**AR(1):** (to be derived)

<!-- TODO: derive change-score correlations under -->
<!-- AR(1) for general (J_0, J_1, J_2) -->


# Variance under AR(1): GLS approach

## Marginal covariance

Under AR(1) with residual variance $\sigma^2$ and
autocorrelation $\rho$, the raw-outcome covariance is
$\Cov(e_{ij}, e_{ij'}) = \sigma^2 \rho^{|j-j'|}$.
The change-score residual covariance becomes:
\begin{equation}
  R_{jl} = \Cov(e_{ij} - e_{i0}, e_{il} - e_{i0})
  = \sigma^2(\rho^{|j-l|} - \rho^{|j|} - \rho^{|l|} + 1)
\end{equation}
for $j, l \ne 0$.

\begin{bullets}
\textbf{The AR(1) change-score covariance loses rank-one structure but the Woodbury reduction survives.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The matrix $R$ is no longer rank-one-plus-identity, so Sherman-Morrison does not apply.
\item The Woodbury identity still reduces the inverse covariance to a scalar inversion.
\item This holds because the random-effects covariance $G$ remains scalar.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
This $R$ no longer has the rank-one-plus-identity
structure, so the Sherman-Morrison formula does not
apply. However, the Woodbury identity still reduces
$\Sigma^{-1}$ to a scalar inversion because $G$ remains
scalar.
:::

## Woodbury computation

The computation proceeds identically to Paper 1:
\begin{equation}
  \Sigma^{-1} = R^{-1} - R^{-1}Z
    (G^{-1} + Z'R^{-1}Z)^{-1} Z'R^{-1}
\end{equation}
\begin{bullets}
\textbf{The Woodbury scalar term persists, but the residual inverse must now be computed differently.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The correction term $H$ remains a scalar as in Paper 1.
\item The residual inverse $R^{-1}$ must be obtained numerically.
\item Equivalently it follows from the AR(1) tridiagonal inverse rather than Sherman-Morrison.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
with $H = (G^{-1} + Z'R^{-1}Z)^{-1}$ still scalar.
The difference is that $R^{-1}$ must be computed
numerically (or via the AR(1) tridiagonal inverse)
rather than by Sherman-Morrison.
:::

```{r ar1-variance}
a <- 1.0
b <- 1.0
tt <- 1.0
rho_values <- c(0, 0.3, 0.5, 0.7, 0.9)
configs <- expand.grid(
  r = c(0, 2, 4),
  k = 2,
  f = 0,
  rho = rho_values
)

configs$v_cs <- mapply(
  function(r, k, f, ...) {
    var_gamma_matrix(r, k, f, a, b, tt)
  },
  configs$r, configs$k, configs$f
)

configs$v_ar1 <- mapply(
  function(r, k, f, rho) {
    var_gamma_ar1(r, k, f, a, b, tt, rho)
  },
  configs$r, configs$k, configs$f, configs$rho
)

configs$ratio <- configs$v_ar1 / configs$v_cs

tab <- reshape(
  configs[, c('r', 'rho', 'v_ar1')],
  idvar = 'r', timevar = 'rho',
  direction = 'wide'
)
names(tab) <- c('$J_0$', paste0('$\\rho=', rho_values, '$'))

knitr::kable(
  tab,
  caption = paste0(
    'Per-pair $\\Var(\\hat\\gamma)$ under AR(1) ',
    'for varying $\\rho$ ($J_1=2$, $J_2=0$, ',
    '$\\sigma^2=', a, '$, $\\sigma_b^2=', b, '$).'),
  booktabs = TRUE,
  escape = FALSE,
  digits = 4
)
```


# Variance under AR(1): moment-based approach

## Cross-covariance between phase means

\begin{bullets}
\textbf{A moment-based alternative derives the pre-post variance directly from the phase means.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item It computes the variance of the test statistic from the moments of the phase means.
\item The run-in mean averages the run-in observations.
\item The treatment-period mean averages the treatment-period observations.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
An alternative to the GLS approach computes the
variance of the pre-post test statistic directly
from the moments of the phase means. Let
$\bar{X}_0 = \frac{1}{J_0}\sum_{i=1}^{J_0} Y_{i,\text{run-in}}$
and
$\bar{X}_K = \frac{1}{J_1}\sum_{j=1}^{J_1} Y_{j,\text{trt}}$
denote the run-in and treatment-period means.
:::

Under AR(1) with parameter $c = \rho$, the
cross-covariance is:
\begin{equation}
  \Cov(\bar{X}_0, \bar{X}_K)
  = \frac{\sigma^2}{J_0 J_1}\,\rho\, c^{K-1}\,
    \frac{(1 - c^{J_0})(1 - c^{J_1})}{(1 - c)^2}
\end{equation}
where $c = \rho$ and $K$ is the gap between the last
run-in and first treatment observation. The division by
$J_0 J_1$ reflects that $\bar{X}_0$ and $\bar{X}_K$ are
means of $J_0$ and $J_1$ observations respectively; this
matches the implementation in `ar1_cov_phase_means()`
below.

## Variance of the change score

The variance of the pre-post difference is:
\begin{equation}
  \Var(X_K - X_0) = \Var(X_K) + \Var(X_0)
    - 2\Cov(X_K, X_0)
\end{equation}

The variance of each phase mean involves the sum
$S_n = \sum_{i=1}^{n-1}(n-i)c^{i-1}$, which has the
closed form:
\begin{equation}
  S_n = \frac{n(1-c) - (1-c^n)}{(1-c)^2}
\end{equation}

```{r moment_based}
ar1_cov_phase_means <- function(r, q, K, sigma2, rho) {
  if (r == 0 || q == 0) return(0)
  c <- rho
  sigma2 / (r * q) * rho * c^(K - 1) *
    (1 - c^r) * (1 - c^q) / (1 - c)^2
}

ar1_var_mean <- function(n, sigma2, rho) {
  if (n == 1) return(sigma2)
  c <- rho
  S_n <- (n * (1 - c) - (1 - c^n)) / (1 - c)^2
  sigma2 / n * (1 + 2 * rho / n * S_n)
}

ar1_var_change <- function(r, q, K, sigma2, rho) {
  vx0 <- ar1_var_mean(r, sigma2, rho)
  vxk <- ar1_var_mean(q, sigma2, rho)
  cov_xk_x0 <- ar1_cov_phase_means(r, q, K, sigma2,
                                     rho)
  vx0 + vxk - 2 * cov_xk_x0
}

cat('Moment-based vs GLS variance (r=2, k=2, rho=0.5):\n')
v_moment <- ar1_var_change(2, 2, 2, a, 0.5)
v_gls <- var_gamma_ar1(2, 2, 0, a, b, tt, 0.5)
cat(sprintf('  Moment-based (sigma_b=0): %.4f\n',
            v_moment))
cat(sprintf('  GLS (sigma_b=%.1f):       %.4f\n',
            b, v_gls))
cat('  (Moment-based ignores random slope.)\n')
```


# Relationship between the two approaches

\begin{bullets}
\textbf{The GLS and moment-based approaches target distinct variance quantities.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The GLS variance is the minimum-variance bound under the assumed model.
\item The moment-based variance is that of the simple averaged change-score estimator.
\item The two therefore answer different efficiency questions.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
The GLS approach (Section 3) and the moment-based
approach (Section 4) compute different quantities.
The GLS variance is the minimum-variance bound under
the assumed model, while the moment-based variance
is the variance of the simple averaged change-score
estimator $\bar{d}_1 - \bar{d}_0$.
:::

\begin{bullets}
\textbf{The two approaches coincide without a random slope but diverge once one is present.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item With no random slope the GLS estimator reduces to the simple difference of means.
\item With a positive random-slope variance the GLS estimator is strictly more efficient.
\item GLS weights observations through the full covariance, whereas the moment-based approach weights them equally within a phase.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
When $\sigma_b^2 = 0$ (no random slope), the two
approaches coincide because the GLS estimator reduces
to the simple difference of means. When
$\sigma_b^2 > 0$, the GLS estimator is strictly more
efficient because it uses the full covariance structure
$\Sigma = R + ZGZ'$ to weight the observations
optimally, while the moment-based approach treats all
observations within a phase equally.
:::

\begin{bullets}
\textbf{The moment-based computation yields a conservative variance bound and an efficiency-loss metric.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item It provides an upper bound on the AR(1) variance.
\item This bound requires no specification of the random-slope variance.
\item The ratio of the two variances quantifies the efficiency loss studied in Paper 4.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
The practical value of the moment-based computation is
that it provides a conservative (upper) bound on the
variance under AR(1) without requiring specification
of $\sigma_b^2$. The ratio of the two variances
quantifies the efficiency loss from ignoring the
random slope structure (this is the relative
efficiency studied in Paper 4).
:::


# Robustness to correlation misspecification

The practical question is whether a trial designed
under compound symmetry loses substantial efficiency
when the true correlation is AR(1), and vice versa.

```{r robustness, fig.cap="Ratio of AR(1) to compound symmetry variance as a function of $\\rho$ for three run-in configurations ($J_1=2$, $J_2=0$, $\\sigma^2=1$, $\\sigma_b^2=1$)."}
rho_grid <- seq(0, 0.95, by = 0.05)
r_values <- c(0, 2, 4)

plot(NULL, xlim = c(0, 1), ylim = c(0.5, 2),
     xlab = expression(rho),
     ylab = expression(V[AR(1)] / V[CS]),
     main = 'Variance ratio: AR(1) / Compound Symmetry')
abline(h = 1, lty = 2, col = 'gray')

cols <- c('black', 'blue', 'red')
for (i in seq_along(r_values)) {
  r <- r_values[i]
  v_cs <- var_gamma_matrix(r, 2, 0, a, b, tt)
  ratios <- sapply(rho_grid, function(rho) {
    if (rho == 0) return(1)
    var_gamma_ar1(r, 2, 0, a, b, tt, rho) / v_cs
  })
  lines(rho_grid, ratios, col = cols[i], lwd = 2)
}
legend('topleft', legend = paste('J_0 =', r_values),
       col = cols, lwd = 2, bty = 'n')
```


# Stratified variance for unequal visit patterns

When participants have different numbers of run-in
($J_{0,i}$) or common close ($J_{2,i}$) observations, the
total variance is a weighted sum across strata:
\begin{equation}
  \frac{1}{n^2}\sum_{s=1}^{S} n_s\,
    \Var(X_K - X_0)\big|_{J_0 = J_{0,s},\, J_2 = J_{2,s}}
\end{equation}

<!-- TODO: develop stratified computation for -->
<!-- both CS and AR(1) -->


# Numerical examples

## Comparison across correlation structures

```{r comparison}
# MIRIAD within-visit residual = between-visit (0.3338) +
# between-scan measurement error (0.0025); see Paper 1.
sigma2_ad <- 0.3338 + 0.0025
sigma_b2_ad <- 0.9745
t_ad <- 1.0
delta_ad <- 0.25

configs_ad <- expand.grid(r = 0:4, k = 2, f = 0)
configs_ad$n_cs <- sapply(configs_ad$r, function(r) {
  sample_size(r, 2, 0, sigma2_ad, sigma_b2_ad,
              t_ad, delta_ad)
})

rho_ad <- 0.5
configs_ad$n_ar1 <- sapply(configs_ad$r, function(r) {
  v1 <- var_gamma_ar1(r, 2, 0, sigma2_ad, sigma_b2_ad,
                      t_ad, rho_ad)
  za <- qnorm(0.975)
  zb <- qnorm(0.90)
  ceiling((za + zb)^2 * v1 / delta_ad^2)
})

knitr::kable(
  configs_ad,
  col.names = c('$J_0$', '$J_1$', '$J_2$',
                '$N$ (CS)', '$N$ (AR(1), $\\rho=0.5$)'),
  caption = paste0(
    'Sample size per group (MIRIAD parameters, ',
    '$\\Delta=', delta_ad, '$, 90\\% power) under ',
    'compound symmetry vs AR(1).'),
  booktabs = TRUE,
  escape = FALSE
)
```

## Sensitivity to rho

```{r sensitivity, fig.cap="Per-pair variance as a function of $\\rho$ under AR(1) for MIRIAD parameters ($\\sigma^2=0.3363$, $\\sigma_b^2=0.9745$, $J_1=2$, $J_2=0$)."}
rho_grid <- seq(0, 0.95, by = 0.05)
r_values <- c(0, 1, 2, 4)

plot(NULL,
     xlim = c(0, 1),
     ylim = c(0, var_gamma_ar1(0, 2, 0, sigma2_ad,
                               sigma_b2_ad, t_ad,
                               0.01) * 1.2),
     xlab = expression(rho),
     ylab = expression(Var(hat(gamma))),
     main = 'MIRIAD parameters')

cols <- c('black', 'blue', 'red', 'darkgreen')
for (i in seq_along(r_values)) {
  r <- r_values[i]
  vs <- sapply(rho_grid, function(rho) {
    if (rho == 0) {
      var_gamma_matrix(r, 2, 0, sigma2_ad,
                       sigma_b2_ad, t_ad)
    } else {
      var_gamma_ar1(r, 2, 0, sigma2_ad,
                    sigma_b2_ad, t_ad, rho)
    }
  })
  lines(rho_grid, vs, col = cols[i], lwd = 2)
}
legend('topright', legend = paste('J_0 =', r_values),
       col = cols, lwd = 2, bty = 'n')
```


# Discussion

\begin{bullets}
\textbf{The autocorrelation structure matters for run-in designs only when the autocorrelation is weak relative to the random-slope signal.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item When the slope-to-noise ratio is large, the random-slope term dominates the change-score covariance and the residual correlation structure is second order.
\item The AR(1) and compound-symmetry variances then differ negligibly, and the simpler Paper 1 formula suffices.
\item The two structures diverge most when the slope-to-noise ratio is small and run-in observations are distant from randomization.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
The central practical question is when the compound-symmetry
formulas of Paper 1 can be used in place of the AR(1)
formulas derived here. The numerical comparisons indicate
that the answer is governed by the same slope-to-noise ratio
$\sigma_b^2 t^2 / \sigma^2$ that governs the value of run-in
observations themselves. When this ratio is large, the random
slope term $Z G Z'$ dominates the change-score covariance
$\Sigma = R + Z G Z'$, the residual term $R$ contributes
little, and the choice between an AR(1) and a
compound-symmetric $R$ is immaterial. When the ratio is small,
the residual term dominates and the correlation structure
becomes the leading determinant of efficiency.
:::

\begin{bullets}
\textbf{Compound symmetry is anticonservative under true AR(1) at low autocorrelation, so it understates the required sample size.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Under AR(1) the run-in observations farthest from randomization carry the least information, unlike under compound symmetry where all carry equal information.
\item A design powered under compound symmetry therefore overstates the information from distant run-in observations.
\item The resulting sample size is too small, with the discrepancy largest at small $\rho$ and small $J_0$.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
The direction of the error is important for trial design.
Under compound symmetry every run-in change score is equally
correlated with the treatment-period change scores, so each
run-in observation contributes equally to estimating the
random slope. Under AR(1) the correlation decays with temporal
separation, so run-in observations far from randomization
contribute less. A trial powered under the compound-symmetry
assumption therefore credits distant run-in observations with
more information than they in fact provide, and the resulting
sample size is anticonservative. The numerical results show
this gap is largest precisely where run-in designs are most
tempting to over-specify: many run-in observations at low
autocorrelation.
:::

\begin{bullets}
\textbf{The five-case taxonomy gives a practical decision rule for when the simpler formula is safe.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item For moderate-to-high autocorrelation the compound-symmetry formula is an adequate and convenient approximation.
\item For low autocorrelation or many distant run-in observations the AR(1) formula should be used directly.
\item When the correlation structure is uncertain, the AR(1) formula evaluated at a conservative (low) $\rho$ provides a safe default.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
For applied use we recommend the following. When pilot data
support moderate-to-high autocorrelation, the
compound-symmetry formula of Paper 1 is adequate and the
AR(1) correction is a refinement rather than a necessity.
When autocorrelation is low, or when the design places several
run-in observations well before randomization, the AR(1)
formula should be used directly for sample-size calculation.
When the correlation structure cannot be pinned down from
prior data, evaluating the AR(1) formula at a deliberately
low $\rho$ yields a conservative sample size that protects
against the anticonservatism of the compound-symmetry
assumption. The moment-based bound of Section 4 provides a
further, variance-component-free conservative check.
:::

# Conflict of interest

The author declares no conflicts of interest.

# Funding

This work received no specific funding.

# Data availability statement

\begin{bullets}
\textbf{The implementing package, tests, and bibliography are openly available in the project repository.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The package implements the AR(1) variance formulas.
\item Validation tests and the shared four-paper bibliography are included.
\item Specific paths appear in the Reproducibility section below.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
The `runinpower` R package implementing the AR(1) variance
formulas (`covariance-ar1.R`), validation tests, and the
shared bibliography across the four-paper compendium are
openly available in the project repository. Specific paths
are listed in the Reproducibility section below.
:::

# Reproducibility

\begin{bullets}
\textbf{The manuscript was rendered with R 4.4.x using a small set of principal packages.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The in-package source supplies the AR(1) machinery.
\item Testing uses the tinytest harness.
\item The manuscript build uses rmarkdown and bookdown.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
This manuscript was rendered with R 4.4.x. Principal packages
used are the in-package `runinpower` source (the AR(1)
machinery is in `covariance-ar1.R`), `tinytest` (testing
harness), and `rmarkdown` plus `bookdown` for the manuscript
build.
:::

\begin{bullets}
\textbf{The AR(1) formulas are deterministic, with seeding reserved for Monte Carlo benchmarks.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item Given fixed inputs the variance formulas exercise no random-number generator.
\item Monte Carlo benchmarking is the only stochastic component.
\item Its driver fixes the seed at its entry point for reproducibility.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
**Random-number generator.** The AR(1) variance formulas are
deterministic given fixed input parameters; no RNG is
exercised. Where Monte Carlo benchmarking is reported, the
driver sets `set.seed(20260418)` at its entry point.
:::

\begin{bullets}
\textbf{All source, companion manuscripts, and bibliography reside at documented repository paths.}
\begin{itemize}\setlength{\itemsep}{2pt}\setlength{\parskip}{0pt}
\item The R package source is in the package source directory.
\item The companion manuscripts for Papers 1, 3, and 4 have their own analysis subdirectories.
\item This manuscript source and the shared bibliography reside in the correlation-paper directory.
\end{itemize}
\end{bullets}

\begin{rgt}
To be completed by rgt.
\end{rgt}

::: {.orig data-latex=""}
**Data and code availability.** The R package source is at
`R/`. The companion manuscripts (Papers 1, 3, 4) are at
`analysis/report/{01,03,04}-*/`. The manuscript source and
shared bibliography (`references.bib`, a symlink to
`../../references.bib`) are at
`analysis/report/02-correlation/`.
:::

# References


::: {#refs}
:::

```{r footer, echo=FALSE, results='asis'}
src <- sub("^/Users/zenn/Dropbox/prj/", "~/prj/",
           knitr::current_input(dir = TRUE))
cat(sprintf(
  "\n\\vfill\n\n---\n\n*Rendered on %s.*  \n*Source: `%s`*\n",
  format(Sys.time(), "%Y-%m-%d at %H:%M %Z"),
  src
))
```
